### 2020-01-22 gradu/gradu.tex
adds gradu template

%% History:
%% May 2019 Tomi Männistö, Antti-Pekka Tuovinen proofreading; 30 vs. 40 cr theses, etc.
%% May 2019 Tomi Männistö changed from babelbib to bibtex; Abstract page (and other pages as well) reformatting.
%% January–May 2019 several issues fixed by Niko Mäkitalo; long fields in abstract
%% March 2018 template file extended by Lea Kutvonen to exploit HYthesisML.cls.
%% Feb2018 This template file for the use of HYgraduML.cls was  modified by Veli Mäkinen from HY_fysiikka_LuKtemplate.tex
%% authored by Roope Halonen ja Tomi Vainio in 2017.
%% Some text is also inherited from engl_malli.tex versions by Kutvonen, Erkiö, Mäkelä, Verkamo, Kurhila, and
%% Nykänen, to accompany tktltiki.cls (by Puolakka 2002).

%% Follow comments to support use.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 1: Choose options for MSc / BSc layout and your bibliographic style
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%  Language: 
%%      finnish, swedish, or english
%%  Pagination (use twoside by default)  
%%      oneside or twoside,
%%  Study programme / kind of report
%%      csm  = pro gradu in new Computer science MSc;
%%      cs = pro gradu in old Computer Science MSc;
%%      tkt = BSc thesis in new curricula;
%%      tktl= BSc thesis in old curricula;
%%  For MSc choose your line or track:
%%      (30 cr thesis, 2020 onwards, Master of Computer Science programme = csm)
%%      software-track-2020 = Software study track
%%      algorithms-track-2020 = Algorithms study track
%%      networking-track-2020 = Networking study track
%%
%%      (30 cr thesis, Master of Computer Science programme = csm)
%%      sw-track-2018 = Software Systems study track
%%      alko-track-2018 = Algorithms study track
%%      nodes-track-2018 = Networking and Services study track
%%
%%      (30 cr thesis, Master of Computer Science programme = csm)
%%      sw-line-2017 =  Software systems subprogramme
%%      alko-line-2017 = Algorithms, Data Analytics and Machine Learning subprogramme
%%      bio-line-2017 = Algorithmic Bioinformatics subprogramme
%%      nodes-line-2017 = Networking and Services subprogramme
%%
%%      (40 cr thesis, = cs)
%%      sw-line = Software Systems specialisation line 
%%      alko-line = Algorithms specialisation line
%%      bio-line = Algorithmic bioinformatics specialisation line
%%      nodes-line = Networking and Services specialisation line

\documentclass[english,twoside,censored,tkt,alko-line]{HYthesisML}

% In theses, open new chapters only at right page.
% For other types of documents, may ask "openany" in document.
\PassOptionsToClass{openright,twoside,a4paper}{report}
%\PassOptionsToClass{openany,twoside,a4paper}{report}

\usepackage{csquotes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% REFERENCES
%% Some notes on bibliography usage and options:
%% natbib -> you can use, e.g., \citep{} or \parencite{} for (Einstein, 1905); with APA \cite -> Einstein, 1905 without ()
%% maxcitenames=2 -> only 2 author names in text citations, if more -> et al. is used
%% maxbibnames=99 as no great need to suppress the biliography list in a thesis
%% for more information see biblatex package documentation, e.g., from https://ctan.org/pkg/biblatex

%% Reference style: select one 
%% for APA = Harvard style = authoryear -> (Einstein, 1905) use:
\usepackage[style=authoryear,bibstyle=authoryear,backend=biber,natbib=true,maxnames=99,maxcitenames=2,giveninits=true,uniquename=init]{biblatex}
%% for numeric = Vancouver style -> [1] use:
%\usepackage[style=numeric,bibstyle=numeric,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%% for alpahbetic -> [Ein05] use:
%\usepackage[style=alphabetic,bibstyle=alphabetic,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%

\addbibresource{bibliography.bib}
% in case you want the final delimiter between authors & -> (Einstein & Zweistein, 1905) 
% \renewcommand{\finalnamedelim}{ \& }
% List the authors in the Bibilipgraphy as Lastname F, Familyname G,
\DeclareNameAlias{sortname}{family-given}
% remove the punctuation between author names in Bibliography 
%\renewcommand{\revsdnamepunct}{ }

%% Block of definitions for fonts and packages for picture management.
%% In some systems, the figure packages may not be happy together.
%% Choose the ones you need.

%\usepackage[utf8]{inputenc} % For UTF8 support, in some systems. Use UTF8 when saving your file.

\usepackage{lmodern}         % Font package, again in some systems.
\usepackage{textcomp}        % Package for special symbols
\usepackage[pdftex]{color, graphicx} % For pdf output and jpg/png graphics
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage[pdftex, plainpages=false]{hyperref} % For hyperlinks and pdf metadata
\usepackage{fancyhdr}        % For nicer page headers
\usepackage{tikz}            % For making vector graphics (hard to learn but powerful)
%\usepackage{wrapfig}        % For nice text-wrapping figures (use at own discretion)
\usepackage{amsmath, amssymb} % For better math

\singlespacing               %line spacing options; normally use single

\fussy
%\sloppy                      % sloppy and fussy commands can be used to avoid overlong text lines
% if you want to see which lines are too long or have too little stuff, comment out the following lines
% \overfullrule=1mm
% to see more info in the detailed log about under/overfull boxes...
% \showboxbreadth=50 
% \showboxdepth=50

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 2:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Set up personal information for the title page and the abstract form.
%% Replace parameters with your information.
\title{Tutkielman otsikko}

% TM: Contributors to template editors now listed in the beginning of the file in comments
\author{Jussi Timonen}
\date{\today}

% Set supervisors and examiners, use the titles according to the thesis language
% Prof. 
% Dr. or in Finnish toht. or tri or FT, TkT, Ph.D. or in Swedish... 
\supervisors{Prof.~D.U.~Mind, Dr.~O.~Why}
\examiners{Prof.~D.U.~Mind, Dr.~O.~Why}

\keywords{ulkoasu, tiivistelmä, lähdeluettelo}
\additionalinformation{\translate{\track}}

%% For seminar papers and such, cover page and abstract
%% requires these three basic items of information.
%% Label as needed and remove comment marks.
%%\level{Seminaariraportti}
%%\programme{Tietojenkäsittelytieteen andidaattiohjelma}
%%\subject{Seminaarisarjan nimi}

%% Replace classification terms with the ones that match your work. ACM
%% ACM Digital library provides a taxonomy and a tool for classification
%% in computer science. Use 1-3 paths, and use right arrows between the
%% about three levels in the path; each path requires a new line.

\classification{\protect{\ \\
\  General and reference $\rightarrow$ Document types  $\rightarrow$ Surveys and overviews\  \\
\  Applied computing  $\rightarrow$ Document management and text processing  $\rightarrow$ Document management $\rightarrow$ Text editing
}}

%% if you want to quote someone special. You can comment this line out and there will be nothing on the document.
%\quoting{Bachelor's degrees make pretty good placemats if you get them laminated.}{Jeph Jacques}

%% OPTIONAL STEP: Set up properties and metadata for the pdf file that pdfLaTeX makes.
%% Your name, work title, and keywords are recommended.
\hypersetup{
    unicode=true,           % to show non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={},            % title
    pdfauthor={},           % author
    pdfsubject={},          % subject of the document
    pdfcreator={},          % creator of the document
    pdfproducer={pdfLaTeX}, % producer of the document
    pdfkeywords={something} {something else}, % list of keywords for
    pdfnewwindow=true,      % links in new window
    colorlinks=true,        % false: boxed links; true: colored links
    linkcolor=black,        % color of internal links
    citecolor=black,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}

%%-----------------------------------------------------------------------------------

\begin{document}

% Generate title page.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 3:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Write your abstract to be positioned here.
%% You can make several abstract pages (if you want it in different languages),
%% but you should also then redefine some of the above parameters in the proper
%% language as well, in between the abstract definitions.

\begin{abstract}

Tämä dokumentti on tarkoitettu Helsingin yliopiston tietojenkäsittelytieteen osaston opin\-näyt\-teiden ja harjoitustöiden ulkoasun ohjeeksi ja mallipohjaksi. Ohje soveltuu kanditutkielmiin, ohjelmistotuotantoprojekteihin, seminaareihin ja maisterintutkielmiin. Tämän ohjeen lisäksi on seurattava niitä ohjeita, jotka opastavat valitsemaan kuhunkin osioon tieteellisesti kiinnostavaa, syvällisesti pohdittua sisältöä.

Työn aihe luokitellaan  
ACM Computing Classification System (CCS) mukaisesti, 
ks.\ \url{https://www.acm.org/about-acm/class},
käyttäen komentoa \verb+\classification{}+. 
Käytä muutamaa termipolkua (1--3), jotka alkavat juuritermistä ja joissa polun tarkentuvat luokat erotetaan toisistaan oikealle osoittavalla nuolella.

\end{abstract}

\begin{otherlanguage}{english}
\begin{abstract}
Use this otherlanguage environment to write your abstract in another language if needed.

Topics are classified according to the ACM Computing Classification System
(CCS), see 
\url{https://www.acm.org/about-acm/class}:
check command \verb+\classification{}+. A small set of paths (1--3) should be used, starting from any top nodes
referred to bu the root term CCS leading to the leaf nodes. The elements
in the path are separated by right arrow, and emphasis of each element individually can be indicated
by the use of bold face for high importance or italics for intermediate
level. The combination of individual boldface terms may give the reader
additional insight. 
\end{abstract}
\end{otherlanguage}

% Place ToC
\newpage
\mytableofcontents
\frontmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 4: Write the thesis.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Your actual text starts here. You shouldn't mess with the code above the line except
%% to change the parameters. Removing the abstract and ToC commands will mess up stuff.
%%
%% You may wish to include material to avoid browsing the definitions
%% above. Command \include{file} includes the file of name file.tex.
%% As a side effect, subsequent inclusions may force a page break.

% BSc instructions
%\include{bsc_finnish_contents}
%\include{bsc_english_contents}
% MSc instructions
%\include{msc_finnish_contents}
\include{msc_english_contents}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cleardoublepage                          %fixes the position of bibliography in bookmarks
\phantomsection
\addcontentsline{toc}{chapter}{\bibname}  % This lines adds the bibliography to the ToC
\printbibliography

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
\begin{appendices}

\input{instructions_english}
%\input{instructions_finnish}

\appendix{Sample Appendix\label{appendix:model}}
usually starts on its own page, with the name and number of the appendix at the top. 
The appendices here are just models of the table of contents and the presentation. Each appendix
Each appendix is paginated separately.

In addition to complementing the main document, each appendix is also its own, independent entity.
This means that an appendix cannot be just an image or a piece of programming, but the appendix must explain its contents and meaning.

\end{appendices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-01-22 gradu/msc_english_contents.tex
adds gradu template

\chapter{Introduction}

The following gives some superficial instructions for using this template for a Master's thesis. For guidelines on thesis writing you can consult various sources, for example, the Bachelor thesis template.

The thesis should have an introduction chapter. Other chapters can be named according to the topic. In the end, some summary chapter is needed; see Chapter~\ref{chapter:conclusions} for an example.

\chapter{Figures and Tables}

\section{Figures}
Figure~\ref{fig:logo} gives an example how to add figures to the document. Remember always to cite the figure in the main text.

\begin{figure}[h!] 
\centering 
\includegraphics[width=0.3\textwidth]{HY-logo-ml.png}
\caption{University of Helsinki flame-logo for Faculty of Science.\label{fig:logo}}
\end{figure}

\section{Tables}

Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.

\begin{table}
\centering
\caption{Experimental results.\label{table:results}}
\begin{tabular}{l||l c c r} 
Experiment & 128 & 256 & 32768 & 65536 \\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03\\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94\\

\hline
%
\end{tabular}
\end{table}

\chapter{Citations}

\section{Citations to literature}

References are listed in a separate .bib-file. In this case it is named \texttt{bibliography.bib} including the following content:
\begin{verbatim}
@article{einstein,
    author =       "Albert Einstein",
    title =        "{Zur Elektrodynamik bewegter K{\"o}rper}. ({German})
        [{On} the electrodynamics of moving bodies]",
    journal =      "Annalen der Physik",
    volume =       "322",
    number =       "10",
    pages =        "891--921",
    year =         "1905",
    DOI =          "http://dx.doi.org/10.1002/andp.19053221004"
}

@book{latexcompanion,
    author    = "Michel Goossens and Frank Mittelbach and Alexander Samarin",
    title     = "The \LaTeX\ Companion",
    year      = "1993",
    publisher = "Addison-Wesley",
    address   = "Reading, Massachusetts"
}

@misc{knuthwebsite,
    author    = "Donald Knuth",
    title     = "Knuth: Computers and Typesetting",
    url       = "http://www-cs-faculty.stanford.edu/%7Eknuth/abcde.html"
}
\end{verbatim}

In the last reference url field the code \verb+%7E+ will translate into \verb+~+ once clicked in the final pdf.

References are created using command \texttt{\textbackslash cite\{einstein\}}, showing as \citep{einstein}. Other examples: \citep{latexcompanion,knuthwebsite}.

Citation style can be negotiated with the supervisor. See some options in \url{https://www.sharelatex.com/learn/Bibtex_bibliography_styles}.

\section{Crossreferences}

Appendix~\ref{appendix:model} on page~\pageref{appendix:model} contains some additional material.

\chapter{From tex to pdf}

In Linux, run \texttt{pdflatex filename.tex} and \texttt{bibtex filename.tex} repeatedly until no more warnings are shown. This process can be automised using make-command.

\chapter{Conclusions\label{chapter:conclusions}}

It is good to conclude with a summary of findings. You can also use separate chapter for discussion and future work. These details you can negotiate with your supervisor.

### 2020-01-22 vb_encode_7bit_byteoptimized.cpp

@@ -74,8 +74,7 @@ int main(int argc, char *argv[]) {
     iv[index] = *i%cap;
   }
   cout << "VByte data vector size: " << sizeof(std::vector<uint32_t>) + sizeof(uint32_t)*vdata.size() << endl;
-  rank_support_v<> rb(&b);
-  cout << "rank " << rb(vdata.size()) << endl;
+
   select_support_mcl<> sls(&b);
 
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
@@ -105,8 +104,8 @@ int main(int argc, char *argv[]) {
     }
     z = z^val;
   }
-  cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::chrono::steady_clock::time_point end = std::chrono::steady_clock::now();
+  cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::cout << "Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (end - begin).count() << "[ms]" << std::endl;
   return 0;
 }

### 2020-01-22 vb_encode_8bit_byteoptimized.cpp

@@ -106,8 +106,8 @@ int main(int argc, char *argv[]) {
 
     z = z^val;
   }
-  cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::chrono::steady_clock::time_point end = std::chrono::steady_clock::now();
+  cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::cout << "Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (end - begin).count() << "[ms]" << std::endl;
   return 0;
 }

### 2020-01-22 vb_encode_7bit_byteoptimized.cpp
adds large number packaging

@@ -66,9 +66,11 @@ int main(int argc, char *argv[]) {
   cout << "Original data size: " << fsize << endl;
 
   bit_vector b(vdata.size(), 0);
-  uint8_t iv [vdata.size()];
-  index = 0;
 
+  cout << "got here, vdata=" <<vdata.size() << endl;
+  uint8_t *iv  = new uint8_t[vdata.size()];
+  cout << "got here" << endl;
+  index = 0;
   for (vector<uint32_t>::const_iterator i = vdata.begin(); i != vdata.end(); i++, index++) {
     b[index] = (*i>>7) & 1;
     iv[index] = *i%cap;

### 2020-01-26 select_query_test.cpp
Changes

@@ -1,0 +1,232 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <sdsl/util.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+
+using namespace std;
+using namespace sdsl;
+
+const uint32_t cap = (uint32_t) 256;
+
+//NOTE: take note of your system's endianness when storing these
+vector<uint32_t> vb_encode_number(uint64_t number) {
+  vector<uint32_t> v;
+  while(1) {
+    v.push_back(number%cap);
+    if(number < cap) {
+      break;
+    }
+    number/=cap;
+  }
+  v.back() += cap;
+  return v;
+}
+
+
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  size_t fsize = util::file_size(argv[1]);
+  if(fsize%8) {
+    cout << "Wrong sized file" << endl;
+    return 1;
+  }
+  vector<uint32_t> vdata;
+  vector<uint64_t> original(fsize/8, 0);
+  ifstream idt;
+  idt.open(argv[1], ios::in|ios::binary);
+  size_t index = 0;
+  while( idt ) {
+      uint32_t val[2] = {0};
+      if (idt.read((char*)val, sizeof(val))) {
+          original[index] = (uint64_t)(val[1]) << 32 | (uint64_t)(val[0]);
+          index++;
+
+      }
+  }
+  idt.close();
+
+  //following for trimming out gaps
+//  sort(original.begin(), original.end());
+//  for(std::vector<uint64_t>::size_type i = original.size()-1; i > 1; i--) {
+//    original[i] -= original[i-1];
+//  }
+  //
+
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i);
+    vdata.insert(vdata.end(), v.begin(), v.end());
+  }
+  cout << "Original data size: " << fsize << endl;
+
+  bit_vector b(vdata.size(), 0);
+  uint8_t *iv  = new uint8_t[vdata.size()];
+  index = 0;
+
+  for (vector<uint32_t>::const_iterator i = vdata.begin(); i != vdata.end(); i++, index++) {
+    b[index] = (*i>>8) & 1;
[…]

### 2020-01-26 vb_encode.cpp

@@ -1,30 +1,16 @@
 #include <sdsl/bit_vectors.hpp>
 #include <sdsl/int_vector.hpp>
-#include <sdsl/util.hpp>
 #include <string>
 #include <vector>
 #include <ctime>
 #include <chrono>
+#include "vbyte_helpers.hpp"
 
 using namespace std;
 using namespace sdsl;
 
-const uint8_t cap = (uint8_t) 128;
-
-
-vector<uint8_t> vb_encode_number(uint64_t number) {
-  vector<uint8_t> v;
-  while(1) {
-    v.push_back(number%cap);
-    if(number < cap) {
-      break;
-    }
-    number/=cap;
-  }
-  *v.begin() += cap;
-  return v;
-}
-
+uint32_t bit_length = 7;
+unsigned int random_accesses = 1000000;
 
 int main(int argc, char *argv[]) {
   if(argc<=1) {
@@ -32,47 +18,21 @@ int main(int argc, char *argv[]) {
     return 1;
   }
 
-  size_t fsize = util::file_size(argv[1]);
-  if(fsize%8) {
-    cout << "Wrong sized file" << endl;
-    return 1;
-  }
-  vector<uint8_t> data;
-  vector<uint64_t> original(fsize/8, 0);
-  ifstream idt;
-  idt.open(argv[1], ios::in|ios::binary);
-  size_t index = 0;
-  while( idt ) {
-      uint32_t val[2] = {0};
-      if (idt.read((char*)val, sizeof(val))) {
-          original[index] = (uint64_t)(val[1]) << 32 | (uint64_t)(val[0]);
-          index++;
+  vector<uint64_t> original = read_data_from_file(argv[1]);
 
-      }
-  }
-  idt.close();
-
-  //following for trimming out gaps
-//  sort(original.begin(), original.end());
-//  for(std::vector<uint64_t>::size_type i = original.size()-1; i > 1; i--) {
-//    original[i] -= original[i-1];
-//  }
-  //
+  vector<uint32_t> data;
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint8_t> v = vb_encode_number(*i);
+    vector<uint32_t> v = vb_encode_number(*i, 1<<bit_length);
     data.insert(data.end(), v.rbegin(), v.rend());
   }
 
-  cout << "Original data size: " << fsize << endl;
[…]

### 2020-01-26 vb_encode_7bit_byteoptimized.cpp

@@ -1,113 +1,0 @@
-#include <sdsl/bit_vectors.hpp>
-#include <sdsl/int_vector.hpp>
-#include <sdsl/util.hpp>
-#include <string>
-#include <vector>
-#include <ctime>
-#include <chrono>
-
-using namespace std;
-using namespace sdsl;
-
-const uint32_t cap = (uint32_t) 128;
-
-//NOTE: take note of your system's endianness when storing these
-vector<uint32_t> vb_encode_number(uint64_t number) {
-  vector<uint32_t> v;
-  while(1) {
-    v.push_back(number%cap);
-    if(number < cap) {
-      break;
-    }
-    number/=cap;
-  }
-  v.back() += cap;
-  return v;
-}
-
-
-int main(int argc, char *argv[]) {
-  if(argc<=1) {
-    cerr << "Give uint64_t file as parameter" << endl;
-    return 1;
-  }
-
-  size_t fsize = util::file_size(argv[1]);
-  if(fsize%8) {
-    cout << "Wrong sized file" << endl;
-    return 1;
-  }
-  vector<uint32_t> vdata;
-  vector<uint64_t> original(fsize/8, 0);
-  ifstream idt;
-  idt.open(argv[1], ios::in|ios::binary);
-  size_t index = 0;
-  while( idt ) {
-      uint32_t val[2] = {0};
-      if (idt.read((char*)val, sizeof(val))) {
-          original[index] = (uint64_t)(val[1]) << 32 | (uint64_t)(val[0]);
-          index++;
-
-      }
-  }
-  idt.close();
-
-  //following for trimming out gaps
-//  sort(original.begin(), original.end());
-//  for(std::vector<uint64_t>::size_type i = original.size()-1; i > 1; i--) {
-//    original[i] -= original[i-1];
-//  }
-  //
-
-  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint32_t> v = vb_encode_number(*i);
-    vdata.insert(vdata.end(), v.begin(), v.end());
-  }
-  cout << "Original data size: " << fsize << endl;
-
-  bit_vector b(vdata.size(), 0);
-
-  cout << "got here, vdata=" <<vdata.size() << endl;
-  uint8_t *iv  = new uint8_t[vdata.size()];
-  cout << "got here" << endl;
-  index = 0;
-  for (vector<uint32_t>::const_iterator i = vdata.begin(); i != vdata.end(); i++, index++) {
-    b[index] = (*i>>7) & 1;
[…]

### 2020-01-26 vb_encode_8bit_byteoptimized.cpp

@@ -66,7 +66,7 @@ int main(int argc, char *argv[]) {
   cout << "Original data size: " << fsize << endl;
 
   bit_vector b(vdata.size(), 0);
-  uint8_t iv [vdata.size()];
+  uint8_t *iv  = new uint8_t[vdata.size()];
   index = 0;
 
   for (vector<uint32_t>::const_iterator i = vdata.begin(); i != vdata.end(); i++, index++) {
@@ -89,7 +89,9 @@ int main(int argc, char *argv[]) {
   for(vector<uint64_t>::size_type i = 0; i < indices.size(); i++) {
     indices[i] = rand()%original.size();
   }
-  std::chrono::steady_clock::time_point begin = std::chrono::steady_clock::now();
+
+
+  std::chrono::steady_clock::time_point time_begin = std::chrono::steady_clock::now();
   uint64_t z = 0;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
@@ -106,8 +108,9 @@ int main(int argc, char *argv[]) {
 
     z = z^val;
   }
-  std::chrono::steady_clock::time_point end = std::chrono::steady_clock::now();
+  std::chrono::steady_clock::time_point time_end = std::chrono::steady_clock::now();
   cout << "making sure optimizer doesn't steal our code: " << z << endl;
-  std::cout << "Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (end - begin).count() << "[ms]" << std::endl;
+  std::cout << "Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (time_end - time_begin).count() << "[ms]" << std::endl;
+
   return 0;
 }

### 2020-01-26 vb_encode_ref.cpp

@@ -5,11 +5,11 @@
 #include <ctime>
 #include <chrono>
 #include "vbyte_helpers.hpp"
-
+#include <cmath>
 using namespace std;
 using namespace sdsl;
 
-uint32_t bit_length = 8;
+uint32_t bit_length = 7;
 unsigned int random_accesses = 1000000;
 
 int main(int argc, char *argv[]) {
@@ -30,6 +30,9 @@ int main(int argc, char *argv[]) {
   bit_vector b(data.size(), 0);
   int_vector<> iv(data.size(), 0, bit_length);
   size_t index = 0;
+  unsigned int block_len = ((int)sqrt(original.size()-1))+1;
+  cout << "block_len" << block_len << endl;
+  int* blocks = new int[block_len];
 
   for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
     b[index] = (*i>>bit_length) & 1;
@@ -37,15 +40,19 @@ int main(int argc, char *argv[]) {
   }
 
   select_support_mcl<> sls(&b);
+  blocks[0] = 0; // because first data begins at 0
+  for(int b_i = 1; b_i < block_len; b_i++) {
+    blocks[b_i] = sls(b_i*block_len)+1;
+  }
 
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
   cout << "data vector memory size: " << size_in_bytes(iv) + sizeof(int_vector<>) << "bytes" << endl;
   cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;
 
   srand((unsigned) time(0));
-
+  uint32_t bitmask = ~((~0)<<bit_length);
+  cout << bitmask << endl;
   cout << "number of numbers: " << original.size() << endl;
-
   vector<unsigned int> indices(random_accesses, 0);
   for(vector<uint64_t>::size_type i = 0; i < indices.size(); i++) {
     indices[i] = rand()%original.size();
@@ -53,17 +60,29 @@ int main(int argc, char *argv[]) {
 
   chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   uint64_t z = 0;
+  uint32_t v;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
-
-
-    int begin = index == 0 ? 0 : sls(index)+1;
-    int end = sls(index+1);
-    uint64_t val = 0;
[…]

### 2020-01-26 vbyte_helpers.cpp

@@ -1,0 +1,42 @@
+#include <vector>
+#include <sdsl/util.hpp>
+
+#include <cstdint> // include this header for uint64_t
+using namespace std;
+
+vector<uint32_t> vb_encode_number(uint64_t number, uint32_t cap = 128) {
+  vector<uint32_t> v;
+  while(1) {
+    v.push_back(number%cap);
+    if(number < cap) {
+      break;
+    }
+    number/=cap;
+  }
+  *v.begin() += cap;
+  return v;
+}
+
+vector<uint64_t> read_data_from_file(char* filename) {
+  size_t fsize = sdsl::util::file_size(filename);
+  if(fsize%8) {
+    cout << "Wrong sized file" << endl;
+    exit(1);
+  }
+  cout << "Original data size: " << fsize << endl;
+
+  vector<uint64_t> original(fsize/8, 0);
+  ifstream idt;
+  idt.open(filename, ios::in|ios::binary);
+  size_t index = 0;
+  while( idt ) {
+      uint32_t val[2] = {0};
+      if (idt.read((char*)val, sizeof(val))) {
+          original[index] = (uint64_t)(val[1]) << 32 | (uint64_t)(val[0]);
+          index++;
+
+      }
+  }
+  idt.close();
+  return original;
+}

### 2020-01-26 vbyte_helpers.hpp

@@ -1,0 +1,11 @@
+#include <vector>
+using namespace std;
+
+#ifndef VBYTE_H
+#define VBYTE_H
+
+vector<uint32_t> vb_encode_number(uint64_t number, uint32_t cap);
+
+vector<uint64_t> read_data_from_file(char* filename);
+
+#endif

### 2020-01-27 generate.cpp
Fixed refence encoding, added distributed dataset generation

@@ -21,9 +21,10 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-
+//  int maxvalues [4] = {128,256,32768,65536};
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
+//    uint64_t i = dis(gen)%maxvalues[rand];
     uint64_t i = dis(gen)%max;
     file.write(reinterpret_cast<const char *>(&i), sizeof(uint64_t));
   }

### 2020-01-27 vb_encode_ref.cpp

@@ -30,9 +30,11 @@ int main(int argc, char *argv[]) {
   bit_vector b(data.size(), 0);
   int_vector<> iv(data.size(), 0, bit_length);
   size_t index = 0;
-  unsigned int block_len = ((int)sqrt(original.size()-1))+1;
+  //unsigned int block_len = ((int)sqrt(original.size()-1))+1;
+  unsigned int block_len = 1024;
+  unsigned int block_amount = original.size()/block_len +1;
   cout << "block_len" << block_len << endl;
-  int* blocks = new int[block_len];
+  int* blocks = new int[block_amount];
 
   for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
     b[index] = (*i>>bit_length) & 1;
@@ -41,10 +43,10 @@ int main(int argc, char *argv[]) {
 
   select_support_mcl<> sls(&b);
   blocks[0] = 0; // because first data begins at 0
-  for(int b_i = 1; b_i < block_len; b_i++) {
+  for(int b_i = 1; b_i < block_amount; b_i++) {
     blocks[b_i] = sls(b_i*block_len)+1;
   }
-
+  cout << "index vector size: " << sizeof(int)*block_amount << "bytes" << endl;
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
   cout << "data vector memory size: " << size_in_bytes(iv) + sizeof(int_vector<>) << "bytes" << endl;
   cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;

### 2020-01-27 vb_encode.cpp
modified 8bit optimized

@@ -11,7 +11,7 @@ using namespace sdsl;
 
 uint32_t bit_length = 7;
 unsigned int random_accesses = 1000000;
-
+uint32_t cap = 1<<bit_length;
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
@@ -23,7 +23,8 @@ int main(int argc, char *argv[]) {
   vector<uint32_t> data;
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint32_t> v = vb_encode_number(*i, 1<<bit_length);
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    *v.begin() += cap;
     data.insert(data.end(), v.rbegin(), v.rend());
   }
 

### 2020-01-27 vb_encode_8bit.cpp

@@ -11,7 +11,7 @@ using namespace sdsl;
 
 uint32_t bit_length = 8;
 unsigned int random_accesses = 1000000;
-
+uint32_t cap = 1<<bit_length;
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
@@ -23,7 +23,8 @@ int main(int argc, char *argv[]) {
   vector<uint32_t> data;
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint32_t> v = vb_encode_number(*i, 1<<bit_length);
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    *v.begin() += cap;
     data.insert(data.end(), v.rbegin(), v.rend());
   }
 

### 2020-01-27 vb_encode_8bit_byteoptimized.cpp

@@ -1,116 +1,90 @@
 #include <sdsl/bit_vectors.hpp>
 #include <sdsl/int_vector.hpp>
-#include <sdsl/util.hpp>
 #include <string>
 #include <vector>
 #include <ctime>
 #include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
 
 using namespace std;
 using namespace sdsl;
 
-const uint32_t cap = (uint32_t) 256;
-
-//NOTE: take note of your system's endianness when storing these
-vector<uint32_t> vb_encode_number(uint64_t number) {
-  vector<uint32_t> v;
-  while(1) {
-    v.push_back(number%cap);
-    if(number < cap) {
-      break;
-    }
-    number/=cap;
-  }
-  v.back() += cap;
-  return v;
-}
-
-
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
     return 1;
   }
 
-  size_t fsize = util::file_size(argv[1]);
-  if(fsize%8) {
-    cout << "Wrong sized file" << endl;
-    return 1;
-  }
-  vector<uint32_t> vdata;
-  vector<uint64_t> original(fsize/8, 0);
-  ifstream idt;
-  idt.open(argv[1], ios::in|ios::binary);
-  size_t index = 0;
-  while( idt ) {
-      uint32_t val[2] = {0};
-      if (idt.read((char*)val, sizeof(val))) {
-          original[index] = (uint64_t)(val[1]) << 32 | (uint64_t)(val[0]);
-          index++;
+  vector<uint64_t> original = read_data_from_file(argv[1]);
 
-      }
-  }
-  idt.close();
-
-  //following for trimming out gaps
-//  sort(original.begin(), original.end());
-//  for(std::vector<uint64_t>::size_type i = original.size()-1; i > 1; i--) {
-//    original[i] -= original[i-1];
-//  }
-  //
+  vector<uint32_t> data;
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint32_t> v = vb_encode_number(*i);
-    vdata.insert(vdata.end(), v.begin(), v.end());
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.back() += cap;
+    data.insert(data.end(), v.begin(), v.end());
   }
[…]

### 2020-01-27 vb_encode_ref.cpp

@@ -11,6 +11,7 @@ using namespace sdsl;
 
 uint32_t bit_length = 7;
 unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
 
 int main(int argc, char *argv[]) {
   if(argc<=1) {
@@ -23,7 +24,8 @@ int main(int argc, char *argv[]) {
   vector<uint32_t> data;
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
-    vector<uint32_t> v = vb_encode_number(*i, 1<<bit_length);
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    *v.begin() += cap;
     data.insert(data.end(), v.rbegin(), v.rend());
   }
 
@@ -31,7 +33,7 @@ int main(int argc, char *argv[]) {
   int_vector<> iv(data.size(), 0, bit_length);
   size_t index = 0;
   //unsigned int block_len = ((int)sqrt(original.size()-1))+1;
-  unsigned int block_len = 1024;
+  unsigned int block_len = 128;
   unsigned int block_amount = original.size()/block_len +1;
   cout << "block_len" << block_len << endl;
   int* blocks = new int[block_amount];

### 2020-01-27 vbyte_helpers.cpp

@@ -13,7 +13,6 @@ vector<uint32_t> vb_encode_number(uint64_t number, uint32_t cap = 128) {
     }
     number/=cap;
   }
-  *v.begin() += cap;
   return v;
 }
 

### 2020-01-27 vb_encode_8bit_byteoptimized.cpp
adds data

@@ -7,6 +7,7 @@
 #include <iterator>
 #include <iostream>
 #include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
 
 using namespace std;
 using namespace sdsl;
@@ -57,31 +58,37 @@ int main(int argc, char *argv[]) {
 
   chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   uint64_t z = 0;
+  bit_vector::iterator iter;
+  CALLGRIND_START_INSTRUMENTATION;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
 
 
     int begin = index == 0 ? 0 : sls(index)+1;
+
 //    bit_vector::iterator iter = b.begin()+begin;
 //    while(*iter == 0) {
 //      iter++;
 //    }
+//    uint8_t diff = std::distance(b.begin()+begin, iter);
 
     uint64_t *test = (uint64_t *)&iv[begin];
 
-//    uint8_t diff = std::distance(b.begin()+begin, iter);
     int end = sls(index+1);
     uint8_t diff = end-begin;
+
     uint64_t val = *test%(256<<(8*diff));
 
     z = z^val;
 
 
-    if(val != original[index]) {
+    if(0 && val != original[index]) {
       cout << "Did not match: " << val << " vs " << original[index] << endl;
       cout << "index " << index << endl;
     }
   }
+  CALLGRIND_STOP_INSTRUMENTATION;
+  CALLGRIND_DUMP_STATS;
 
   chrono::steady_clock::time_point time_end = chrono::steady_clock::now();
   cout << "checksum: " << z << endl;

### 2020-01-27 vb_encode_ref.cpp

@@ -83,7 +83,8 @@ int main(int argc, char *argv[]) {
       iter++;
     } while (((v>>bit_length) & 1) == 0);
     z = z^val;
-    if(val != original[index]) {
+
+    if(0 && val != original[index]) {
       cout << "Did not match: " << val << " vs " << original[index] << endl;
       cout << "index " << index << endl;
     }

### 2020-01-29 gradu/msc_english_contents.tex
adds large number packaging
Changes exponent
added outline
adds gradu

@@
 \chapter{Introduction}
+Enormous datasets are a common case in today's applications. Compressing the datasets is beneficial, because they 
+naturally decrease memory requirements but also are faster when compressed data is read from disk \citep{Zob95}.
 
-The following gives some superficial instructions for using this template for a Master's thesis.
-For guidelines on thesis writing you can consult various sources, for example, the Bachelor thesis template.
+One of the leading methods of data compression is variable-length coding \citep{Sal99}, where frequent sequences of data
+are represented with shorter codewords. Because the sequences of data have different lenghts when compressed, it is 
+not trivial to determine the exact location of a certain element. If this is required, the usual data compression algorithms are
+inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill.
 
-The thesis should have an introduction chapter. Other chapters can be named according to the topic. In the end, some summary chapter is needed; see Chapter~\ref{chapter:conclusions} for an example.
+However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
+to this is to decompress data from the beginning. An integer compressing method with fast random access is explained and compared
+existing state-of-the-art methods.
 
-\chapter{Figures and Tables}
+\chapter{Basic Concepts}
+Variable-byte encoding \citep{Wil99} is a method of compressing integers. It splits an integer into blocks of $b$ bits and then encodes
+the block as $b + c$. $c$ is a bit denoting whether $b$ has the most significant bits of the number or not. A good dataset for variable-byte encoding
+is a set with both very small and very large numbers, since he compression comes from omitting the leading zeros in the bit array. This method 
+loses in compression performance to other methods \citep{Bri09}, but they are very fast to decode and allow fast random access.
 
+- Explain rank / select
+
+\chapter{Previous Work}
+bl
+\chapter{Algorithm}
+here
+\chapter{Results}
+hmll
 \section{Figures}
@@
 \centering
-\caption{Experimental results.\label{table:results}}
-\begin{tabular}{l||l c c r} 
-Experiment & 128 & 256 & 32768 & 65536 \\ 
+\caption{Results with 100k entries (in milliseconds).\label{table:results}}
+\begin{tabular}{l||l c c c c r} 
+Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
 \hline \hline 
-$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18\\
-$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11\\
-$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03\\
-$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94\\
+$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
+$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
+$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
+$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\
+
+\hline
+%
+\end{tabular}
+\end{table}
+
+\begin{table}
+\centering
+\caption{Results with 1M entries (in milliseconds).\label{table:results}}
+\begin{tabular}{l||l c c c c r} 
+Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
+\hline \hline 
+$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
+$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
+$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
+$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\
 
@@
 \begin{verbatim}
+
 @article{einstein,

### 2020-01-29 generate.cpp

@@ -21,11 +21,12 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-//  int maxvalues [4] = {128,256,32768,65536};
+  int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31};
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
-//    uint64_t i = dis(gen)%maxvalues[rand];
-    uint64_t i = dis(gen)%max;
+    int rand = dis(gen)%8;
+    uint64_t i = dis(gen)%maxvalues[rand];
+//    uint64_t i = dis(gen)%max;
     file.write(reinterpret_cast<const char *>(&i), sizeof(uint64_t));
   }
   file.close();

### 2020-01-29 vb_encode_8bit_bris.cpp

@@ -1,0 +1,143 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[8];
+  std::array<std::vector<uint8_t>, 8> iv;
+  size_t index = 0;
+
+  for(int i = 0; i < 8; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  select_support_mcl<> sls[8];
+  rank_support_v<0> rb[8];
+  for(int i =0; i < 8; i++) {
+    select_support_mcl<> s(&b[i]);
+    rank_support_v<0> r(&b[i]);
+    sls[i] = s;
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
[…]

### 2020-01-29 vb_encode_8bit_byteoptimized.cpp

@@ -41,11 +41,12 @@ int main(int argc, char *argv[]) {
   }
 
   bit_vector::select_1_type sls(&b);
-
+  
   //select_support_mcl<> sls(&b);
 
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
   cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;
+  cout << "Data vector memory size: " << sizeof(uint8_t) * data.size() << "bytes" << endl;
 
   srand((unsigned) time(0));
 

### 2020-01-29 vb_encode_ref.cpp

@@ -25,7 +25,7 @@ int main(int argc, char *argv[]) {
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
     vector<uint32_t> v = vb_encode_number(*i, cap);
-    *v.begin() += cap;
+    *v.front() += cap;
     data.insert(data.end(), v.rbegin(), v.rend());
   }
 

### 2020-02-02 gradu/msc_english_contents.tex
Adds more text

@@
 Variable-byte encoding \citep{Wil99} is a method of compressing integers. It splits an integer into blocks of $b$ bits and then encodes
-the block as $b + c$. $c$ is a bit denoting whether $b$ has the most significant bits of the number or not. A good dataset for variable-byte encoding
-is a set with both very small and very large numbers, since he compression comes from omitting the leading zeros in the bit array. This method 
-loses in compression performance to other methods \citep{Bri09}, but they are very fast to decode and allow fast random access.
+the block as $b + c$. $c$ is a bit denoting whether $b$ has the most significant bits of the number or not. (example?) A good dataset for variable-byte encoding
+is a set with both very small and very large numbers, since the compression comes from omitting the leading zeros in the bit array. This method 
+loses in compression performance to other methods \citep{Bri09}, but decoding is fast.
 
-- Explain rank / select
+Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
+the index of ith 1 bit in the bit array. Both functions work in constant time (citation?) and they require only a few percents of extra space over
+the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
+version of random access has already been introduced \citep{Bri09}. It uses one rank call for each block after the first which makes the fetching very
+fast if the number in question fits to one block.
+
+An effective version with select is also possible. The index of the block where ith number starts is obtainable with select(i-1)+1, when i > 0 and 0 when i=0.
+By separating the $c$ bit array and $b$ block array, $b$ now contains variable-byte integers in readable form. Another upside is that functions 
+next(i) and previous(i) are conveniently available without any extra calculation and the runtime is not dependent on the size of the value.
+
+- explain how random access is good
 
@@
 bl
+
 \chapter{Algorithm}
-here
+ - modifications to basic implementation
+
 \chapter{Results}
-hmll
+ - comparison to basic implementation + Bri09
+ - 
+\chapter{Conclusion}
+ - here
+
+\chapter{Future work}
+ - something to improve / research?
+
 \section{Figures}

### 2020-02-03 gradu/msc_english_contents.tex
Adds more text

@@
 
-\chapter{Basic Concepts}
-Variable-byte encoding \citep{Wil99} is a method of compressing integers. It splits an integer into blocks of $b$ bits and then encodes
-the block as $b + c$. $c$ is a bit denoting whether $b$ has the most significant bits of the number or not. (example?) A good dataset for variable-byte encoding
-is a set with both very small and very large numbers, since the compression comes from omitting the leading zeros in the bit array. This method 
-loses in compression performance to other methods \citep{Bri09}, but decoding is fast.
+\chapter{Variable-byte encoding}
 
+Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. A good data set for 
+VB encoding is a list of different sized numbers. As an example, an inverted index used by document search engine stores word frequencies and 
+locations in documents. VB saves space here because index numbers need to support large values, but are usually significantly smaller. 
+It is also proven to improve search speed, because less bytes need to be read from a hard drive \citep{Sch02}.
+
+VB splits an integer into blocks of $b$ bits. A continuation bit is added to the block to form a chunk $c$ and then chunk non-empty chunks are stored.
+The continuation bit is set to 1 for the last block of the integer. During decoding, chunks are read and their blocks are joined until a continuation 
+bit is set to 1. An example implementation is showcased in Figure~\ref{fig:vbyte}.
+
+\begin{figure}[h!] 
+\centering 
+blblblblb
+testataan
+tätä
+\caption{VB encoding and decoding.\label{fig:vbyte}}
+\end{figure}
+
+Small lengths of $c$ can yield better
+compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and offer less effective compression. 
+Generally chunk length of 8 has been used because it gives a good average and handling bytes is convenient \citep{IRbook}.
+
+bl 
+It splits an integer into blocks of $b$ bits and then encodes
+the block as $b + c$ where $c$ is a bit denoting whether $b$ has the most significant bits of the number or not.
+
+This method loses in compression performance to other methods \citep{Bri09}, but decoding is fast.
+
+\chapter{Directly addressable codes?}
 Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
@@
 the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
-version of random access has already been introduced \citep{Bri09}. It uses one rank call for each block after the first which makes the fetching very
-fast if the number in question fits to one block.
+version of random access has already been introduced \citep{Bri09}.
 
-An effective version with select is also possible. The index of the block where ith number starts is obtainable with select(i-1)+1, when i > 0 and 0 when i=0.
-By separating the $c$ bit array and $b$ block array, $b$ now contains variable-byte integers in readable form. Another upside is that functions 
-next(i) and previous(i) are conveniently available without any extra calculation and the runtime is not dependent on the size of the value.
+Random access with select query is also possible.  By separating the $c$ bit array and $b$ block array, $b$ contains variable-byte integers 
+in readable form. Another upside is that functions next(i) and previous(i) are conveniently available. Rank implementation has
 
@@
 \chapter{Algorithm}
+
  - modifications to basic implementation
@@
  - comparison to basic implementation + Bri09
- - 
-\chapter{Conclusion}
- - here
-
-\chapter{Future work}
- - something to improve / research?
-
-\section{Figures}
-Figure~\ref{fig:logo} gives an example how to add figures to the document. Remember always to cite the figure in the main text.
-
-\begin{figure}[h!] 
-\centering 
-\includegraphics[width=0.3\textwidth]{HY-logo-ml.png}
-\caption{University of Helsinki flame-logo for Faculty of Science.\label{fig:logo}}
-\end{figure}
-
-\section{Tables}
-
-Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.
-
 \begin{table}
@@
 \end{table}
+
+\chapter{Conclusion}
+ - here
+
+\chapter{Future work}
+ - something to improve / research?
+
+\section{Figures}
+Figure~\ref{fig:logo} gives an example how to add figures to the document. Remember always to cite the figure in the main text.
+
+\begin{figure}[h!] 
+\centering 
+\includegraphics[width=0.3\textwidth]{HY-logo-ml.png}
+\caption{University of Helsinki flame-logo for Faculty of Science.\label{fig:logo}}
+\end{figure}
+
+\section{Tables}
+
+Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.
 

### 2020-02-06 gradu/msc_english_contents.tex
more stuff

@@
 However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
-to this is to decompress data from the beginning. An integer compressing method with fast random access is explained and compared
+to this is to decompress data from the beginning. An integer compressing method with fast random access is explained later and compared
 existing state-of-the-art methods.
@@
 
-Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. A good data set for 
-VB encoding is a list of different sized numbers. As an example, an inverted index used by document search engine stores word frequencies and 
-locations in documents. VB saves space here because index numbers need to support large values, but are usually significantly smaller. 
-It is also proven to improve search speed, because less bytes need to be read from a hard drive \citep{Sch02}.
+Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
+compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
+and allows constant time random access (TODO: link to later chapters). A good data set for VB encoding is a list of mostly small numbers with a 
+need to support larger ones.
 
-VB splits an integer into blocks of $b$ bits. A continuation bit is added to the block to form a chunk $c$ and then chunk non-empty chunks are stored.
-The continuation bit is set to 1 for the last block of the integer. During decoding, chunks are read and their blocks are joined until a continuation 
-bit is set to 1. An example implementation is showcased in Figure~\ref{fig:vbyte}.
+A search engine may use an inverted index of words in documents. For each word, a list of document IDs where the word appears is stored. 
+It may also store locations of the word in document for advanced search purposes. Usually these lists are preprocessed and stored 
+as gaps, storing the difference to previous number instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, 
+but because of gap storing the numbers are small. In contrast, rare words have only a few entries but the numbers stored are larger. These lists 
+are excellent data sets for VB. Apache Lucene has variable bytes as vInt and IBM DB2 has them as Variable Byte).
 
-\begin{figure}[h!] 
-\centering 
-blblblblb
-testataan
-tätä
-\caption{VB encoding and decoding.\label{fig:vbyte}}
+TODO: a chapter about existing work
+Usage of VB to compress inverted 
+index lists was first experimented in \citep{Sch02}. (TODO: find better examples or find references to these)
+
+VB splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
+on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
+integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
+\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
+are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
+
+\medskip
+\noindent\begin{minipage}{.5\textwidth}
+\begin{algorithmic}
+\Function{VBEncodeNumber}{$n$}
+\State $bytes\gets $ list
+\While{true}
+\State $bytes$.prepend($n \bmod 128$)
+\If{$n < 128$} \State break \EndIf
+\State $n\gets n$ div $128$
+\EndWhile
+\State $bytes$.last() += $128$
+\State \textbf{return} bytes
+\EndFunction
+\captionof{figure}{VByte encoding} \label{vbyte_enc}
+\end{algorithmic}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+\begin{algorithmic}
+\Function{VBEncode}{$numbers$}
+\State $bytestream\gets $ list
+\ForEach {$n \in numbers$}
+\State $bytes \gets$ VBEncodeNumber($n$)
+\State $bytestream$.extend($bytes$)
+\EndFor
+\State \textbf{return} bytestream
+\EndFunction
+\end{algorithmic}
+\end{minipage}
+\medskip
+
+Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
+are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
+is \texttt{00010 11000} and $n = 0$. Block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to left 
+equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
+$n = 42$. Because the previous continuous bit was 1, decoding this number has finished. An example implementation of encode and decode with block 
+length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.
+
+Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
+offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
+convenient \citep{Man08}.
+
+\begin{figure}[ht]
+\begin{algorithmic}
+\Function{VBDecode}{$bytestream$}
+\State $numbers\gets $ list
+\State $n\gets 0$
+\ForEach {$b \in bytestream$}
+\If{$b < 128$}
+\State $n\gets 128\times n $ + $b$
+\Else
+\State $n\gets 128\times n $ + $b$ - $128$
+\State $numbers$.append($n$)
+\State $n\gets 0$
+\EndIf
+\EndFor
+\State \textbf{return} numbers
+\EndFunction
+\end{algorithmic}
+\caption{VByte decoding} \label{vbyte_dec}
 \end{figure}
 
-Small lengths of $c$ can yield better
-compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and offer less effective compression. 
-Generally chunk length of 8 has been used because it gives a good average and handling bytes is convenient \citep{IRbook}.
++description
 
-bl 
-It splits an integer into blocks of $b$ bits and then encodes
-the block as $b + c$ where $c$ is a bit denoting whether $b$ has the most significant bits of the number or not.
+\medskip
+-motivation (see papers below)
 
-This method loses in compression performance to other methods \citep{Bri09}, but decoding is fast.
+\medskip
++pseudocode
+
+\medskip
++a few examples (1)
+
+\medskip
+-a review of related work should include:
+
+\medskip
+---the text book from which the initial description I sent you comes from
+
+\medskip
+---First use for inverted index compression (at least that I'm aware of):
+
+\medskip
+https://dl.acm.org/doi/10.1145/564376.564416
+
+\medskip
+---More recent incarnations of vbyte (just mention them):
+
+\medskip
+https://arxiv.org/abs/1709.08990
+
+\medskip
+https://arxiv.org/abs/1503.07387
+
+\medskip
 
@@
 \centering
-\caption{Results with 100k entries (in milliseconds).\label{table:results}}
+\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
 \begin{tabular}{l||l c c c c r} 
@@
 \centering
-\caption{Results with 1M entries (in milliseconds).\label{table:results}}
+\caption{Results with 1M entries (in milliseconds).\label{table:results2}}
 \begin{tabular}{l||l c c c c r} 
@@
 \section{Figures}
-Figure~\ref{fig:logo} gives an example how to add figures to the document. Remember always to cite the figure in the main text.
-
-\begin{figure}[h!] 
-\centering 
-\includegraphics[width=0.3\textwidth]{HY-logo-ml.png}
-\caption{University of Helsinki flame-logo for Faculty of Science.\label{fig:logo}}
-\end{figure}
+Figure gives an example how to add figures to the document. Remember always to cite the figure in the main text.
 
@@
 
-Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.
-
-\chapter{Citations}
-
-\section{Citations to literature}
-
-References are listed in a separate .bib-file. In this case it is named \texttt{bibliography.bib} including the following content:
-\begin{verbatim}
-
-@article{einstein,
-    author =       "Albert Einstein",
-    title =        "{Zur Elektrodynamik bewegter K{\"o}rper}. ({German})
-        [{On} the electrodynamics of moving bodies]",
-    journal =      "Annalen der Physik",
-    volume =       "322",
-    number =       "10",
-    pages =        "891--921",
-    year =         "1905",
-    DOI =          "http://dx.doi.org/10.1002/andp.19053221004"
-}
-
-@book{latexcompanion,
-    author    = "Michel Goossens and Frank Mittelbach and Alexander Samarin",
-    title     = "The \LaTeX\ Companion",
-    year      = "1993",
-    publisher = "Addison-Wesley",
-    address   = "Reading, Massachusetts"
-}
-
-@misc{knuthwebsite,
-    author    = "Donald Knuth",
-    title     = "Knuth: Computers and Typesetting",
-    url       = "http://www-cs-faculty.stanford.edu/%7Eknuth/abcde.html"
-}
-\end{verbatim}
-
-In the last reference url field the code \verb+%7E+ will translate into \verb+~+ once clicked in the final pdf.
-
-References are created using command \texttt{\textbackslash cite\{einstein\}}, showing as \citep{einstein}. Other examples: \citep{latexcompanion,knuthwebsite}.
-
-Citation style can be negotiated with the supervisor. See some options in \url{https://www.sharelatex.com/learn/Bibtex_bibliography_styles}.
-
-\section{Crossreferences}
-
-Appendix~\ref{appendix:model} on page~\pageref{appendix:model} contains some additional material.
+Table gives an example how to report experimental results. Remember always to cite the table in the main text.
 

### 2020-02-06 gradu/temp/msc_english_contents.tex
more stuff

--- 2020-02-03 gradu/msc_english_contents.tex
@@
 
-\begin{figure}[h!] 
-\centering 
-blblblblb
-testataan
-tätä
-\caption{VB encoding and decoding.\label{fig:vbyte}}
-\end{figure}
+\begin{algorithm}
+\KwData{this text}
+\KwResult{how to write algorithm with \LaTeX2e}
+\LinesNumbered
+\While{not at end of this document}{read current\;
+  \eIf{understand}{
+  go to next section\;
+  current section becomes this one\;
+  }{
+    go back to the beginning of current section\;
+  }
+}
+\caption{VB encoding.\label{fig:vbyte}}
+\end{algorithm}
 

### 2020-02-07 gradu/msc_english_contents.tex
adds more text

@@
 
-Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
+Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
 compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
-and allows constant time random access (TODO: link to later chapters). A good data set for VB encoding is a list of mostly small numbers with a 
-need to support larger ones.
+and allows constant time random access. This is later explained in (TODO: link to later chapter).
 
-A search engine may use an inverted index of words in documents. For each word, a list of document IDs where the word appears is stored. 
-It may also store locations of the word in document for advanced search purposes. Usually these lists are preprocessed and stored 
-as gaps, storing the difference to previous number instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, 
-but because of gap storing the numbers are small. In contrast, rare words have only a few entries but the numbers stored are larger. These lists 
-are excellent data sets for VB. Apache Lucene has variable bytes as vInt and IBM DB2 has them as Variable Byte).
+A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
+words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
+advanced search purposes. Usually these lists are preprocessed and stored as gaps, storing the difference to previous number instead of the actual 
+number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
+have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.
 
-TODO: a chapter about existing work
-Usage of VB to compress inverted 
-index lists was first experimented in \citep{Sch02}. (TODO: find better examples or find references to these)
-
-VB splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
+VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
 on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
@@
 
-Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
-offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
-convenient \citep{Man08}.
-
 \begin{figure}[ht]
@@
 
-+description
+Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
+offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
+convenient \citep{Man08}.
 
-\medskip
--motivation (see papers below)
+VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
+in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
+(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first used to compress
+lists of word locations in documents \citep{Sch02}. It yielded excellent records, and since then many different approaches have been introduced.
 
-\medskip
-+pseudocode
+history and current implementations etc here
+ - mainitse IR book, Lem15, Pla18
 
-\medskip
-+a few examples (1)
-
-\medskip
--a review of related work should include:
-
-\medskip
----the text book from which the initial description I sent you comes from
-
-\medskip
----First use for inverted index compression (at least that I'm aware of):
-
-\medskip
-https://dl.acm.org/doi/10.1145/564376.564416
-
-\medskip
----More recent incarnations of vbyte (just mention them):
-
-\medskip
-https://arxiv.org/abs/1709.08990
-
-\medskip
-https://arxiv.org/abs/1503.07387
-
-\medskip
+TODO: a chapter about existing work
 

### 2020-02-08 gradu/gradu.tex
more gradu stuff

@@
 
-\documentclass[english,twoside,censored,tkt,alko-line]{HYthesisML}
+\documentclass[english,twoside,censored,cs,alko-line]{HYthesisML}
 
@@
 %% for APA = Harvard style = authoryear -> (Einstein, 1905) use:
-\usepackage[style=authoryear,bibstyle=authoryear,backend=biber,natbib=true,maxnames=99,maxcitenames=2,giveninits=true,uniquename=init]{biblatex}
+\usepackage[style=authoryear,bibstyle=authoryear,backend=bibtex,natbib=true,maxnames=99,maxcitenames=2,giveninits=true,uniquename=init]{biblatex}
 %% for numeric = Vancouver style -> [1] use:
@@
 \usepackage{amsmath, amssymb} % For better math
+\usepackage{algorithm}  
+\usepackage[noend]{algpseudocode}
+\singlespacing               %line spacing options; normally use single
 
-\singlespacing               %line spacing options; normally use single
+\algnewcommand\algorithmicforeach{\textbf{for each}}
+\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
 
@@
 %% Replace parameters with your information.
-\title{Tutkielman otsikko}
+\title{something about variable-byte encoding}
 
@@
 \begin{abstract}
-
-Tämä dokumentti on tarkoitettu Helsingin yliopiston tietojenkäsittelytieteen osaston opin\-näyt\-teiden ja harjoitustöiden ulkoasun ohjeeksi ja mallipohjaksi.
-Ohje soveltuu kanditutkielmiin, ohjelmistotuotantoprojekteihin, seminaareihin ja maisterintutkielmiin.
-Tämän ohjeen lisäksi on seurattava niitä ohjeita, jotka opastavat valitsemaan kuhunkin osioon tieteellisesti kiinnostavaa, syvällisesti pohdittua sisältöä.
-
-Työn aihe luokitellaan  
-ACM Computing Classification System (CCS) mukaisesti, 
-ks.\ \url{https://www.acm.org/about-acm/class},
-käyttäen komentoa \verb+\classification{}+. 
-Käytä muutamaa termipolkua (1--3), jotka alkavat juuritermistä ja joissa polun tarkentuvat luokat erotetaan toisistaan oikealle osoittavalla nuolella.
-
-\end{abstract}
-
-\begin{otherlanguage}{english}
-\begin{abstract}
 Use this otherlanguage environment to write your abstract in another language if needed.
@@
 \end{abstract}
-\end{otherlanguage}
 
@@
 \backmatter
-\begin{appendices}
+%\begin{appendices}
 
-\input{instructions_english}
+%\input{instructions_english}
 %\input{instructions_finnish}
 
-\appendix{Sample Appendix\label{appendix:model}}
-usually starts on its own page, with the name and number of the appendix at the top. 
-The appendices here are just models of the table of contents and the presentation. Each appendix
-Each appendix is paginated separately.
-
-In addition to complementing the main document, each appendix is also its own, independent entity.
-This means that an appendix cannot be just an image or a piece of programming, but the appendix must explain its contents and meaning.
-
-\end{appendices}
+%\end{appendices}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-02-08 gradu/msc_english_contents.tex
more gradu stuff

@@
 compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
-and allows constant time random access. This is later explained in (TODO: link to later chapter).
+and sometimes preferred due to it's simplicity. allows constant time random access. This is later explained in (TODO: link to later chapter).
 
@@
 words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
-advanced search purposes. Usually these lists are preprocessed and stored as gaps, storing the difference to previous number instead of the actual 
+advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number instead of the actual 
 number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
@@
 in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
-(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first used to compress
-lists of word locations in documents \citep{Sch02}. It yielded excellent records, and since then many different approaches have been introduced.
+(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first experimented in 
+compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different approaches 
+have been introduced.
 
-history and current implementations etc here
- - mainitse IR book, Lem15, Pla18
+Modern studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) architecture to VB \citep{Lem18,Pla15}. The bit 
+operations in VB are simple and therefore modifying the code to use SIMD instructions is simple and the speed improvements are significant.
 
-TODO: a chapter about existing work
+TODO: might need 1-2 more chapters
 
 \chapter{Directly addressable codes?}
+
+Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
+data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not possible
+to decode i-th element in known well-performing compression algorithms like Huffman or Lempel-Ziv, because the position of the element depends on the data 
+before it. While VB encoding doesn't natively support direct access, it is achievable with two succinct data structure operations: rank and select.
+
+<explain rank>
+
+<explain select>
+
+<introduce Bri09>
+
+<example of Bri09>
+
+- TODO: find out where rank+select initially come from.
+
 Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
@@
 
-\chapter{Previous Work}
+\chapter{Rearranged VB codes}
 bl
 
-\chapter{Algorithm}
-
- - modifications to basic implementation
-
-\chapter{Results}
+\chapter{Experimental results}
  - comparison to basic implementation + Bri09

### 2020-02-10 gradu/msc_english_contents.tex
.

@@
 Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
-data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not possible
-to decode i-th element in known well-performing compression algorithms like Huffman or Lempel-Ziv, because the position of the element depends on the data 
-before it. While VB encoding doesn't natively support direct access, it is achievable with two succinct data structure operations: rank and select.
+data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not natively
+possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the preceding data. Direct access 
+is achievable with supporting data structures. An obvious solution is to store each element's location, but that adds a very large overhead which removes the 
+benefit of compression, but for some compression techniques it adds a very large overhead and thus is not applicable.
 
-<explain rank>
+Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
+$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
+space over the bit array B (citation?). Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, 
+this requires B to be created in addition to the existing data and the length of B usually has to be close to the length of the data.
 
-<explain select>
+VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which significantly shrinks the length of B. Also 
+the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to 
+i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.
 
-<introduce Bri09>
+\section{DAC via rank}
+Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
+$A_n$ and then use $rank_1$ to get direct access. For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second 
+least significant bits to $A_2$ and the most significant bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. 
+If the continuation bit is set at 0 (meaning there are more chunks to this element), getting the correct index for $A_2$ is calculated with $rank_1$($A_1$, i). 
+Small values benefit from this reordering, because the first chunk does not need any calculation. This method causes a small overhead on the data, because each 
+continuation bit array needs a support data structure for rank.
+
+(TODO: check if actually first introduced or just first with rank)
+
+do rank/select need to be explained better? (or just referenced to)
+
+<introduce Bri09 better?>
 
@@
 
-Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
-the index of ith 1 bit in the bit array. Both functions work in constant time (citation?) and they require only a few percents of extra space over
-the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
+ An effective
 version of random access has already been introduced \citep{Bri09}.
 
-Random access with select query is also possible.  By separating the $c$ bit array and $b$ block array, $b$ contains variable-byte integers 
-in readable form. Another upside is that functions next(i) and previous(i) are conveniently available. Rank implementation has
+- explain how random access is good ?
 
-- explain how random access is good
+\chapter{DAC with select query}
 
-\chapter{Rearranged VB codes}
-bl
+Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
+in $B$ and a single $select_1$ query gets the desired starting point.
 
@@
  - comparison to basic implementation + Bri09
+
+ - do we need to support search()?
+
 \begin{table}
@@
 
-\chapter{From tex to pdf}
-
-In Linux, run \texttt{pdflatex filename.tex} and \texttt{bibtex filename.tex} repeatedly until no more warnings are shown. This process can be automised using make-command.
-
 \chapter{Conclusions\label{chapter:conclusions}}

### 2020-02-12 gradu/msc_english_contents.tex
DAC with select

@@
 $select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
-space over the bit array B (citation?). Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, 
-this requires B to be created in addition to the existing data and the length of B usually has to be close to the length of the data.
+space over the bit array B. See Table~\ref{supportsize} for $rank$ and $select$ size requirement approximations on different sizes of bit arrays.
 
-VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which significantly shrinks the length of B. Also 
-the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to 
-i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.
+Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, this requires B to be created in addition 
+to the existing data and the length of B usually has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data 
+is compressed in blocks of equal length which significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the 
+locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending 
+byte of i-th compressed element.
+
+\begin{table}
+\centering
+\caption{Size of support structure with SDSL-lite library}
+\begin{tabular}{l||c c c c c c} 
+Structure & TODO & get & multiple & size & calculations & here\\ 
+\hline \hline 
+$rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
+$select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
+\hline
+%
+\end{tabular}
+\label{supportsize}
+\end{table}
 
@@
 
-(TODO: check if actually first introduced or just first with rank)
-
-do rank/select need to be explained better? (or just referenced to)
+do rank/select algorithms + data structures need to be explained better? (or just referenced to)
 
@@
 <example of Bri09>
-
-- TODO: find out where rank+select initially come from.
-
- An effective
-version of random access has already been introduced \citep{Bri09}.
 
@@
 Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
-in $B$ and a single $select_1$ query gets the desired starting point.
+in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
+the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
+consists of a few bit operations.
+
+To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. $Select$(i) returns the
+location of the end byte of i-th element. Therefore the start of j-th element in the block array is at block $b_s$ = $select_1$(j-1) + 1. To simplify calculations even further,
+block sizes of 2, 4 and 8 were used to avoid splitting of blocks between bytes.
+
+The blocks should be stored in an order that, depending on the endianness of the system, when reading an unsigned integer from the block array, the blocks are already 
+in correct order. This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
+element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
+is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
+32-b*k to re-align bits and remove leading bits.
+
+Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract already calculated start block index from it. This causes a second $select_1$
+query, which is suboptimal. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from the bit array, fix
+it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See Table~\ref{endblockcall} for 
+detailed times.
+
+\begin{table}
+\centering
+\caption{Time to calculate end blocks}
+\begin{tabular}{l||c c c c c c} 
+Version & Average time over 10Mall \\ 
+\hline \hline 
+select(i+1) & 34.97 \\
+iterate bit array & 34.97 \\
+bit shift magic & 34.97 \\
+\hline
+%
+\end{tabular}
+\label{endblockcall}
+\end{table}
+
+(maybe to future work) Storing bytes in this fashion may have benefits when reading subsequent (fast next() and previous()?) elements.
+With some modifications, SIMD instructions can possibly be applied.
+
+TODO: check if rank query takes less when done on single byte array instead of multiples
+
+ - get actual results of rank vs select support size
+ - take advantage of how data is stored
 
 \chapter{Experimental results}
+ - data sets
+
 - comparison to basic implementation + Bri09
+  - compare with b=2,4,8
 
  - do we need to support search()?
+ - SIMD?
 
@@
 
-\section{Figures}
-Figure gives an example how to add figures to the document. Remember always to cite the figure in the main text.
-
-\section{Tables}
-
-Table gives an example how to report experimental results. Remember always to cite the table in the main text.
-
 \chapter{Conclusions\label{chapter:conclusions}}

### 2020-02-12 vb_encode_8bit_bris.cpp

@@ -7,7 +7,7 @@
 #include <iterator>
 #include <iostream>
 #include "vbyte_helpers.hpp"
-#include <valgrind/callgrind.h>
+//#include <valgrind/callgrind.h>
 
 using namespace std;
 using namespace sdsl;
@@ -83,10 +83,12 @@ int main(int argc, char *argv[]) {
 
   chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   uint64_t z = 0;
+  int level = 0;
+  uint64_t val = 0;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
-    int level = 0;
-    uint64_t val = 0;
+    level = 0;
+    val = 0;
     while(b[level][index] == 0) {
 //      cout << "orig:" << original[*i] << endl;
 //      cout << "l:" <<level << " " << index << endl;
@@ -121,7 +123,7 @@ int main(int argc, char *argv[]) {
     z = z^val;
 
 
-    if(val != original[*i]) {
+    if(0 && val != original[*i]) {
       cout << "Did not match: " << val << " vs " << original[*i] << endl;
       cout << original[*i]%cap << " " << original[*i]/cap << endl;
 //      cout << "index " << index << endl;

### 2020-02-12 vb_encode_8bit_byteoptimized.cpp

@@ -6,6 +6,7 @@
 #include <chrono>
 #include <iterator>
 #include <iostream>
+#include <bitset>
 #include "vbyte_helpers.hpp"
 #include <valgrind/callgrind.h>
 
@@ -41,13 +42,12 @@ int main(int argc, char *argv[]) {
   }
 
   bit_vector::select_1_type sls(&b);
-  
+
   //select_support_mcl<> sls(&b);
 
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
   cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;
   cout << "Data vector memory size: " << sizeof(uint8_t) * data.size() << "bytes" << endl;
-
   srand((unsigned) time(0));
 
   cout << "number of numbers: " << original.size() << endl;
@@ -57,34 +57,56 @@ int main(int argc, char *argv[]) {
     indices[i] = rand()%original.size();
   }
 
-  chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   uint64_t z = 0;
-  bit_vector::iterator iter;
-  CALLGRIND_START_INSTRUMENTATION;
+//  CALLGRIND_START_INSTRUMENTATION;
+uint64_t *test;
+int end;
+uint8_t diff;
+uint64_t val;
+int begin;
+int bsize = sizeof(b.data()) * 8;
+cout << "bsize" << bsize << endl;
+int shiftamt;
+uint64_t maxuint = 0-1;
+cout << maxuint << endl;
+chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
+CALLGRIND_START_INSTRUMENTATION;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
 
-
-    int begin = index == 0 ? 0 : sls(index)+1;
-
+//TODO CALLGRIND THIS
+    begin = index == 0 ? 0 : sls(index)+1;
+    test = (uint64_t *)&iv[begin];
+//    while(*(b.begin()+begin+diff)==0) diff++;
 //    bit_vector::iterator iter = b.begin()+begin;
 //    while(*iter == 0) {
 //      iter++;
 //    }
 //    uint8_t diff = std::distance(b.begin()+begin, iter);
-
-    uint64_t *test = (uint64_t *)&iv[begin];
-
-    int end = sls(index+1);
[…]

### 2020-02-18 gradu/msc_english_contents.tex
adds v5 and 4bit

@@
 Modern studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) architecture to VB \citep{Lem18,Pla15}. The bit 
-operations in VB are simple and therefore modifying the code to use SIMD instructions is simple and the speed improvements are significant.
+operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
 
@@
 is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
-32-b*k to re-align bits and remove leading bits.
+32-b*k to re-align bits.
 
-Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract already calculated start block index from it. This causes a second $select_1$
-query, which is suboptimal. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from the bit array, fix
-it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See Table~\ref{endblockcall} for 
-detailed times.
+Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
+$select_1$ query, which costs a lot timewise. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
+the bit array, fix it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See 
+Table~\ref{endblockcall} for detailed times.
 
@@
 
+ - emphasis on size!
+
+     RANK seems a lot faster, check algorithms!!!
+TODO: try bitvector SD performance + size (select)
+
+TODO: look into select implementation, may need optimizing
+
 \chapter{Experimental results}
- - data sets
 
-- comparison to basic implementation + Bri09
-  - compare with b=2,4,8
+The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server>.
 
- - do we need to support search()?
- - SIMD?
+Five different kinds of VB decoding algorithms were used. Bris-implementations are from \citep{Bri09}. Rank and select functions are from
+ \citep{gbmp2014sea}.
+
+\begin{itemize}
+  \item Bris4v5 - Bris implementation with block size 4, using rank support v5
+  \item Bris8 - Bris implementation with block size 8, using rank support v
+  \item Bris8v5 - Bris implementation with block size 8, using rank support v5
+  \item Our4 - Our implementation with block size 4, using select support mcl
+  \item Our8 - Our implementation with block size 8, using select support mcl
+\end{itemize}
+
+Eight kinds of datasets were used. For each number in the dataset, $p$ was randomly selected from $P$ and then number was randomly selected from
+range [0,$2^p$].
+
+\begin{itemize}
+  \item all5M - 5 million numbers, $P$ = {7,8,15,16,23,24,30,31}
+  \item all50M - 50 million numbers, $P$ = {7,8,15,16,23,24,30,31}
+  \item byte5M - 5 million numbers, $P$ = {7,7,7,8,8,8,16,31}
+  \item byte50M - 50 million numbers, $P$ = {7,7,7,8,8,8,16,31}
+  \item small5M - 5 million numbers, $P$ = {3,4,5,6,7,8,16,31}
+  \item small50M - 50 million numbers, $P$ = {3,4,5,6,7,8,16,31}
+  \item vsmall5M - 5 million numbers, $P$ = {2,2,3,3,3,4,4,15}
+  \item vsmall50M - 50 million numbers, $P$ = {2,2,3,3,3,4,4,15}
+\end{itemize}
+
+All algorithms save the data to memory and randomize one million index numbers to an array. The time taken is calculated from looping through the 
+index array and VB decoding the number in the index. The times shown in Table~\ref{table:results1} are an average of 100 such runs.
 
@@
 \centering
-\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
-\begin{tabular}{l||l c c c c r} 
-Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
+\caption{Results in milliseconds, smaller is better.\label{table:results1}}
+\begin{tabular}{l||c c c c c c c c} 
+Experiment & all5M & all50M & byte5M & byte50M & small5M & small50M & vsmall5M & vsmall50M\\ 
 \hline \hline 
-$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
-$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
-$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
-$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\
+Bris4v5 & 348.41 & 768.83 & 345.16 & 772.84 & 344.54 & 768.45 & 346.86 & 771.61 \\
+Our4    & 127.92 & 263.22 & 127.44 & 263.5  & 127.9  & 262.67 & 127.88 & 262.58 \\
+Bris8   & \textbf{97.75}  & 308.17 & \textbf{98.47}  & 309.6  & \textbf{98.14}  & 307.81 & \textbf{97.99}  & 308.09 \\
+Bris8v5 & 127.83 & 287.46 & 127.7  & 288.63 & 128.14 & 291.32 & 127.99 & 287.84 \\
+Our8    & 103.85 & \textbf{230.16} & 103.92 & \textbf{230.14} & 103.86 & \textbf{230.02} & 103.95 & \textbf{230.6} \\
 
@@
 
-\begin{table}
-\centering
-\caption{Results with 1M entries (in milliseconds).\label{table:results2}}
-\begin{tabular}{l||l c c c c r} 
-Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
-\hline \hline 
-$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
-$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
-$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
-$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\
+ - use array instead of vector<uint8 t> for bris?
+- comparison to basic implementation + Bri09
+  - compare with b=2,4,8
 
-\hline
-%
-\end{tabular}
-\end{table}
+ - do we need to support search()?
+ - SIMD?
 

### 2020-02-18 generate.cpp

@@ -21,7 +21,10 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-  int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31};
+  // int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
+  // int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- smallheavy
+  // int maxvalues [8] = {1<<3,1<<4,1<<5,1<<6,1<<7,1<<8,1<<16,1<<31}; // <- vsmall
+  int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- vvsmallheavy
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
     int rand = dis(gen)%8;

### 2020-02-18 select_query_test.cpp

@@ -112,27 +112,6 @@ int main(int argc, char *argv[]) {
   cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::cout << "SLS_MCL Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (time_end - time_begin).count() << "[ms]" << std::endl;
 
-  select_support_scan<> sls_scan(&b);
-  time_begin = std::chrono::steady_clock::now();
-  z = 0;
-  for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
-    index = *i;
-
-//    cout << "original index " << index << " has " << original[index];
-    int begin = index == 0 ? 0 : sls_scan(index)+1; //
-    int end = sls_scan(index+1); // takes around 20ms
-    //int end = begin+1;
-//    cout << "Compressed data is between indices " << begin << "-" << end << endl;
-    //uint64_t val = 0;
-    uint64_t *test = (uint64_t *)&iv[begin];
-    uint8_t diff = end-begin;
-    uint64_t val = *test%(256<<(8*diff));
-
-    z = z^val;
-  }
-    time_end = std::chrono::steady_clock::now();
-  cout << "making sure optimizer doesn't steal our code: " << z << endl;
-  std::cout << "SLS_scan Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (time_end - time_begin).count() << "[ms]" << std::endl;
 
   bit_vector_il<> b_il(b);
   select_support_il<> sls_il(&b_il);
@@ -156,6 +135,8 @@ int main(int argc, char *argv[]) {
     time_end = std::chrono::steady_clock::now();
   cout << "making sure optimizer doesn't steal our code: " << z << endl;
   std::cout << "SLS_il Time taken: " << std::chrono::duration_cast<std::chrono::milliseconds> (time_end - time_begin).count() << "[ms]" << std::endl;
+  cout << "SLS_IL support vector memory size: " << size_in_bytes(sls_il) + sizeof(select_support_mcl<>) << "bytes" << endl;
+
   sd_vector<> b_sd(b);
   select_support_sd<> sls_sd(&b_sd);
   time_begin = std::chrono::steady_clock::now();
@@ -178,6 +159,8 @@ int main(int argc, char *argv[]) {
     time_end = std::chrono::steady_clock::now();
[…]

### 2020-02-18 vb_encode_4bit_bris_v5.cpp

@@ -12,7 +12,7 @@
 using namespace std;
 using namespace sdsl;
 
-uint32_t bit_length = 8;
+uint32_t bit_length = 4;
 unsigned int random_accesses = 1000000;
 uint32_t cap = 1<<bit_length;
 int main(int argc, char *argv[]) {
@@ -23,7 +23,7 @@ int main(int argc, char *argv[]) {
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
 
-  std::array<std::vector<uint32_t>, 8> data;
+  std::array<std::vector<uint32_t>, 16> data;
   for(int i = 0; i < 8; i++) {
     vector<uint32_t> vec;
     data[i] = vec;
@@ -41,11 +41,10 @@ int main(int argc, char *argv[]) {
     }
   //  data.insert(data.end(), v.begin(), v.end());
   }
-  bit_vector b[8];
-  std::array<std::vector<uint8_t>, 8> iv;
+  bit_vector b[16];
+  std::array<std::vector<uint8_t>, 16> iv;
   size_t index = 0;
-
-  for(int i = 0; i < 8; i++) {
+  for(int i = 0; i < 16; i++) {
     vector<uint8_t> vec(data[i].size(),0);
     bit_vector bv(data[i].size(), 0);
     index = 0;
@@ -59,12 +58,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  select_support_mcl<> sls[8];
-  rank_support_v<0> rb[8];
-  for(int i =0; i < 8; i++) {
-    select_support_mcl<> s(&b[i]);
-    rank_support_v<0> r(&b[i]);
-    sls[i] = s;
+  rank_support_v5<0> rb[16];
+  for(int i =0; i < 16; i++) {
+    rank_support_v5<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);

### 2020-02-18 vb_encode_4bit_byteoptimized.cpp

@@ -13,9 +13,11 @@
 using namespace std;
 using namespace sdsl;
 
-uint32_t bit_length = 8;
+uint32_t bit_length = 4;
 unsigned int random_accesses = 1000000;
 uint32_t cap = 1<<bit_length;
+int bsize = 64;
+
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
@@ -23,8 +25,18 @@ int main(int argc, char *argv[]) {
   }
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
+//  uint8_t *siv  = new uint8_t[8];
+//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
+//  uint64_t *x = (uint64_t *)&siv[0];
+//  *x = original[0]<<4;
+//  cout << *x << endl;
+//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
+//  exit(1);
 
   vector<uint32_t> data;
+
+//  REDO THIS:
+//  calculate bits on the fly and save number to block array intact
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
     vector<uint32_t> v = vb_encode_number(*i, cap);
@@ -32,13 +44,35 @@ int main(int argc, char *argv[]) {
     data.insert(data.end(), v.begin(), v.end());
   }
 
+
   bit_vector b(data.size(), 0);
   uint8_t *iv  = new uint8_t[data.size()];
   size_t index = 0;
-
+  size_t num_index = -1;
+  size_t num_p = 0;
+  uint8_t overflow = 0;
   for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
     b[index] = (*i>>bit_length) & 1;
-    iv[index] = *i;
+    if(b[index] == 1) {
+        uint64_t orig = original[num_p];
+        int diff = index-num_index;
+        int shiftsize = bsize-(diff*bit_length);
+        orig = (orig<<shiftsize)>>shiftsize;
+        num_index++;
+        if(num_index%2 == 1) {
+          if(diff == 16) {
+            overflow = orig>>60;
+          }
+          orig = orig<<4;
+        }
[…]

### 2020-02-18 vb_encode_8bit_bris_v5.cpp

@@ -59,12 +59,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  select_support_mcl<> sls[8];
-  rank_support_v<0> rb[8];
+  rank_support_v5<0> rb[8];
   for(int i =0; i < 8; i++) {
-    select_support_mcl<> s(&b[i]);
-    rank_support_v<0> r(&b[i]);
-    sls[i] = s;
+    rank_support_v5<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);

### 2020-02-19 gradu/msc_english_contents.tex
Added pseudocodes

@@
 $rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
+$rank_v5$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
 $select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
@@
 continuation bit array needs a support data structure for rank. 
+A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. In the actual implementation, different block size requires
+extra calculation to get the block location from the byte array. (TODO: do i need to explain how to get the block location?) (TODO: does this need an example?)
 
-do rank/select algorithms + data structures need to be explained better? (or just referenced to)
+\begin{figure}[ht]
+\begin{algorithmic}
+\State $i \gets $ wanted index
+\State $A \gets $ block arrays
+\State $B \gets $ continuation bit arrays
+\State $level \gets 0$
+\State $number \gets 0$
+\While {$B[level][index] = 0$}
+\State $block \gets A[level][index]$
+\State $number \gets number \mathbin{\ll} 8$
+\State $number \gets number \mathbin{|} block$
+\State $index \gets \Call{Rank}{B[level], index}$
+\State $level \gets level + 1$
+\EndWhile
+\State $block \gets A[level][index]$
+\State $number \gets number \mathbin{\ll} 8$
+\State $number \gets number \mathbin{|} block$
 
-<introduce Bri09 better?>
-
+\end{algorithmic}
+\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}
+\end{figure}
 <example of Bri09>
@@
 
-The blocks should be stored in an order that, depending on the endianness of the system, when reading an unsigned integer from the block array, the blocks are already 
-in correct order. This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
+The data in blocks should be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order. 
+This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
 element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
@@
 Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
-$select_1$ query, which costs a lot timewise. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
-the bit array, fix it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See 
-Table~\ref{endblockcall} for detailed times.
+$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
+the bit array, fix it's offset and count trailing zeros. Traveling the bit array is preferred, since it is very simple to implement and is not noticably slower. See 
+Table~\ref{endblockcall} for detailed times. See Figure~\ref{select_pseudo} for a DAC with $select$ and block size 8. With different block sizes, extra calculation to get 
+the block location from the byte array is needed. In the example, $CalculateLength$ returns the length of the block and $ByteMask(k)$ returns a bit mask for k bytes. $number$
+is a long unsigned integer with a length of multiple bytes.
 
@@
 
+\begin{figure}[ht]
+\begin{algorithmic}
+\State $i \gets $ wanted index
+\State $A \gets $ block byte array
+\State $B \gets $ continuation bit array
+\State $begin \gets \Call{Select}{index}$
+\State $len \gets \Call{CalculateLength}{begin}$
+\State $mask \gets \Call{ByteMask}{len}$
+\State $number \gets A[begin]$ \Comment{read a larger number from the byte array}
+\State $number \gets number \mathbin{\&} mask$
+
+\end{algorithmic}
+\caption{Pseudo code of DAC with select with block length 8} \label{select_pseudo}
+\end{figure}
+
 TODO: check if rank query takes less when done on single byte array instead of multiples
@@
  - get actual results of rank vs select support size
- - take advantage of how data is stored
 
- - emphasis on size!
+ - compare data structure size!
 
-     RANK seems a lot faster, check algorithms!!!
+     RANK seems a lot faster on small arrays, check both algorithms!
 TODO: try bitvector SD performance + size (select)
@@
 
-The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server>.
+The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server and write specs here>.
 
@@
 
+ - do cachegrind
+ - 
  - use array instead of vector<uint8 t> for bris?

### 2020-02-19 vb_encode_8bit_byteoptimized.cpp

@@ -70,13 +70,19 @@ int shiftamt;
 uint64_t maxuint = 0-1;
 cout << maxuint << endl;
 chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
+uint64_t bitmasks[8];
+val = 0;
+for(int i = 0;i < 8; i++) {
+  val = val << 8;
+  val = val | 0xF;
+  bitmasks[i] = val;
+}
 CALLGRIND_START_INSTRUMENTATION;
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
 
 //TODO CALLGRIND THIS
     begin = index == 0 ? 0 : sls(index)+1;
-    test = (uint64_t *)&iv[begin];
 //    while(*(b.begin()+begin+diff)==0) diff++;
 //    bit_vector::iterator iter = b.begin()+begin;
 //    while(*iter == 0) {
@@ -93,7 +99,7 @@ CALLGRIND_START_INSTRUMENTATION;
       uint64_t blokki2 = *(b.data()+block+1);
       val = val | (blokki2 <<(64-offset));
     }
-    diff = bits::lo(val) +1;
+    diff = bits::lo(val); //ADD +1 for shiftamt stuff
 
 //    if ((val & 0x0000000F) == 0) {diff = diff + 4; val = val >> 4;}
 //    if ((val & 0x00000003) == 0) {diff = diff + 2; val = val >> 2;}
@@ -101,9 +107,12 @@ CALLGRIND_START_INSTRUMENTATION;
 //    uint8_t *nro = (uint8_t *)&b[begin];
 //      end = sls(index+1); //TODO this has to be optimized
 //      diff = 1;//end-begin;
-    shiftamt = 64-bit_length*diff;
-    val = *test<<(shiftamt)>>shiftamt;
+//  TODO: try with bit mask!
+//    shiftamt = 64-bit_length*diff;
+//    val = *test<<(shiftamt)>>shiftamt;
+    test = (uint64_t *)&iv[begin];
 
+    val = *test & bitmasks[diff];
     z = z^val;
 
     if(0 && val != original[index]) {

### 2020-02-21 gradu/msc_english_contents.tex
minor adds

@@
 
+\section{Rank and Select}
 Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
-$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
-space over the bit array B. See Table~\ref{supportsize} for $rank$ and $select$ size requirement approximations on different sizes of bit arrays.
+$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?). See Table~\ref{supportsize} for $rank$ and $select$ 
+size requirement approximations over the bit array using \citep{gbmp2014sea}.
 
@@
 
+For every 512th bit in the array, rank stores the absolute value of 1's preceding that bit in a superblock. For every 64th bit in a superblock, a relative 
+count is stored. With this setup, an rank can be instantly calculated for every 64th bit. To get rank(i), superblock[i/512] and it's relative value (i\%512)/64 
+are fetched. Then 1 bits from (i-i\%64) to i are counted and added to get the correct sum. For Rankv5, superblock size is 2048 and relative count is stored every
+384th bit. This offers a decent tradeoff between performance and memory usage.
+
+- explain select here
+<<explain rank and select algorithms here>>
+
 \begin{table}
 \centering
-\caption{Size of support structure with SDSL-lite library}
-\begin{tabular}{l||c c c c c c} 
-Structure & TODO & get & multiple & size & calculations & here\\ 
+\caption{Memory requirements of SDSL rank and select data structures\label{table:supportsize}}
+\begin{tabular}{l||c} 
+Structure & Extra size taken\\ 
 \hline \hline 
-$rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
-$rank_v5$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
-$select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
+Rank   & \text{25\% of bit array} \\
+Rankv5 & \text{6.25\% of bit array}\\
+Select & \text{8.3\% of bit array}\\
+TODO & see if data quality or size matters here\\
+
 \hline
@@
 \end{tabular}
-\label{supportsize}
 \end{table}
@@
 Table~\ref{endblockcall} for detailed times. See Figure~\ref{select_pseudo} for a DAC with $select$ and block size 8. With different block sizes, extra calculation to get 
-the block location from the byte array is needed. In the example, $CalculateLength$ returns the length of the block and $ByteMask(k)$ returns a bit mask for k bytes. $number$
-is a long unsigned integer with a length of multiple bytes.
+the block location from the byte array is needed. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns a bit mask for k bytes.
 
@@
 
-The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server and write specs here>.
+The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server and write specs here>. The algorithms 
+were implemented with C++ (TODO: list compile tags and whatnot here) using data structures and functions from \citep{gbmp2014sea}.
 
-Five different kinds of VB decoding algorithms were used. Bris-implementations are from \citep{Bri09}. Rank and select functions are from
- \citep{gbmp2014sea}.
-
+Five different kinds of VB decoding algorithms were used. Bris-implementations are from \citep{Bri09}. 
 \begin{itemize}

### 2020-02-21 vb_encode_4bit_bris.cpp

@@ -58,9 +58,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  rank_support_v<0> rb[16];
+  rank_support_v5<0> rb[16];
   for(int i =0; i < 16; i++) {
-    rank_support_v<0> r(&b[i]);
+    rank_support_v5<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);

### 2020-02-23 gradu/msc_english_contents.tex
Refactoring of text

@@
 data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not natively
-possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the preceding data. Direct access 
-is achievable with supporting data structures. An obvious solution is to store each element's location, but that adds a very large overhead which removes the 
-benefit of compression, but for some compression techniques it adds a very large overhead and thus is not applicable.
+possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the length of the preceding 
+compressed data.
+
+Direct access is achievable with supporting data structures. An obvious solution is to store the location of each element, but it adds a very large overhead 
+which removes the benefit of compression.
 
@@
 Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
-$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?). See Table~\ref{supportsize} for $rank$ and $select$ 
-size requirement approximations over the bit array using \citep{gbmp2014sea}.
+$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?). Locations of 1 bits in B should represent locations 
+of the items in the encoded data. For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
+has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which 
+significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) 
+would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.
 
-Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, this requires B to be created in addition 
-to the existing data and the length of B usually has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data 
-is compressed in blocks of equal length which significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the 
-locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending 
-byte of i-th compressed element.
+\section{Rank and Select implementation}
+
+The rank and select implementations used in this article are C++ libraries from \citep{gbmp2014sea}. The library has several implementations of both rank and 
+select as well as an implementaion of a bit array. Table~\ref{table:supportsize} has $rank$ and $select$ size requirements of implementations used in this experiment.
+Both rank implementations have a constant space requirement, while select's needed size depends on the number of 1's in the data.
 
@@
 are fetched. Then 1 bits from (i-i\%64) to i are counted and added to get the correct sum. For Rankv5, superblock size is 2048 and relative count is stored every
-384th bit. This offers a decent tradeoff between performance and memory usage.
+384th bit, which offers a tradeoff between performance and memory usage.
 
-- explain select here
-<<explain rank and select algorithms here>>
+Select data structure works similarly to rank. Index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+relative to the superblock.
 
@@
 Rankv5 & \text{6.25\% of bit array}\\
-Select & \text{8.3\% of bit array}\\
-TODO & see if data quality or size matters here\\
-
+Select & \text{~8.3\% of bit array (on average)}\\
 \hline
-%
 \end{tabular}
@@
 Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
-$A_n$ and then use $rank_1$ to get direct access. For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second 
-least significant bits to $A_2$ and the most significant bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. 
-If the continuation bit is set at 0 (meaning there are more chunks to this element), getting the correct index for $A_2$ is calculated with $rank_1$($A_1$, i). 
-Small values benefit from this reordering, because the first chunk does not need any calculation. This method causes a small overhead on the data, because each 
-continuation bit array needs a support data structure for rank. 
-A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. In the actual implementation, different block size requires
-extra calculation to get the block location from the byte array. (TODO: do i need to explain how to get the block location?) (TODO: does this need an example?)
+$A_n$ and then use $rank_0$ to get direct access. If i is the requested index, then $rank_0$($A_k$, i) returns the number of elements preceding i that have 
+continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$.
+
+For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second least significant bits to $A_2$ and the most significant 
+bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. If the continuation bit is set at 0 (meaning there are more chunks 
+to this element), getting the correct index for $A_2$ is calculated with $rank_0$($A_1$, i). A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
+with block length of 8. (TODO: example needed?)
 
@@
 \State $number \gets number \mathbin{|} block$
-
 \end{algorithmic}
@@
 \end{figure}
-<example of Bri09>
-
-- explain how random access is good ?
 
@@
 location of the end byte of i-th element. Therefore the start of j-th element in the block array is at block $b_s$ = $select_1$(j-1) + 1. To simplify calculations even further,
-block sizes of 2, 4 and 8 were used to avoid splitting of blocks between bytes.
+block sizes of 4 and 8 were used to avoid splitting of blocks between bytes.
 
-The data in blocks should be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order. 
-This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
-element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
-is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
-32-b*k to re-align bits.
+Unlike the standard VB encoding, the continuation bits are not in the middle of the data blocks. The compressed number can be read from the memory block and concatenation of blocks
+is not needed. The data in blocks is be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order.
 
-Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
-$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
-the bit array, fix it's offset and count trailing zeros. Traveling the bit array is preferred, since it is very simple to implement and is not noticably slower. See 
-Table~\ref{endblockcall} for detailed times. See Figure~\ref{select_pseudo} for a DAC with $select$ and block size 8. With different block sizes, extra calculation to get 
-the block location from the byte array is needed. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns a bit mask for k bytes.
+If the original element size is 32 bits and the data was compressed to k blocks of length b. The starting byte s of the element is $select$(i) * b div 8 where div is the integer division. 
+Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit value is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
+bits and bit shift right 32-b*k to re-align bits. If block size is one byte, offset calculation is not needed and thus the process is slightly faster.
 
-\begin{table}
-\centering
-\caption{Time to calculate end blocks}
-\begin{tabular}{l||c c c c c c} 
-Version & Average time over 10Mall \\ 
-\hline \hline 
-select(i+1) & 34.97 \\
-iterate bit array & 34.97 \\
-bit shift magic & 34.97 \\
-\hline
-%
-\end{tabular}
-\label{endblockcall}
-\end{table}
+The intuitive way to calculate block length k of i-th element is from $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
+$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until the next 1. Alternatively, the block length can be calculated by reading an integer 
+from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB encoding with DAC with $select$ and block size 8. 
+Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
+a bit mask for k bytes.
 
@@
 
-TODO: check if rank query takes less when done on single byte array instead of multiples
-
- - get actual results of rank vs select support size
-
- - compare data structure size!
-
-     RANK seems a lot faster on small arrays, check both algorithms!
 TODO: try bitvector SD performance + size (select)
@@
 
-\chapter{Conclusion}
- - here
-
 \chapter{Future work}

### 2020-02-26 gradu/msc_english_contents.tex
mods

@@
 One of the leading methods of data compression is variable-length coding \citep{Sal99}, where frequent sequences of data
-are represented with shorter codewords. Because the sequences of data have different lenghts when compressed, it is 
-not trivial to determine the exact location of a certain element. If this is required, the usual data compression algorithms are
-inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill.
+are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
+not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
+data compression algorithms are inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill. However, 
+random access of compressed data is useful in compressed data structures. It saves storage space, bandwith and increases the likelihood
+of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data from the beginning.
 
-However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
-to this is to decompress data from the beginning. An integer compressing method with fast random access is explained later and compared
-existing state-of-the-art methods.
+A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used clever block 
+reorganizing and $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
+has been widely adopted.
+
+An alternative solution with $select$ data structure is shown later. (TODO: refer to results)
+
+--contributions?
 
@@
 
-Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
-compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
-and sometimes preferred due to it's simplicity. allows constant time random access. This is later explained in (TODO: link to later chapter).
+Variable-byte (VB) encoding \citep{Wil99} is a method for compressing integers via omitting leading zero bits that would be present in a longer fixed 
+width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
+is generally faster to decode \citep{Bri09} and sometimes preferred due to its simplicity. As later shown in Chapter~\ref{chapter:DAC}, constant time 
+random access is possible with VB encoding.
 
@@
 words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
-advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number instead of the actual 
-number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
-have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.
+advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number 
+instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. 
+In contrast, rare words have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.
+
+Variable-byte encoding originates from and is used in MIDI music file \citep{Mid96} and several applications have a similar implementation of VB. Apache 
+Lucene has vInt datatype. Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
+ Microsoft .Net framework offers "7BitEncodedInt" in BinaryReader and BinaryWriter classes and IBM DB2 has a variable byte  \citep{Bha09}.
+
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small numbers, 
+which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
+of their changing code length.
 
@@
 are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
-
-\medskip
-\noindent\begin{minipage}{.5\textwidth}
+\begin{figure}[ht]
 \begin{algorithmic}
@@
 \While{true}
-\State $bytes$.prepend($n \bmod 128$)
+\State \Call{prepend}{bytes,$n \bmod 128$}
 \If{$n < 128$} \State break \EndIf
@@
 \EndWhile
-\State $bytes$.last() += $128$
+\State $bytes$[\Call{len}{$bytes$}-1] += $128$
 \State \textbf{return} bytes
 \EndFunction
-\captionof{figure}{VByte encoding} \label{vbyte_enc}
-\end{algorithmic}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-\begin{algorithmic}
+
 \Function{VBEncode}{$numbers$}
@@
 \EndFunction
+
 \end{algorithmic}
-\end{minipage}
-\medskip
+\captionof{figure}{VByte encoding} \label{vbyte_enc}
+\end{figure}
 
@@
 
-\chapter{Directly addressable codes?}
+\chapter{Directly addressable codes} \label{chapter:DAC}
 
@@
 
+ - Run with 32 bit values, see what happens (some operations may be 32b?)
  - do cachegrind
@@
 
+ - $rank$ vs rank? should formulas have a specific style?
+
 \chapter{Future work}

### 2020-02-28 gradu/msc_english_contents.tex
Moved algorithm
centered 2.1
Fixes

@@
 are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
+
 \begin{figure}[ht]
-\begin{algorithmic}
+\centering
+  \begin{minipage}{0.5\linewidth}
+\begin{algorithmic}[H]
 \Function{VBEncodeNumber}{$n$}
@@
 \EndFunction
-
+\medskip
+\medskip
 \Function{VBEncode}{$numbers$}
@@
 \ForEach {$n \in numbers$}
-\State $bytes \gets$ VBEncodeNumber($n$)
-\State $bytestream$.extend($bytes$)
+\State $bytes \gets$ \Call{VBEncodeNumber}{$n$}
+\State \Call{extend}{$bytestream$,$bytes$}
 \EndFor
@@
 \end{algorithmic}
+\end{minipage}
 \captionof{figure}{VByte encoding} \label{vbyte_enc}
@@
 are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
-is \texttt{00010 11000} and $n = 0$. Block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to left 
+is \texttt{00010 11000} and $n = 0$. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to the left 
 equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
-$n = 42$. Because the previous continuous bit was 1, decoding this number has finished. An example implementation of encode and decode with block 
-length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.
+$n = 42$. Because the previous continuation bit was 1, decoding for this number is complete. More examples shown in Table~\ref{table:vbytes}. 
+An example implementation of encode and decode with block length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.
 
 \begin{figure}[ht]
-\begin{algorithmic}
+\centering
+  \begin{minipage}{0.5\linewidth}
+\begin{algorithmic}[H]
 \Function{VBDecode}{$bytestream$}
@@
 \end{algorithmic}
+\end{minipage}
 \caption{VByte decoding} \label{vbyte_dec}
@@
 Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
-offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
+offer less effective compression. Generally block length of 7 has been used because it tends to perform well on average and handling chunks as bytes is 
 convenient \citep{Man08}.
@@
 VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
-in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
-(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first experimented in 
-compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different approaches 
-have been introduced.
+in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
+(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was was to be found efficient in compressing integer lists. It was first experimented 
+as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different 
+approaches have been introduced.
 
-Modern studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) architecture to VB \citep{Lem18,Pla15}. The bit 
+More recent studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) instructions to VB \citep{Lem18,Pla15}. The bit 
 operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
 
-TODO: might need 1-2 more chapters
+\begin{table}
+\centering
+\begin{tabular}{l||c c c c c} 
+Original number & first block & second block & third block & fourth block &\\ 
+\hline \hline 
+4  & \texttt{\underline{1}4000}    &                             &                           &  &  \\
+17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
+620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
+60201 & \texttt{\underline{1}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
+
+\hline
+\end{tabular}
+\caption{VByte encoded numbers, block size 4. Continuation bit underlined.\label{table:vbytes}}
+\end{table}
 
@@
 
-Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
-data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not natively
-possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the length of the preceding 
-compressed data.
-
-Direct access is achievable with supporting data structures. An obvious solution is to store the location of each element, but it adds a very large overhead 
-which removes the benefit of compression.
+The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
+data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
+strutures. It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
+list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
+of each element, but it adds a very large overhead which removes the benefit of compression.
 
@@
 Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
-$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?). Locations of 1 bits in B should represent locations 
+$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep{gbmp2014sea}. Locations of 1 bits in B should represent locations 
 of the items in the encoded data. For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
@@
 $A_n$ and then use $rank_0$ to get direct access. If i is the requested index, then $rank_0$($A_k$, i) returns the number of elements preceding i that have 
-continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$.
+continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$. See Figure~\ref{bris_ds} for visualization of data.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.4]{bris.png}
+\caption{Data structure by Brisaboa et al., visualized} \label{bris_ds}
+\end{figure}
 
@@
 
-To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. $Select$(i) returns the
-location of the end byte of i-th element. Therefore the start of j-th element in the block array is at block $b_s$ = $select_1$(j-1) + 1. To simplify calculations even further,
-block sizes of 4 and 8 were used to avoid splitting of blocks between bytes.
+To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
+only on it's last block, $Select_1$(i) returns the location of the end byte of i-th element. Therefore the start of j-th element in the block array is at 
+block $b_s$ = $select_1$(j-1) + 1. Implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.
 
-Unlike the standard VB encoding, the continuation bits are not in the middle of the data blocks. The compressed number can be read from the memory block and concatenation of blocks
-is not needed. The data in blocks is be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order.
+Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own array, which leaves the blocks in their own array. This allows the 
+compressed number to be read from the memory block and removes the need of block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
+when reading a word from the block byte array, the bits and bytes are already in correct order.
 
 If the original element size is 32 bits and the data was compressed to k blocks of length b. The starting byte s of the element is $select$(i) * b div 8 where div is the integer division. 
-Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit value is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
+Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit word is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
 bits and bit shift right 32-b*k to re-align bits. If block size is one byte, offset calculation is not needed and thus the process is slightly faster.
 
-The intuitive way to calculate block length k of i-th element is from $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
+The most intuitive way to calculate block length of i-th element is from $select_1$(i+1) and subtract the previously calculated start block index. This however causes a second 
 $select_1$ query, which is costly. A much faster option is to iterate forward in bit array until the next 1. Alternatively, the block length can be calculated by reading an integer 
-from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB encoding with DAC with $select$ and block size 8. 
+from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
 Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
@@
 
-(maybe to future work) Storing bytes in this fashion may have benefits when reading subsequent (fast next() and previous()?) elements.
-With some modifications, SIMD instructions can possibly be applied.
-
 \begin{figure}[ht]
+\centering
 \begin{algorithmic}
@@
 \State $mask \gets \Call{ByteMask}{len}$
-\State $number \gets A[begin]$ \Comment{read a larger number from the byte array}
-\State $number \gets number \mathbin{\&} mask$
+\State $word \gets A[begin]$ \Comment{read a word from the byte array}
+\State $word \gets word \mathbin{\&} mask$
 
@@
 
+Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. Length of the compressed 
+element is obtainable e.g. via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$. Because element length in this case is 3, blocks not belonging to the element
+are removed either with a bitmask or by double bit shifting, resulting in decoded k-th element.
+
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+[node distance=0pt,
+ start chain = A going right,
+ start chain = B going right,
+ arrow/.style = {draw=#1,-{Stealth[]}, 
+                shorten >=1mm, shorten <=1mm}, % styles of arrows
+ arrow/.default = black,
+ X/.style = {rectangle, draw,% styles of nodes in string (chain)
+                minimum width=9ex, minimum height=5ex,
+                outer sep=0pt, on chain}
+ ]
+
+\node[X, join] (start) {...};
+\node[X] (first){0};
+\node[X, very thick] (select){1};
+\node[X] (Btarget){0};
+\node[X, join] (second){0};
+\node[X, very thick, join] (end){1};
+\node[X, join] (third){0};
+\node[X, join] {$B_{...}$};
+
+\node [above=1ex of start]{Bit array};
+\node [below=1ex of start]{index:};
+\node [above=1ex of select]{k-1th 1};
+\node [below=1ex of first]{$i-1$};
+\node [below=1ex of select]{$i$};
+\node [below=1ex of Btarget]{$i+1$};
+\node [below=1ex of second]{$i+2$};
+\node [below=1ex of end]{$i+3$};
+\node [below=1ex of third]{$i+4$};
+\node [above=1ex of end]{kth 1};
+\node[X, below=12ex of start] (x) {...};
+\node[X, join] {$A_{i-2}$};
+\node[X, join] {$A_{i-1}$};
+\node[X, ultra thick] (Atarget){$A_{i}$};
+\node[X, ultra thick] {$A_{i+1}$};
+\node[X, ultra thick] {$A_{i+2}$};
+\node[X, join] {$A_{i+3}$};
+\node[X, join] {$A_{...}$};
+\node [above=1ex of x]{Block array};
+
+\draw[arrow] (Btarget.south) to (Atarget.north);
+\end{tikzpicture}
+\caption{Data} \label{VBarrays}
+\end{figure}
 TODO: try bitvector SD performance + size (select)
-
-TODO: look into select implementation, may need optimizing
 

### 2020-02-28 gradu/msc_english_contents.tex
Changes

@@
 
-A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used clever block 
+A variable-byte encoding based integer compression method with constant time random access was first introduced by \citep{Bri09}. They used clever block 
 reorganizing and $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
@@
 
-An alternative solution with $select$ data structure is shown later. (TODO: refer to results)
+An alternative solution with $select$ data structure is explained in detail and compared to $rank$ by Brisaboa et al. This method does not use data 
+reorganizing, but rather capitalizes on the assumption that data is encoded the same way it's in the source. Because variable-byte encoding does not assign 
+new codewords to data when encoding, reading the data straight from the memory is much faster than reassembling the word from blocks.
 
---contributions?
+(TODO: refer to results)
 
@@
 \includegraphics[scale=0.4]{bris.png}
-\caption{Data structure by Brisaboa et al., visualized} \label{bris_ds}
+\caption{Data structure by Brisaboa et al., visualized} (TODO: create this with tikz)\label{bris_ds}
 \end{figure}
@@
 to this element), getting the correct index for $A_2$ is calculated with $rank_0$($A_1$, i). A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
-with block length of 8. (TODO: example needed?)
+with block length of 8.
 
@@
 \end{figure}
-TODO: try bitvector SD performance + size (select)
+TODO: try bitvector SD performance + size (select) (maybe write about different select implementations)
 

### 2020-02-28 gradu/msc_english_contents.tex
Fixes

@@
 
-An alternative solution with $select$ data structure is explained in detail and compared to $rank$ by Brisaboa et al. This method does not use data 
-reorganizing, but rather capitalizes on the assumption that data is encoded the same way it's in the source. Because variable-byte encoding does not assign 
+An alternative solution with $select$ data structure is proposed and explained in detail later. Comparison to $rank$ by Brisaboa et al is shown with different
+implementations of $rank$ and $select$ and several different data sets are used. Proposed method does not use data 
+reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
 new codewords to data when encoding, reading the data straight from the memory is much faster than reassembling the word from blocks.
@@
 
-Select data structure works similarly to rank. Index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
-relative to the superblock.
+$Select$ data structure works similarly to rank. Index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+relative to the superblock. $Select$
 

### 2020-02-28 gradu/msc_english_contents.tex
Fixes

@@
 
-An alternative solution with $select$ data structure is proposed and explained in detail later. Comparison to $rank$ by Brisaboa et al is shown with different
+An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ by Brisaboa et al is shown with different
 implementations of $rank$ and $select$ and several different data sets are used. Proposed method does not use data 
@@
 
-The rank and select implementations used in this article are C++ libraries from \citep{gbmp2014sea}. The library has several implementations of both rank and 
+The rank and select implementations used in this article are from C++ library by \citep{gbmp2014sea}. The library has several implementations of both rank and 
 select as well as an implementaion of a bit array. Table~\ref{table:supportsize} has $rank$ and $select$ size requirements of implementations used in this experiment.
-Both rank implementations have a constant space requirement, while select's needed size depends on the number of 1's in the data.
+Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on the number of 1's in the data.
 
@@
 $Select$ data structure works similarly to rank. Index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
-relative to the superblock. $Select$
+relative to the superblock.
 

### 2020-03-01 gradu/msc_english_contents.tex
Fixes

@@
 Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
-$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep{gbmp2014sea}. Locations of 1 bits in B should represent locations 
-of the items in the encoded data. For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
-has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which 
-significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) 
-would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.
+$select_1$(B, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep{gbmp2014sea}. To use rank or select in indexing, the
+set 1 bits in B should reflect to the element locations in the encoded data.
+
+For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
+has to be close to the length of the data, because encoded element lenghts vary. VB encoding has several advantages with search and rank: the data is compressed in 
+blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
+as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks up to i-th index and $select_1$(C, i) would give the location of the ending block 
+of i-th compressed element.
 
@@
 
-For every 512th bit in the array, rank stores the absolute value of 1's preceding that bit in a superblock. For every 64th bit in a superblock, a relative 
-count is stored. With this setup, an rank can be instantly calculated for every 64th bit. To get rank(i), superblock[i/512] and it's relative value (i\%512)/64 
-are fetched. Then 1 bits from (i-i\%64) to i are counted and added to get the correct sum. For Rankv5, superblock size is 2048 and relative count is stored every
-384th bit, which offers a tradeoff between performance and memory usage.
+The data structure of rank has two layers. First layer is the superblock array. For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
+The second layer is the relative count block. It stores a relative count of 1's for every 64th bit. To calculate rank(i), the superblock index is s = i/512 and 
+relative count block index is r = i-s / 64, both being integer divisions. Then number of 1 bits from index r to i are calculated. All three values added together add up to rank(i). 
+Rankv5 is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, but reduces 
+memory requirement to a quarter.
 
-$Select$ data structure works similarly to rank. Index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
-relative to the superblock.
+$Select$ data structure works similarly to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+relative to the superblock. With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until required amount of bits is reached.
+
+In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only differiating factor is the manual bit count from relative
+block's index to wanted index. This is at most the size of one relative block and thus both of the functions work in constant time \citep{Gon05}.
 
@@
 \section{DAC via rank}
-Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
-$A_n$ and then use $rank_0$ to get direct access. If i is the requested index, then $rank_0$($A_k$, i) returns the number of elements preceding i that have 
-continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$. See Figure~\ref{bris_ds} for visualization of data.
+Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. In their solution, the chunks are sorted in separate arrays by their significance. For each integer, the block 
+of the least significant chunk is stored in the first array $A_1$, and its bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
+stored to $A_2$ and the bit to $B_2$ and so on. After the data is set, rank data structure is built for each bit array. Figure~\ref{bris_ds} contains a visualization of the data structure.
 
@@
 
-For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second least significant bits to $A_2$ and the most significant 
-bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. If the continuation bit is set at 0 (meaning there are more chunks 
-to this element), getting the correct index for $A_2$ is calculated with $rank_0$($A_1$, i). A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
+To decode an element from index i, first block is fetched from $A_1$[i]. Because each is composed of at least one block, the first block is obtainable via straight indexing. With small numbers, this 
+is all that is required.
+If the data has a lot of small numbers, the $rank$ method has a huge advantage.
+However if the element was stored in multiple blocks, $i \gets rank_0$($A_1$, i) is called.
+$Rank_0$
+returns the number of zeros preceding i. In other words, the number of blocks that continue to the next array. This in turn means that $i$ now has the index of the element's next block in array $A_2$.
+This process is then repeated until ith bit of the current bit array is set.
+The decompressed integer is constructed from the fetched blocks.
+A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo}
 with block length of 8.
@@
 \begin{figure}[ht]
+\centering
+\begin{minipage}{0.5\linewidth}
 \begin{algorithmic}
@@
 \end{algorithmic}
+\end{minipage}
 \caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}
+
 \end{figure}
+
+--TODO: create another version: 7bit bris.
+Store bits in both bit array and chunk (to see if it's faster to get the continuation bit from the chunk than from the bit array) (also needs 7bit compliant dataset)
 
@@
 
-The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server and write specs here>. The algorithms 
-were implemented with C++ (TODO: list compile tags and whatnot here) using data structures and functions from \citep{gbmp2014sea}.
+The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64Kb+64Kb L1 cache, 512Kb L2 cache, 6144Kb L3 cache and 32GB of DDR3 1333MHz.
+The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The code was compiled with g++-7 -std=c++11 -DNDEBUG and with libraries -lsdsl -ldivsufsort -ldivsufsort64.
 
-Five different kinds of VB decoding algorithms were used. Bris-implementations are from \citep{Bri09}. 
+Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
+library \citep{gbmp2014sea}. Each implementation is decoding a uint64\_t integer.
+
 \begin{itemize}
@@
   \item Bris8v5 - Bris implementation with block size 8, using rank support v5
-  \item Our4 - Our implementation with block size 4, using select support mcl
-  \item Our8 - Our implementation with block size 8, using select support mcl
+  \item Select4 - Proposed implementation with block size 4, using select support mcl
+  \item Select8 - Proposed implementation with block size 8, using select support mcl
 \end{itemize}
 
-Eight kinds of datasets were used. For each number in the dataset, $p$ was randomly selected from $P$ and then number was randomly selected from
-range [0,$2^p$].
+Five differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).
 
 \begin{itemize}
-  \item all5M - 5 million numbers, $P$ = {7,8,15,16,23,24,30,31}
-  \item all50M - 50 million numbers, $P$ = {7,8,15,16,23,24,30,31}
-  \item byte5M - 5 million numbers, $P$ = {7,7,7,8,8,8,16,31}
-  \item byte50M - 50 million numbers, $P$ = {7,7,7,8,8,8,16,31}
-  \item small5M - 5 million numbers, $P$ = {3,4,5,6,7,8,16,31}
-  \item small50M - 50 million numbers, $P$ = {3,4,5,6,7,8,16,31}
-  \item vsmall5M - 5 million numbers, $P$ = {2,2,3,3,3,4,4,15}
-  \item vsmall50M - 50 million numbers, $P$ = {2,2,3,3,3,4,4,15}
+  \item all - integers equally 1-4 bytes long
+  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
+  \item twolarge2 - Same as twolarge, but third of the 1 byte numbers 4 bits or less
+  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
+  \item onebyte - all integers 4 bits or less
 \end{itemize}
 
-All algorithms save the data to memory and randomize one million index numbers to an array. The time taken is calculated from looping through the 
-index array and VB decoding the number in the index. The times shown in Table~\ref{table:results1} are an average of 100 such runs.
+All implementation read the data set from a file and compres it to memory. Then one million index numbers are randomized to an array. The times shown in Table~\ref{table:results1} 
+are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.
 
@@
 \caption{Results in milliseconds, smaller is better.\label{table:results1}}
-\begin{tabular}{l||c c c c c c c c} 
-Experiment & all5M & all50M & byte5M & byte50M & small5M & small50M & vsmall5M & vsmall50M\\ 
+\begin{tabular}{l||c c c c c c} 
+ & Bris4v5 & Select4 & Bris8 & Bris8v5 & Select8 \\ 
 \hline \hline 
-Bris4v5 & 348.41 & 768.83 & 345.16 & 772.84 & 344.54 & 768.45 & 346.86 & 771.61 \\
-Our4    & 127.92 & 263.22 & 127.44 & 263.5  & 127.9  & 262.67 & 127.88 & 262.58 \\
-Bris8   & \textbf{97.75}  & 308.17 & \textbf{98.47}  & 309.6  & \textbf{98.14}  & 307.81 & \textbf{97.99}  & 308.09 \\
-Bris8v5 & 127.83 & 287.46 & 127.7  & 288.63 & 128.14 & 291.32 & 127.99 & 287.84 \\
-Our8    & 103.85 & \textbf{230.16} & 103.92 & \textbf{230.14} & 103.86 & \textbf{230.02} & 103.95 & \textbf{230.6} \\
+all (5M)        & - & 189.8 & 194.0 & 230.3 & 139.2 \\
+all (50M)       & - & 292.0 & 256.0 & 304.4 & 232.7 \\
+all (500M)      & - &  -    & 299.9 & 369.2 & 335.1 \\
+twolarge (5M)   & - & 164.1 & 93.9  & 216.7 & 117.5 \\
+twolarge (50M)  & - & 270.4 & 140.0 & 159.8 & 210.6 \\
+twolarge (500M) & - & 381.4 & 169.7 & 383.6 & 307.8 \\
 
@@
  - Run with 32 bit values, see what happens (some operations may be 32b?)
- - do cachegrind
  - 
@@
  - do we need to support search()?
- - SIMD?
 

### 2020-03-01 gradu/msc_english_contents.tex
Fixes

@@
 reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
-new codewords to data when encoding, reading the data straight from the memory is much faster than reassembling the word from blocks.
+new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.
 
@@
 
-A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
+A good data set for VB encoding is a list of mostly small integers with a need to support larger ones. A search engine may use an inverted index of 
 words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
 advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number 
-instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. 
-In contrast, rare words have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.
+instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the integers are small. 
+In contrast, rare words have only a few entries but the integers stored are larger. These lists are excellent data sets for VB encoding.
 
@@
 
-Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small numbers, 
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small integers, 
 which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
@@
 
-VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
+VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the integers
 in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
@@
 \end{tabular}
-\caption{VByte encoded numbers, block size 4. Continuation bit underlined.\label{table:vbytes}}
+\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
 \end{table}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
 \centering
-\includegraphics[scale=0.4]{bris.png}
-\caption{Data structure by Brisaboa et al., visualized} (TODO: create this with tikz)\label{bris_ds}
+\begin{tikzpicture}
+[node distance=0pt,
+ start chain = A going right,
+ arrow/.style = {draw=#1,-{Stealth[]}, 
+                shorten >=1mm, shorten <=1mm}, % styles of arrows
+ arrow/.default = black,
+ X/.style = {rectangle, draw,% styles of nodes in string (chain)
+                minimum width=7ex, minimum height=4ex,
+                outer sep=0pt, on chain}
+ ]
+\node[X, join] (start) {$C_{1,2}C_{1,1}$};
+\node [left=1ex of start]{\textbf{C =}};
+\node[X] {$C_{2,1}$};
+\node[X] {$C_{3,3}C_{3,2}C_{3,1}$};
+\node[X] {$C_{4,2}C_{4,1}$};
+\node[X] {$C_{5,1}$...};
+
+\node[X, below=5ex of start] (a1) {$A_{1,1}$};
+\node [left=1ex of a1]{\textbf{$A_1$ =}};
+\node[X] {$A_{2,1}$};
+\node[X] {$A_{3,1}$};
+\node[X] {$A_{4,1}$};
+\node[X] {$A_{5,1}$};
+\node[X] {...};
+
+\node[X, below=0ex of a1] (b1) {1};
+\node [left=1ex of b1]{\textbf{$B_1$ =}};
+\node[X] {0};
+\node[X] {1};
+\node[X] {1};
+\node[X] {0};
+\node[X] {...};
+
+\node[X, below=5ex of b1] (a2) {$A_{1,2}$};
+\node [left=1ex of a2]{\textbf{$A_1$ =}};
+\node[X] {$A_{3,2}$};
+\node[X] {$A_{4,2}$};
+\node[X] {...};
+
+\node[X, below=0ex of a2] (b2) {0};
+\node [left=1ex of b2]{\textbf{$B_2$ =}};
+\node[X] {1};
+\node[X] {0};
+\node[X] {...};
+
+\node[X, below=5ex of b2] (a3) {$A_{3,3}$};
+\node [left=1ex of a3]{\textbf{$A_3$ =}};
+\node[X] {...};
+
+\node[X, below=0ex of a3] (b3) {0};
+\node [left=1ex of b3]{\textbf{$B_3$ =}};
+\node[X] {...};
+
+\end{tikzpicture}
+\caption{Data structure by Brisaboa et al., visualized}\label{bris_ds}
+
 \end{figure}
 
-To decode an element from index i, first block is fetched from $A_1$[i]. Because each is composed of at least one block, the first block is obtainable via straight indexing. With small numbers, this 
+To decode an element from index i, first block is fetched from $A_1$[i].
+Because each encoded element is composed of at least one block, the first block is obtainable via straight indexing.
+With small integers, this
 is all that is required.
-If the data has a lot of small numbers, the $rank$ method has a huge advantage.
+If the data has a lot of small integers, the $rank$ method has a huge advantage.
 However if the element was stored in multiple blocks, $i \gets rank_0$($A_1$, i) is called.
 $Rank_0$
-returns the number of zeros preceding i. In other words, the number of blocks that continue to the next array. This in turn means that $i$ now has the index of the element's next block in array $A_2$.
+returns the number of zeros in the bit array preceding index i.
+In other words, the number of blocks that continue to the next array.
+This in turn means that $i$ now has the index of the element's next block in array $A_2$.
 This process is then repeated until ith bit of the current bit array is set.
@@
 \State $number \gets number \mathbin{\ll} 8$
-\State $number \gets number \mathbin{|} block$
+\State $number \gets number + block$
 \end{algorithmic}
@@
   \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
-  \item twolarge2 - Same as twolarge, but third of the 1 byte numbers 4 bits or less
+  \item twolarge2 - Same as twolarge, but third of the 1 byte integers 4 bits or less
   \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
@@
 
-All implementation read the data set from a file and compres it to memory. Then one million index numbers are randomized to an array. The times shown in Table~\ref{table:results1} 
+All implementation read the data set from a file and compres it to memory. Then one million index integers are randomized to an array. The times shown in Table~\ref{table:results1} 
 are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.

### 2020-03-01 gradu/msc_english_contents.tex
Changes

@@
 \end{figure}
-TODO: try bitvector SD performance + size (select) (maybe write about different select implementations)
 
@@
 
-All implementation read the data set from a file and compres it to memory. Then one million index integers are randomized to an array. The times shown in Table~\ref{table:results1} 
+All implementations read the data set from a file and compress it to memory. Then one million index integers are randomized and stored to an array. The times shown in Table~\ref{table:results1} 
 are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.

### 2020-03-02 gradu/msc_english_contents.tex
Changes

@@
 data compression algorithms are inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill. However, 
-random access of compressed data is useful in compressed data structures. It saves storage space, bandwith and increases the likelihood
+random access of compressed data is useful in compressed data structures. It saves storage space, bandwidth and increases the likelihood
 of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data from the beginning.
@@
 
-An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ by Brisaboa et al is shown with different
-implementations of $rank$ and $select$ and several different data sets are used. Proposed method does not use data 
+An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ implementation by Brisaboa et al is 
+shown with different implementations of $rank$ and $select$ and several different data sets. Proposed method does not use data 
 reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
@@
 
-(TODO: refer to results)
+$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better.
 
-\chapter{Variable-byte encoding}
+\chapter{Variable-bytes}
 
@@
 width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
-is generally faster to decode \citep{Bri09} and sometimes preferred due to its simplicity. As later shown in Chapter~\ref{chapter:DAC}, constant time 
-random access is possible with VB encoding.
+is generally faster to decode \citep{Bri09} and sometimes preferred due to its simplicity.
 
@@
 
-Variable-byte encoding originates from and is used in MIDI music file \citep{Mid96} and several applications have a similar implementation of VB. Apache 
+Variable-byte encoding originates from and is used in MIDI music file format \citep{Mid96} and several applications have a similar implementation of VB. Apache 
 Lucene has vInt datatype. Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
@@
 
-Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small integers, 
-which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
-of their changing code length.
+Later, VB encoding was to 
+be found efficient in compressing integer lists. It was first experimented as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. 
+That yielded excellent records, and since then many different approaches have been introduced.
 
-VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
-on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
-integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
-\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
-are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
+More recent studies have taken a look into the machine code level for VB and applied SIMD (Single instruction, multiple data) instructions to VB \citep{Lem18,Pla15}. The bit 
+operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
+
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. Their encoding process assigns short bit array codes to small integers, 
+which is why they outperform VB encoding on datasets with a lot of small integers \Citep{Wil99}.
+
+\section{VB encoding}
+VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The continuation bit is set to 1 only
+on the block containing the least significant bits of the integer. This information is used in decoding to signal if the current integer continues in the next chunk.
+
+For example, if block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
+\texttt{00000000 00101000}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1000}. The block with least significant bits is given a continuation bit 1.
+The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11000}, 
+which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}.
 
@@
 
-Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
-are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
-is \texttt{00010 11000} and $n = 0$. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to the left 
-equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
-$n = 42$. Because the previous continuation bit was 1, decoding for this number is complete. More examples shown in Table~\ref{table:vbytes}. 
-An example implementation of encode and decode with block length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.
+Smaller block length can yield a better compression rate at the cost of more bit manipulation, while bigger block lenghts need less bit manipulation and 
+offer less effective compression. On the other hand, bigger block length means a smaller percentage of added continuation bits.
+Generally block length of 7 has been used because it tends to perform well on average and handling chunks as bytes is 
+convenient \citep{Man08}.
+
+\begin{table}
+\centering
+\begin{tabular}{l||c c c c c} 
+Original number & first block & second block & third block & fourth block &\\ 
+\hline \hline 
+4  & \texttt{\underline{1}0100}    &                             &                           &  &  \\
+17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
+620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
+60201 & \texttt{\underline{1}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
+
+\hline
+\end{tabular}
+\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
+\end{table}
+
+\section{VB decoding}
+VB decoding is technically just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
+are removed from all the chunks and the blocks are concatenated to form the original number. An example implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.
+
+As in the encoding example, block length is 4, the encoded message 
+is \texttt{00010 11000} and n is a 16 bit unsigned integer. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
+The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
+changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
+$n = $\texttt{101000} (bit representation of 42). Because the continuation bit of this chunk is 1, decoding for this number is complete.
 
@@
 
-Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
-offer less effective compression. Generally block length of 7 has been used because it tends to perform well on average and handling chunks as bytes is 
-convenient \citep{Man08}.
-
-VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the integers
-in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
-(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was was to be found efficient in compressing integer lists. It was first experimented 
-as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different 
-approaches have been introduced.
-
-More recent studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) instructions to VB \citep{Lem18,Pla15}. The bit 
-operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
-
-\begin{table}
-\centering
-\begin{tabular}{l||c c c c c} 
-Original number & first block & second block & third block & fourth block &\\ 
-\hline \hline 
-4  & \texttt{\underline{1}4000}    &                             &                           &  &  \\
-17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
-620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
-60201 & \texttt{\underline{1}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
-
-\hline
-\end{tabular}
-\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
-\end{table}
-
 \chapter{Directly addressable codes} \label{chapter:DAC}
@@
 
-The rank and select implementations used in this article are from C++ library by \citep{gbmp2014sea}. The library has several implementations of both rank and 
+The rank and select implementations used in this thesis are from C++ library by \citep{gbmp2014sea}. The library has several implementations of both rank and 
 select as well as an implementaion of a bit array. Table~\ref{table:supportsize} has $rank$ and $select$ size requirements of implementations used in this experiment.
@@
 \section{DAC via rank}
-Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. In their solution, the chunks are sorted in separate arrays by their significance. For each integer, the block 
+Directly addressable codes in VB was first introduced by \citep{Bri09}. In their solution, the chunks are sorted in separate arrays by their significance. For each integer, the block 
 of the least significant chunk is stored in the first array $A_1$, and its bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
-stored to $A_2$ and the bit to $B_2$ and so on. After the data is set, rank data structure is built for each bit array. Figure~\ref{bris_ds} contains a visualization of the data structure.
+stored to $A_2$ and the bit to $B_2$ and so on. After the data is set, rank data structure is built for each bit array. The data structure is implicit and does not require additional space apart
+from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.
 
@@
 \node[X, join] (start) {$C_{1,2}C_{1,1}$};
+\node [above=1ex of start]{The chunk array. $C_{i,j}$ = $B_{i,j}$ : $A_{i,j}$};
 \node [left=1ex of start]{\textbf{C =}};
@@
 
-\node[X, below=0ex of a1] (b1) {1};
+\node[X, below=0ex of a1] (b1) {0};
 \node [left=1ex of b1]{\textbf{$B_1$ =}};
+\node[X] {1};
+\node[X] {0};
 \node[X] {0};
 \node[X] {1};
-\node[X] {1};
-\node[X] {0};
 \node[X] {...};
@@
 
-\node[X, below=0ex of a2] (b2) {0};
+\node[X, below=0ex of a2] (b2) {1};
 \node [left=1ex of b2]{\textbf{$B_2$ =}};
+\node[X] {0};
 \node[X] {1};
-\node[X] {0};
 \node[X] {...};
@@
 
-\node[X, below=0ex of a3] (b3) {0};
+\node[X, below=0ex of a3] (b3) {1};
 \node [left=1ex of b3]{\textbf{$B_3$ =}};
@@
 With small integers, this
-is all that is required.
-If the data has a lot of small integers, the $rank$ method has a huge advantage.
-However if the element was stored in multiple blocks, $i \gets rank_0$($A_1$, i) is called.
-$Rank_0$
-returns the number of zeros in the bit array preceding index i.
-In other words, the number of blocks that continue to the next array.
+is all that is required. If the data has a lot of small integers, the $rank$ method has a huge advantage.
+
+However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, i). $Rank_0$ 
+returns the number of zeros in the bit array preceding index i. In other words, the number of elements before $i$ that continue to the next array. 
 This in turn means that $i$ now has the index of the element's next block in array $A_2$.
-This process is then repeated until ith bit of the current bit array is set.
-The decompressed integer is constructed from the fetched blocks.
+This process is repeated until the i-th bit of the bit array of current level is set.
+The resulting integer is constructed from the blocks of the fetched chunks.
 A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo}
@@
 \begin{algorithmic}
-\State $i \gets $ wanted index
+\State $index \gets $ wanted index
 \State $A \gets $ block arrays
@@
 \State $number \gets number \mathbin{\ll} 8$
-\State $number \gets number \mathbin{|} block$
+\State $number \gets number + block$
 \State $index \gets \Call{Rank}{B[level], index}$
@@
 
-The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64Kb+64Kb L1 cache, 512Kb L2 cache, 6144Kb L3 cache and 32GB of DDR3 1333MHz.
+The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
 The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The code was compiled with g++-7 -std=c++11 -DNDEBUG and with libraries -lsdsl -ldivsufsort -ldivsufsort64.
@@
 \begin{itemize}
-  \item all - integers equally 1-4 bytes long
+  \item all - all integers randomly 1-4 bytes long
   \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
@@
  -
- - use array instead of vector<uint8 t> for bris?
-- comparison to basic implementation + Bri09
-  - compare with b=2,4,8
 

### 2020-03-02 gradu/msc_english_contents.tex
Adds results

@@
 \begin{tabular}{l||c c c c c c} 
- & Bris4v5 & Select4 & Bris8 & Bris8v5 & Select8 \\ 
+& bris8 & bris4 & bris8v5 & bris4v5 & select8 & select4  \\
  \hline \hline 
-all (5M)        & - & 189.8 & 194.0 & 230.3 & 139.2 \\
-all (50M)       & - & 292.0 & 256.0 & 304.4 & 232.7 \\
-all (500M)      & - &  -    & 299.9 & 369.2 & 335.1 \\
-twolarge (5M)   & - & 164.1 & 93.9  & 216.7 & 117.5 \\
-twolarge (50M)  & - & 270.4 & 140.0 & 159.8 & 210.6 \\
-twolarge (500M) & - & 381.4 & 169.7 & 383.6 & 307.8 \\
+all (5M)   & 194.10 & 578.20 & 232.30 & 461.80 & 132.00 & 187.60 \\
+all (50M)   & 255.40 & 719.10 & 300.70 & 558.90 & 214.20 & 291.00 \\
+all (500M)   & 298.10 & 821.90 & 368.10 & 644.90 & 307.20 & 388.10 \\
+twolarge (5M)   & 96.40 & 331.70 & 107.70 & 260.20 & 119.40 & 165.20 \\
+twolarge (50M)   & 142.30 & 406.50 & 159.30 & 332.10 & 212.10 & 269.20 \\
+twolarge (500M)   & 168.50 & 516.10 & 191.80 & 393.80 & 304.90 & 381.30 \\
+onelarge (5M)   & 25.80 & 50.80 & 26.90 & 42.30 & 94.60 & 106.10 \\
+onelarge (50M)   & 50.00 & 85.30 & 51.00 & 72.30 & 197.90 & 241.00 \\
+onelarge (500M)   & 55.10 & 96.80 & 59.10 & 80.50 & 294.30 & 351.70 \\
+onlysmall (5M)   & 16.40 & 17.60 & 16.60 & 16.50 & 82.40 & 94.90 \\
+onlysmall (50M)   & 31.90 & 36.90 & 32.50 & 31.20 & 186.50 & 226.90 \\
+onlysmall (500M)   & 34.60 & 39.40 & 35.10 & 34.10 & 278.00 & 333.90 \\
 

### 2020-03-02 vb_encode_4bit_byteoptimized.cpp

@@ -46,7 +46,7 @@ int main(int argc, char *argv[]) {
 
 
   bit_vector b(data.size(), 0);
-  uint8_t *iv  = new uint8_t[data.size()];
+  uint8_t *iv  = new uint8_t[data.size()+8];
   size_t index = 0;
   size_t num_index = -1;
   size_t num_p = 0;

### 2020-03-02 gradu/msc_english_contents.tex
Changes

@@
 
-$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better.
+$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better. Additionally, the
+proposed $select$ method offers quick access to next or previous element.
 
@@
 
-For example, if block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
+For example, block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
 \texttt{00000000 00101000}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1000}. The block with least significant bits is given a continuation bit 1.
 The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11000}, 
-which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}.
+which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}. Prepend adds an element to the beginning of the list 
+and extend adds all the elements of the second list to the end of the first list.
 
@@
 \section{VB decoding}
-VB decoding is technically just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
-are removed from all the chunks and the blocks are concatenated to form the original number. An example implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.
+VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
+are removed from all the chunks and the blocks are concatenated to form the original integer. A pseudo code implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.
+Append adds an element to the end of the list. If block length is changed, additional bit operation steps when reading the data may be needed.
 
-As in the encoding example, block length is 4, the encoded message 
-is \texttt{00010 11000} and n is a 16 bit unsigned integer. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
+As how the encoding example ended, the encoded message was \texttt{00010 11000} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
+extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
 The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
 changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
-$n = $\texttt{101000} (bit representation of 42). Because the continuation bit of this chunk is 1, decoding for this number is complete.
+$n = $\texttt{101000} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.
 
@@
 \State $n\gets 128\times n $ + $b$ - $128$
-\State $numbers$.append($n$)
+\State \Call{append}{numbers,n}
 \State $n\gets 0$
@@
 Rankv5 & \text{6.25\% of bit array}\\
-Select & \text{~8.3\% of bit array (on average)}\\
+Select & \text{~8-12\% of bit array}\\
 \hline
@@
 Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
-library \citep{gbmp2014sea}. Each implementation is decoding a uint64\_t integer.
+library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t integers.
 
@@
 
-Five differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).
+Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).
 
@@
   \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
-  \item twolarge2 - Same as twolarge, but third of the 1 byte integers 4 bits or less
   \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
-  \item onebyte - all integers 4 bits or less
+  \item onlysmall - all integers 4 bits or less
 \end{itemize}
@@
 All implementations read the data set from a file and compress it to memory. Then one million index integers are randomized and stored to an array. The times shown in Table~\ref{table:results1} 
-are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.
+are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated multiple times and an average of the runtimes is taken.
 
@@
 
- - Run with 32 bit values, see what happens (some operations may be 32b?)
- -
+ - get compression sizes of 4 and 8 bit versions
+
+ - figure out why data size matters
 

### 2020-03-02 bris4.cpp
Adds changes

@@ -1,0 +1,145 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 4;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 16> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[16];
+  std::array<std::vector<uint8_t>, 16> iv;
+  size_t index = 0;
+  for(int i = 0; i < 16; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  rank_support_v5<0> rb[16];
+  for(int i =0; i < 16; i++) {
+    rank_support_v5<0> r(&b[i]);
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
+//  cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
[…]

### 2020-03-02 bris4_v5.cpp

@@ -1,0 +1,145 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 4;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 16> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[16];
+  std::array<std::vector<uint8_t>, 16> iv;
+  size_t index = 0;
+  for(int i = 0; i < 16; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  rank_support_v<0> rb[16];
+  for(int i =0; i < 16; i++) {
+    rank_support_v<0> r(&b[i]);
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
+//  cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
[…]

### 2020-03-02 bris8.cpp

@@ -1,0 +1,149 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[8];
+  std::array<std::vector<uint8_t>, 8> iv;
+  size_t index = 0;
+
+  for(int i = 0; i < 8; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  select_support_mcl<> sls[8];
+  rank_support_v<0> rb[8];
+  for(int i =0; i < 8; i++) {
+    select_support_mcl<> s(&b[i]);
+    rank_support_v<0> r(&b[i]);
+    sls[i] = s;
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
[…]

### 2020-03-02 bris8_v5.cpp

@@ -1,0 +1,146 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[8];
+  std::array<std::vector<uint8_t>, 8> iv;
+  size_t index = 0;
+
+  for(int i = 0; i < 8; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  rank_support_v5<0> rb[8];
+  for(int i =0; i < 8; i++) {
+    rank_support_v5<0> r(&b[i]);
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
+//  cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
[…]

### 2020-03-02 generate.cpp

@@ -21,10 +21,11 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-  // int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
-  // int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- smallheavy
+   int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
+  //int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- twolarge
   // int maxvalues [8] = {1<<3,1<<4,1<<5,1<<6,1<<7,1<<8,1<<16,1<<31}; // <- vsmall
-  int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- vvsmallheavy
+  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- onelarge
+  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
     int rand = dis(gen)%8;

### 2020-03-02 select4.cpp

@@ -1,0 +1,171 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include <bitset>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 4;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int bsize = 64;
+
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+//  uint8_t *siv  = new uint8_t[8];
+//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
+//  uint64_t *x = (uint64_t *)&siv[0];
+//  *x = original[0]<<4;
+//  cout << *x << endl;
+//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
+//  exit(1);
+
+  vector<uint32_t> data;
+
+//  REDO THIS:
+//  calculate bits on the fly and save number to block array intact
+
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.back() += cap;
+    data.insert(data.end(), v.begin(), v.end());
+  }
+
+
+  bit_vector b(data.size(), 0);
+  uint8_t *iv  = new uint8_t[data.size()+4];
+  size_t index = 0;
+  size_t num_index = -1;
+  size_t num_p = 0;
+  uint8_t overflow = 0;
+  for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
+    b[index] = (*i>>bit_length) & 1;
+    if(b[index] == 1) {
+        uint64_t orig = original[num_p];
+        int diff = index-num_index;
+        int shiftsize = bsize-(diff*bit_length);
+        orig = (orig<<shiftsize)>>shiftsize;
+        num_index++;
+        if(num_index%2 == 1) {
+          if(diff == 16) {
[…]

### 2020-03-02 select8.cpp

@@ -1,0 +1,127 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include <bitset>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  vector<uint32_t> data;
+
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.back() += cap;
+    data.insert(data.end(), v.begin(), v.end());
+  }
+
+  bit_vector b(data.size(), 0);
+  uint8_t *iv  = new uint8_t[data.size()];
+  size_t index = 0;
+
+  for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
+    b[index] = (*i>>bit_length) & 1;
+    iv[index] = *i;
+  }
+
+  bit_vector::select_1_type sls(&b);
+
+  //select_support_mcl<> sls(&b);
+
+  cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
+  cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;
+  cout << "Data vector memory size: " << sizeof(uint8_t) * data.size() << "bytes" << endl;
+  srand((unsigned) time(0));
+
+  cout << "number of numbers: " << original.size() << endl;
+
+  vector<unsigned int> indices(random_accesses, 0);
+  for(vector<uint64_t>::size_type i = 0; i < indices.size(); i++) {
+    indices[i] = rand()%original.size();
+  }
+
+  uint64_t z = 0;
+//  CALLGRIND_START_INSTRUMENTATION;
+uint64_t *test;
+int end;
+uint8_t diff;
[…]

### 2020-03-02 gradu/msc_english_contents.tex
Changed rank4 implementation

@@
 \begin{tabular}{l||c c c c c c} 
-& bris8 & bris4 & bris8v5 & bris4v5 & select8 & select4  \\
+& bris8 & bris4v5 & bris8v5 & bris4 & select8 & select4  \\
  \hline \hline 
@@
 
+ - 4bit is naturally slower, because of smaller block size
+ - select is better with longer numbers, because it only needs 1 select call and reads straight from memory
+
+ - subarray access(?)
+
  - get compression sizes of 4 and 8 bit versions

### 2020-03-02 bris4.cpp

@@ -49,8 +49,16 @@ int main(int argc, char *argv[]) {
     bit_vector bv(data[i].size(), 0);
     index = 0;
     for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
-        bv[index] = (*j>>bit_length) & 1;
-        vec[index] = *j%(cap);
+      uint8_t value = *j;
+      if(index%2 == 0) {
+        vec[index/2] = (value%cap)<<4;
+      }
+      else {
+        vec[index/2] += value%cap;
+
+      }
+//      cout << (int)*j <<" " << (int)value << " " << (int)vec[index/2] << endl;
+      bv[index] = (*j>>bit_length) & 1;
     }
 
     iv[i] = vec;
@@ -58,9 +66,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  rank_support_v5<0> rb[16];
+  rank_support_v<0> rb[16];
   for(int i =0; i < 16; i++) {
-    rank_support_v5<0> r(&b[i]);
+    rank_support_v<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);
@@ -89,8 +97,14 @@ int main(int argc, char *argv[]) {
     while(b[level][index] == 0) {
 //      cout << "orig:" << original[*i] << endl;
 //      cout << "l:" <<level << " " << index << endl;
-      uint8_t d = iv[level][index];
-//      cout << "d " << (int)d << endl;
+      uint8_t d = iv[level][index/2];
+      if(index%2==1) {
+        d=d&15;
+      }
+      else {
+        d=d>>4;
+      }
+//      cout << (int)d << endl;
       val = (val<<bit_length) + d;
 //      cout << "val " << val << endl;
       index = rb[level](index);
@@ -98,8 +112,15 @@ int main(int argc, char *argv[]) {
     }
 //    cout << "orig:" << original[*i] << endl;
 //    cout << "l:"<<level << " " << index << endl;
+    uint8_t e = iv[level][index/2];
+    if(index%2==1) {
+      e=e&15;
+    }
+    else {
+      e=e>>4;
+    }
 
-    val = (val<<bit_length) + iv[level][index];
+    val = (val<<bit_length) + e;
 //    cout << "val " << val << endl;
 
   //  int begin = index == 0 ? 0 : sls(index)+1;

### 2020-03-02 bris4_v5.cpp

@@ -49,8 +49,16 @@ int main(int argc, char *argv[]) {
     bit_vector bv(data[i].size(), 0);
     index = 0;
     for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
-        bv[index] = (*j>>bit_length) & 1;
-        vec[index] = *j%(cap);
+      uint8_t value = *j;
+      if(index%2 == 0) {
+        vec[index/2] = (value%cap)<<4;
+      }
+      else {
+        vec[index/2] += value%cap;
+
+      }
+//      cout << (int)*j <<" " << (int)value << " " << (int)vec[index/2] << endl;
+      bv[index] = (*j>>bit_length) & 1;
     }
 
     iv[i] = vec;
@@ -58,9 +66,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  rank_support_v<0> rb[16];
+  rank_support_v5<0> rb[16];
   for(int i =0; i < 16; i++) {
-    rank_support_v<0> r(&b[i]);
+    rank_support_v5<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);
@@ -89,8 +97,14 @@ int main(int argc, char *argv[]) {
     while(b[level][index] == 0) {
 //      cout << "orig:" << original[*i] << endl;
 //      cout << "l:" <<level << " " << index << endl;
-      uint8_t d = iv[level][index];
-//      cout << "d " << (int)d << endl;
+      uint8_t d = iv[level][index/2];
+      if(index%2==1) {
+        d=d&15;
+      }
+      else {
+        d=d>>4;
+      }
+//      cout << (int)d << endl;
       val = (val<<bit_length) + d;
 //      cout << "val " << val << endl;
       index = rb[level](index);
@@ -98,8 +112,15 @@ int main(int argc, char *argv[]) {
     }
 //    cout << "orig:" << original[*i] << endl;
 //    cout << "l:"<<level << " " << index << endl;
+    uint8_t e = iv[level][index/2];
+    if(index%2==1) {
+      e=e&15;
+    }
+    else {
+      e=e>>4;
+    }
 
-    val = (val<<bit_length) + iv[level][index];
+    val = (val<<bit_length) + e;
 //    cout << "val " << val << endl;
 
   //  int begin = index == 0 ? 0 : sls(index)+1;

### 2020-03-02 vb_encode_4bit_bris.cpp

@@ -58,9 +58,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  rank_support_v5<0> rb[16];
+  rank_support_v<0> rb[16];
   for(int i =0; i < 16; i++) {
-    rank_support_v5<0> r(&b[i]);
+    rank_support_v<0> r(&b[i]);
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);

### 2020-03-02 vb_encode_8bit_byteoptimized.cpp

@@ -92,7 +92,7 @@ CALLGRIND_START_INSTRUMENTATION;
 // TODO: try out with builtin_clz
 //    test = (uint64_t *)&iv[begin];
     int offset = (begin)%bsize;
-    int block = (begin)/bsize;
+    int block = (begin)>>bsize;
     uint64_t blokki = *(b.data()+block);
     val = blokki >> offset;
     if(offset) {

### 2020-03-02 bris4.cpp
Changes to bris4

@@ -45,7 +45,7 @@ int main(int argc, char *argv[]) {
   std::array<std::vector<uint8_t>, 16> iv;
   size_t index = 0;
   for(int i = 0; i < 16; i++) {
-    vector<uint8_t> vec(data[i].size(),0);
+    vector<uint8_t> vec(data[i].size()/2 +1,0);
     bit_vector bv(data[i].size(), 0);
     index = 0;
     for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {

### 2020-03-02 bris4_v5.cpp

@@ -45,7 +45,7 @@ int main(int argc, char *argv[]) {
   std::array<std::vector<uint8_t>, 16> iv;
   size_t index = 0;
   for(int i = 0; i < 16; i++) {
-    vector<uint8_t> vec(data[i].size(),0);
+    vector<uint8_t> vec(data[i].size()/2 + 1,0);
     bit_vector bv(data[i].size(), 0);
     index = 0;
     for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {

### 2020-03-02 gradu/msc_english_contents.tex
Changes to bris4
Changes

@@
 
-Variable-byte (VB) encoding \citep{Wil99} is a method for compressing integers via omitting leading zero bits that would be present in a longer fixed 
+Variable-byte (VB) encoding \citep{Wil99} is a method for compressing unsigned integers via omitting leading zero bits that would be present in a longer fixed 
 width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
@@
 For example, block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
-\texttt{00000000 00101000}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1000}. The block with least significant bits is given a continuation bit 1.
-The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11000}, 
+\texttt{00000000 00101010}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1010}. The block with least significant bits is given a continuation bit 1.
+The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11010}, 
 which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}. Prepend adds an element to the beginning of the list 
@@
 
-As how the encoding example ended, the encoded message was \texttt{00010 11000} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
+As how the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
 extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
@@
 changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
-$n = $\texttt{101000} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.
+$n = $\texttt{101010} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.
 
@@
 data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
-strutures. It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
+structures. It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
 list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
-of each element, but it adds a very large overhead which removes the benefit of compression.
+of each element to an array, but it adds a very large overhead which removes the benefit of compression.
 
@@
 
-The rank and select implementations used in this thesis are from C++ library by \citep{gbmp2014sea}. The library has several implementations of both rank and 
-select as well as an implementaion of a bit array. Table~\ref{table:supportsize} has $rank$ and $select$ size requirements of implementations used in this experiment.
-Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on the number of 1's in the data.
+The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and  several implementations
+ of both rank and select to support the bit array. Also a few useful functions were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
+size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
+the number of 1's in the data.
 
@@
 
+The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
+with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Bris8 performs better than Bris8v5, but also requires 
+more memory. See Figure~\ref{table:supportsize} for memory consumption.
+
+8-bit versions perform better than 4-bit versions timewise, because they need less operations overall.
+However 4-bit versions have better compression in some data sets, because they are able to compress
+some integers better.
+
+ - see how data length percentage matters (do datasets with k/8 long numbers).
+ - see how cache matters
+
  - 4bit is naturally slower, because of smaller block size

### 2020-03-03 gradu/msc_english_contents.tex
blb

@@
 \caption{Results in milliseconds, smaller is better.\label{table:results1}}
-\begin{tabular}{l||c c c c c c} 
-& bris8 & bris4v5 & bris8v5 & bris4 & select8 & select4  \\
+\begin{tabular}{l||c c c c c c c} 
+& bris8 & bris4 & bris8v5 & bris4v5 & select8 & select4  \\
  \hline \hline 
-all (5M)   & 194.10 & 578.20 & 232.30 & 461.80 & 132.00 & 187.60 \\
-all (50M)   & 255.40 & 719.10 & 300.70 & 558.90 & 214.20 & 291.00 \\
-all (500M)   & 298.10 & 821.90 & 368.10 & 644.90 & 307.20 & 388.10 \\
-twolarge (5M)   & 96.40 & 331.70 & 107.70 & 260.20 & 119.40 & 165.20 \\
-twolarge (50M)   & 142.30 & 406.50 & 159.30 & 332.10 & 212.10 & 269.20 \\
-twolarge (500M)   & 168.50 & 516.10 & 191.80 & 393.80 & 304.90 & 381.30 \\
-onelarge (5M)   & 25.80 & 50.80 & 26.90 & 42.30 & 94.60 & 106.10 \\
-onelarge (50M)   & 50.00 & 85.30 & 51.00 & 72.30 & 197.90 & 241.00 \\
-onelarge (500M)   & 55.10 & 96.80 & 59.10 & 80.50 & 294.30 & 351.70 \\
-onlysmall (5M)   & 16.40 & 17.60 & 16.60 & 16.50 & 82.40 & 94.90 \\
-onlysmall (50M)   & 31.90 & 36.90 & 32.50 & 31.20 & 186.50 & 226.90 \\
-onlysmall (500M)   & 34.60 & 39.40 & 35.10 & 34.10 & 278.00 & 333.90 \\
+all (5M)   & 194.50 & 473.00 & 232.40 & 578.00 & 133.50 & 186.10 \\
+all (50M)   & 270.90 & 580.00 & 323.80 & 786.50 & 214.30 & 291.50 \\
+all (500M)   & 298.90 & 678.80 & 368.80 & 940.30 & 308.10 & 388.10 \\
+twolarge (5M)   & 92.60 & 270.70 & 109.60 & 328.90 & 117.20 & 166.50 \\
+twolarge (50M)   & 140.60 & 378.30 & 159.20 & 458.50 & 210.60 & 269.60 \\
+twolarge (500M)   & 168.50 & 439.90 & 191.30 & 586.00 & 304.50 & 382.10 \\
+onelarge (5M)   & 26.10 & 42.20 & 27.20 & 50.10 & 95.30 & 106.10 \\
+onelarge (50M)   & 50.00 & 91.20 & 51.20 & 104.90 & 199.60 & 239.90 \\
+onelarge (500M)   & 55.20 & 100.50 & 59.00 & 123.90 & 294.70 & 351.30 \\
+onlysmall (5M)   & 16.60 & 21.10 & 16.60 & 23.00 & 80.20 & 95.10 \\
+onlysmall (50M)   & 31.60 & 51.90 & 32.40 & 57.00 & 186.50 & 227.30 \\
+onlysmall (500M)   & 34.40 & 56.00 & 35.00 & 63.50 & 279.30 & 335.40 \\
 

### 2020-03-03 select8.cpp

@@ -64,7 +64,7 @@ int end;
 uint8_t diff;
 uint64_t val;
 int begin;
-int bsize = sizeof(b.data()) * 8;
+int bsize = 64;
 cout << "bsize" << bsize << endl;
 int shiftamt;
 uint64_t maxuint = 0-1;
@@ -74,7 +74,7 @@ uint64_t bitmasks[8];
 val = 0;
 for(int i = 0;i < 8; i++) {
   val = val << 8;
-  val = val | 0xF;
+  val = val | 0xFF;
   bitmasks[i] = val;
 }
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
@@ -98,7 +98,7 @@ for(int i = 0;i < 8; i++) {
       uint64_t blokki2 = *(b.data()+block+1);
       val = val | (blokki2 <<(64-offset));
     }
-    diff = bits::lo(val); //ADD +1 for shiftamt stuff
+    //diff = bits::lo(val); //ADD +1 for shiftamt stuff
 
 //    if ((val & 0x0000000F) == 0) {diff = diff + 4; val = val >> 4;}
 //    if ((val & 0x00000003) == 0) {diff = diff + 2; val = val >> 2;}
@@ -111,7 +111,7 @@ for(int i = 0;i < 8; i++) {
 //    val = *test<<(shiftamt)>>shiftamt;
     test = (uint64_t *)&iv[begin];
 
-    val = *test & bitmasks[diff];
+    val = *test & bitmasks[bits::lo(val)];
     z = z^val;
 
     if(0 && val != original[index]) {

### 2020-03-04 gradu/msc_english_contents.tex
Optimized 4bit more

@@
 
-If the original element size is 32 bits and the data was compressed to k blocks of length b. The starting byte s of the element is $select$(i) * b div 8 where div is the integer division. 
-Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit word is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
-bits and bit shift right 32-b*k to re-align bits. If block size is one byte, offset calculation is not needed and thus the process is slightly faster.
+For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$(i) gets the index location
+of needed block. Because the block length differs from byte length, the the two values need to be calculated. First, the starting byte location $s$ needs to be calculated 
+$(s = select_1(i) \times 4) \div 8$, the latter operand being the integer division.
+Second, the first block might not be at the start of the byte $s$.
+This is why offset $o = select_1(i) \times 4) mod 8$
+is needed (in the example case, o equals to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.
+
+At this point, $w$ contains the wanted bits, but also has extra bits from the previous and next compressed integers.
+If offset was not zero, there are trailing $o$ bits in $w$ from the previous integer.
+These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the requested value. Because most systems are
+able to read memory only from byte addresses, there is a corner case where the value is stored in too many bytes for a word.
+In this setup, this happens if compressed length was 8 blocks and offset was 4.
+Then another step of reading another byte and storing the remaining bits to $w$ is required.
+This is entirely avoidable if integer length is constrained.
+In this case maximum integer length would be 28 bits.
 
@@
 
+ - See Ahn \& Moffat 2005; Lemire et al. 2018
  - subarray access(?)
+
+ - compare to elias-fano?
 

### 2020-03-04 select4.cpp

@@ -102,6 +102,13 @@ cout << "bsize" << bsize << endl;
 int shiftamt;
 uint64_t maxuint = 0-1;
 cout << maxuint << endl;
+uint64_t bitmasks[16];
+val = 0;
+for(int i = 0;i < 16; i++) {
+  val = val << 4;
+  val = val | 0xF;
+  bitmasks[i] = val;
+}
 chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
@@ -123,34 +130,15 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
       uint64_t blokki2 = *(b.data()+block+1);
       val = val | (blokki2 <<(64-offset));
     }
-    diff = bits::lo(val) + 1;
-//    if(diff > 15) {
-//      cout <<"LENGTH:" << (int) diff << endl;
-//    }
-//    if ((val & 0x0000000F) == 0) {diff = diff + 4; val = val >> 4;}
-//    if ((val & 0x00000003) == 0) {diff = diff + 2; val = val >> 2;}
-//    diff = diff -(val & 1);
-//    uint8_t *nro = (uint8_t *)&b[begin];
-//      end = sls(index+1); //TODO this has to be optimized
-//      if(end-begin != diff) {
-//        cout << "bl " << (int)diff << " vs " << (end-begin)<< endl;
-//      }
-//      diff = 1;//end-begin;
+      uint8_t diff = bits::lo(val);
+
+
     test = (uint64_t *)&iv[begin/2];
-//    cout << bitset<64>(*test) << endl;
-    val = *test;
-    if(begin%2==1) {
-      val>>=bit_length;
-      if(diff==16) {
-        uint64_t *test2 = (uint64_t *)&iv[1+(begin/2)];
-//        cout << bitset<64>(*test2) << endl;
-        val = (val&0x0FFFFFFFFFFFFFFF) | ((*test2>>56)<<60);
-      }
-    }
 
-    shiftamt = 64-bit_length*diff;
-    val = val<<(shiftamt)>>shiftamt;
+    val = *test>>((begin%2)*bit_length);
 
+    val = val & bitmasks[diff];
+    
     z = z^val;
 
     if(0 && val != original[index]) {

### 2020-03-04 select8.cpp

@@ -69,7 +69,6 @@ cout << "bsize" << bsize << endl;
 int shiftamt;
 uint64_t maxuint = 0-1;
 cout << maxuint << endl;
-chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
 uint64_t bitmasks[8];
 val = 0;
 for(int i = 0;i < 8; i++) {
@@ -77,19 +76,14 @@ for(int i = 0;i < 8; i++) {
   val = val | 0xFF;
   bitmasks[i] = val;
 }
+chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
 
 //TODO CALLGRIND THIS
     begin = index == 0 ? 0 : sls(index)+1;
-//    while(*(b.begin()+begin+diff)==0) diff++;
-//    bit_vector::iterator iter = b.begin()+begin;
-//    while(*iter == 0) {
-//      iter++;
-//    }
-//    uint8_t diff = std::distance(b.begin()+begin, iter);
-// TODO: try out with builtin_clz
-//    test = (uint64_t *)&iv[begin];
+
+//TODO see if this can be read on byte level instead of 64bit level - would save the extra calculation
     int offset = (begin)%bsize;
     int block = (begin)/bsize;
     uint64_t blokki = *(b.data()+block);

### 2020-03-05 gradu/msc_english_contents.tex
Updated results

@@
  \hline \hline 
-all (5M)   & 194.50 & 473.00 & 232.40 & 578.00 & 133.50 & 186.10 \\
-all (50M)   & 270.90 & 580.00 & 323.80 & 786.50 & 214.30 & 291.50 \\
-all (500M)   & 298.90 & 678.80 & 368.80 & 940.30 & 308.10 & 388.10 \\
-twolarge (5M)   & 92.60 & 270.70 & 109.60 & 328.90 & 117.20 & 166.50 \\
-twolarge (50M)   & 140.60 & 378.30 & 159.20 & 458.50 & 210.60 & 269.60 \\
-twolarge (500M)   & 168.50 & 439.90 & 191.30 & 586.00 & 304.50 & 382.10 \\
-onelarge (5M)   & 26.10 & 42.20 & 27.20 & 50.10 & 95.30 & 106.10 \\
-onelarge (50M)   & 50.00 & 91.20 & 51.20 & 104.90 & 199.60 & 239.90 \\
-onelarge (500M)   & 55.20 & 100.50 & 59.00 & 123.90 & 294.70 & 351.30 \\
-onlysmall (5M)   & 16.60 & 21.10 & 16.60 & 23.00 & 80.20 & 95.10 \\
-onlysmall (50M)   & 31.60 & 51.90 & 32.40 & 57.00 & 186.50 & 227.30 \\
-onlysmall (500M)   & 34.40 & 56.00 & 35.00 & 63.50 & 279.30 & 335.40 \\
+all (5M)   & 191.80 & 471.50 & 234.00 & 569.40 & 149.20 & 175.90 \\
+all (50M)   & 269.20 & 573.80 & 323.00 & 785.00 & 245.40 & 275.00 \\
+all (500M)   & 298.50 & 677.80 & 368.20 & 941.80 & 355.90 & 382.20 \\
+twolarge (5M)   & 96.70 & 275.50 & 110.70 & 329.00 & 125.60 & 151.90 \\
+twolarge (50M)   & 140.40 & 379.70 & 159.80 & 456.80 & 227.70 & 242.20 \\
+twolarge (500M)   & 168.20 & 441.00 & 191.50 & 584.70 & 330.20 & 341.00 \\
+onelarge (5M)   & 25.80 & 40.10 & 28.30 & 49.20 & 95.90 & 100.40 \\
+onelarge (50M)   & 49.80 & 91.30 & 50.90 & 105.70 & 212.70 & 213.20 \\
+onelarge (500M)   & 55.00 & 100.00 & 58.50 & 124.00 & 315.80 & 307.70 \\
+onlysmall (5M)   & 16.50 & 20.20 & 16.50 & 23.80 & 84.70 & 88.40 \\
+onlysmall (50M)   & 31.20 & 51.80 & 32.10 & 57.00 & 201.20 & 198.80 \\
+onlysmall (500M)   & 34.10 & 56.00 & 35.00 & 63.10 & 297.90 & 293.80 \\
 

### 2020-03-05 gradu/msc_english_contents.tex
Changes

@@
 
+ - TODO figure out why 4bit crashes sometimes with 500M_all
+
  - see how data length percentage matters (do datasets with k/8 long numbers).
@@
  - 4bit is naturally slower, because of smaller block size
+
+ - actually may not be the case with select (maybe cache is causing this?)
+
+ - better dataset might be something with larger (like 60bit) numbers
+
  - select is better with longer numbers, because it only needs 1 select call and reads straight from memory

### 2020-03-09 gradu/msc_english_contents.tex
More text

--- 2020-03-05 gradu/msc_english_contents.tex
@@
 Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. Their encoding process assigns short bit array codes to small integers, 
-which is why they outperform VB encoding on datasets with a lot of small integers \Citep{Wil99}.
+which is why they outperform VB encoding on datasets with a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, the compressable data sets will have
+the focus on slightly larger, unordered numbers.
 
@@
 
-For example, block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
-\texttt{00000000 00101010}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1010}. The block with least significant bits is given a continuation bit 1.
-The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11010}, 
-which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}. Prepend adds an element to the beginning of the list 
+For example, the standard bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. When split to blocks of length 4, it becomes \texttt{0000 0000 0010 1010}. The empty blocks 
+are removed and then continuation bits are added.
+1 is added to the least significant block and 0 to the other block, resulting in \texttt{00010 11010}.
+A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}.
+Prepend adds an element to the beginning of the list
 and extend adds all the elements of the second list to the end of the first list.
@@
 With small integers, this
-is all that is required. If the data has a lot of small integers, the $rank$ method has a huge advantage.
+is all that is required.
+If the data has a lot of small integers, the $rank$ method has a huge advantage.
+In addition, the block length of this method can change within different array levels.
+This opens
+a possibility for further optimization.
 
@@
 
---TODO: create another version: 7bit bris.
-Store bits in both bit array and chunk (to see if it's faster to get the continuation bit from the chunk than from the bit array) (also needs 7bit compliant dataset)
-
 \chapter{DAC with select query}
@@
 Second, the first block might not be at the start of the byte $s$.
-This is why offset $o = select_1(i) \times 4) mod 8$
+This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$
 is needed (in the example case, o equals to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.
@@
 
+\section{Elias-Fano encoding}
+Direct access via select can also be implemented with Elias-Fano (TODO citation to Elias-Fano?), but the data has to be sorted. In Elias-Fano encoding, the data is at first split into low and 
+high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bits is trivial. The higher bits are gap-encoded and then encoded in unary. 
+For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}.
+$Select_1$ has an interesting interaction with gap encoded unary arrays.
+$k = Select_1(B,i)$
+returns the location of i-th 1 in B. Up to index $k$, the bit array has had $i$ ones and $i-k$ zeros. The $i$ corresponds to the number of compressed integers in $B$ and every zero in the array 
+means an increment of 1 in the numbers. In other words, $i-k$ equals to the integer at position $i$ before gap-encoding. Therefore $select_1(i)$ - i returns the original higher bit number. The split 
+between bits is made to reduce the gap sizes.
+This approach can be further optimized by finding clusters in the integers and partitioning the indexes \citep{Ott14}.
+
+\chapter{Sub-array access}
+After locating and decompressing one value from a compressed data structure, compressing the subsequent items gets a lot easier.
+Because the next compressed value is usually located next to the fetched
+in memory, locating the data is not needed. This is especially beneficial to the VByte decoding with $select$ method, where calculating the start byte location takes majority of he runtime. In 
+addition to this, the length of the next compressed integer is faster to calculate because already manipulated data in memory.
+
+The shown $rank$ method by Brisaboa et al.
+can also effectively decompress the next value but some modifications are needed.
+When fetching a longer subarray, the goal becomes to minimize $rank$ calls.
+If an additional pointer array is stored during subarray decompression, only one $rank$ call for each accessed block level is needed.
+This can be observed from Figure~\ref{bris_ds}; the blocks of subarray
+are located next to each other in different block levels.
+
+ - examples / visualizaton?
+
+ - TODO get results from this
 \chapter{Experimental results}
@@
 The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
-The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The code was compiled with g++-7 -std=c++11 -DNDEBUG and with libraries -lsdsl -ldivsufsort -ldivsufsort64.
+The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG} and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
 
 Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
-library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t integers.
+library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers.
 
@@
 
-All implementations read the data set from a file and compress it to memory. Then one million index integers are randomized and stored to an array. The times shown in Table~\ref{table:results1} 
+All implementations read the data set from a file and compress it to memory, randomize one million indexes and stored to an array. The times shown in Table~\ref{table:results1} 
 are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated multiple times and an average of the runtimes is taken.
@@
 The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
-with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Bris8 performs better than Bris8v5, but also requires 
-more memory. See Figure~\ref{table:supportsize} for memory consumption.
+with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. The 8-bit versions tend to perform better than the 
+4-bit versions.
+This is because 8-bit version stores more data in one block and less block readings are required.
+However, 4bit versions achieve better compression in data sets containing small integers.
+Select versions do not depend as much on the block size, because all the blocks in 
+the compressed element are read at one go.
 
-8-bit versions perform better than 4-bit versions timewise, because they need less operations overall.
-However 4-bit versions have better compression in some data sets, because they are able to compress
-some integers better.
+In rank, the v5 version naturally has worse performance because more calculations are needed to locate the correct index. As a trade-off, it requires 
+significantly less memory, as seen on Figure~\ref{table:supportsize}.
+
+Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
+in every scenario. This is most likely due to cache, because in the 4-bit version the stored data set is smaller.
+
+ - TODO figure out why 4bit crashes sometimes with 500M\_all
 
  - see how data length percentage matters (do datasets with k/8 long numbers).
+
  - see how cache matters
 
- - 4bit is naturally slower, because of smaller block size
- - select is better with longer numbers, because it only needs 1 select call and reads straight from memory
+ - better dataset might be something with larger (like 60bit) numbers
 
  - See Ahn \& Moffat 2005; Lemire et al. 2018
+
  - subarray access(?)
 
- - compare to elias-fano?
+ - bris has advantage of different size blocks (in different levels)
 
@@
 \chapter{Future work}
- - something to improve / research?
+
+There has already been a few successful attempts to decompress VB encoded integers with SIMD \citep{Lem18,Pla15}.
+It is worthwhile to check if the presented approach benefits from SIMD.
+The data needs
+less manipulation because the data is already written as it is, but calculating the integer block lengths might benefit from SIMD.
+
+Another improvement to VB encoding is to see whether the compressed numbers can be changed.
+This is trivial until the block length of the compressed number changes.
+Then an additional data structure or
+cache is needed and the supporting rank/select data structure needs to be recalculated.
 

### 2020-03-09 select4.cpp

@@ -46,7 +46,7 @@ int main(int argc, char *argv[]) {
 
 
   bit_vector b(data.size(), 0);
-  uint8_t *iv  = new uint8_t[data.size()+4];
+  uint8_t *iv  = new uint8_t[data.size()/2+4];
   size_t index = 0;
   size_t num_index = -1;
   size_t num_p = 0;
@@ -138,7 +138,7 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
     val = *test>>((begin%2)*bit_length);
 
     val = val & bitmasks[diff];
-    
+
     z = z^val;
 
     if(0 && val != original[index]) {

### 2020-03-09 select8.cpp

@@ -84,8 +84,8 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
     begin = index == 0 ? 0 : sls(index)+1;
 
 //TODO see if this can be read on byte level instead of 64bit level - would save the extra calculation
-    int offset = (begin)%bsize;
-    int block = (begin)/bsize;
+    int offset = begin&0x3F;
+    int block = begin>>6;
     uint64_t blokki = *(b.data()+block);
     val = blokki >> offset;
     if(offset) {

### 2020-03-09 select4.cpp
Shaved off a few cycles

@@ -126,7 +126,7 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
     int block = begin/bsize;
     uint64_t blokki = *(b.data()+block);
     val = blokki >> offset;
-    if(offset) {
+    if(!val) {
       uint64_t blokki2 = *(b.data()+block+1);
       val = val | (blokki2 <<(64-offset));
     }

### 2020-03-09 select8.cpp

@@ -82,27 +82,20 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
 
 //TODO CALLGRIND THIS
     begin = index == 0 ? 0 : sls(index)+1;
+//TODO see if ternary above can be substituted with just sls call
+
 
 //TODO see if this can be read on byte level instead of 64bit level - would save the extra calculation
-    int offset = begin&0x3F;
+    int offset = (begin)%bsize;
     int block = begin>>6;
     uint64_t blokki = *(b.data()+block);
     val = blokki >> offset;
-    if(offset) {
+    if(!val) {
       uint64_t blokki2 = *(b.data()+block+1);
       val = val | (blokki2 <<(64-offset));
     }
-    //diff = bits::lo(val); //ADD +1 for shiftamt stuff
 
-//    if ((val & 0x0000000F) == 0) {diff = diff + 4; val = val >> 4;}
-//    if ((val & 0x00000003) == 0) {diff = diff + 2; val = val >> 2;}
-//    diff = diff -(val & 1);
-//    uint8_t *nro = (uint8_t *)&b[begin];
-//      end = sls(index+1); //TODO this has to be optimized
-//      diff = 1;//end-begin;
-//  TODO: try with bit mask!
-//    shiftamt = 64-bit_length*diff;
-//    val = *test<<(shiftamt)>>shiftamt;
+
     test = (uint64_t *)&iv[begin];
 
     val = *test & bitmasks[bits::lo(val)];

### 2020-03-19 gradu/msc_english_contents.tex
new stuff

@@
 If an additional pointer array is stored during subarray decompression, only one $rank$ call for each accessed block level is needed.
-This can be observed from Figure~\ref{bris_ds}; the blocks of subarray
+This can be observed from Figure~\ref{bris_ds}; the blocks of a subarray
 are located next to each other in different block levels.
@@
 \hline
-%
 \end{tabular}
@@
 The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
-with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. The 8-bit versions tend to perform better than the 
-4-bit versions.
-This is because 8-bit version stores more data in one block and less block readings are required.
-However, 4bit versions achieve better compression in data sets containing small integers.
-Select versions do not depend as much on the block size, because all the blocks in 
-the compressed element are read at one go.
+with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
+of one million $rank$ and $select$ calls with different average block lengths. The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior.
+
+The 8-bit versions tend to perform better than the 4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4bit versions achieve 
+better compression in data sets containing small integers. Select versions do not depend as much on the block size, because all the blocks in the compressed element are read at the same time. 
+Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
+in every scenario. This is most likely due to cache, because the stored data set is smaller in the 4-bit version.
 
@@
 
-Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
-in every scenario. This is most likely due to cache, because in the 4-bit version the stored data set is smaller.
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    xlabel={average integer byte length},
+    ylabel={ms per million queries},
+    xmin=1, xmax=8,
+    ymin=0, ymax=800,
+    xtick={1,2,3,4,5,6,7,8},
+    ytick={0,100,200,300,400,500,600,700,800},
+    legend pos=north west,
+    ymajorgrids=true,
+    grid style=dashed,
+]
+
+\addplot[
+color=blue,
+mark=square,
+]
+coordinates {
+(1,31.90)(2,111.00)(3,205.40)(4,304.20)(5,405.80)(6,513.90)(7,616.90)(8,725.10)
+};
+\addplot[
+color=green,
+mark=triangle,
+]
+coordinates {
+(1,32.90)(2,145.30)(3,271.20)(4,399.00)(5,528.20)(6,666.80)(7,767.40)(8,902.00)
+};
+\addplot[
+color=red,
+mark=ball,
+]
+coordinates {
+(1,202.40)(2,219.70)(3,228.70)(4,251.60)(5,261.80)(6,269.60)(7,275.70)(8,284.30)
+};
+\addplot[
+color=black,
+mark=line,
+]
+coordinates {
+(1,218.60)(2,255.20)(3,273.00)(4,282.90)(5,296.00)(6,306.10)(7,317.30)(8,327.50)
+};
+
+\legend{bris8, bris8v5, select8, select4 }
+
+\end{axis}
+\end{tikzpicture}
+\caption{Rank and select performances} \label{graph:data_comparison}
+\end{figure}
+
+Figures \ref{table:results8bit} and \ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures. The data sets consist of 50 million 64-bit numbers and 
+the compressed size is a sum of continuation bit array and the data blocks.
+
+The static memory requirement of $select$ is easily noticeable, 
+as is the relation between the memory requirement of $rank$ and the total number of blocks. Data structures in versions with smaller block size naturally take more space, but smaller block 
+size allows a better data compression in some data sets and therefore a smaller total size. This can be seen when comparing the two forementioned figures: All 4-bit versions of the algorithms require
+more space than the 8-bit versions, but the compressed size of the data is significantly smaller in the bottom two data sets.
+The data structure memory usage for the bottom row is similar in both versions
+due to the fact that the continuation bit array is exactly the same.
+
+\begin{table}
+\centering
+\begin{tabular}{l|| c c c c } 
+dataset   & compressed size & bris8  & bris8v5 & select8  \\
+ \hline \hline 
+all       & 168.4MB         & 4.67MB & 1.17MB   & 1.54MB    \\
+twolarge  & 112.4MB         & 3.12MB & 781kB    & 1.48MB    \\
+onelarge  &  63.2MB         & 1.76MB & 439kB    & 1.43MB    \\
+onlysmall &  56.3MB         & 1.56MB & 391kB    & 1.43MB    \\
+
+\hline
+%
+\end{tabular}
+\caption{Memory requirement for 50M 64-bit numbers 8bit blocks, smaller is better.\label{table:results8bit}}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{l|| c c c c } 
+dataset   & compressed size & bris4  & bris4v5 & select4  \\
+ \hline \hline 
+all       & 177.7MB         & 7.65MB & 1.91MB  & 1.63MB \\
+twolarge  & 116.1MB         & 4.66MB & 1.14MB  & 1.54MB \\
+onelarge  & 42.5MB          & 2.12MB & 531kB   & 1.44MB \\
+onlysmall & 31.3MB          & 1.56MB & 391kB   & 1.43MB \\
+
+\hline
+%
+\end{tabular}
+\caption{Memory requirement for 50M 64-bit numbers 4bit blocks, smaller is better.\label{table:results4bit}}
+\end{table}
 
  - TODO figure out why 4bit crashes sometimes with 500M\_all
-
- - see how data length percentage matters (do datasets with k/8 long numbers).
 
@@
 
- - See Ahn \& Moffat 2005; Lemire et al. 2018
-
- - subarray access(?)
+ - subarray access
 
  - bris has advantage of different size blocks (in different levels)
-
- - get compression sizes of 4 and 8 bit versions
 

### 2020-03-19 bris4.cpp
Adds subarray for select8

@@ -161,6 +161,8 @@ int main(int argc, char *argv[]) {
   }
   cout << "rank support size: " << sum << endl;
   cout << "bit vector size: " << bit_vec_size << endl;
-  cout << "Data size: " << sizeof(uint8_t)*data_size << endl;
+  cout << "Data size: " << sizeof(uint8_t)*data_size/2 << endl;
+  cout << "total data size:" << bit_vec_size + sizeof(uint8_t)*data_size/2 << endl;
+
   return 0;
 }

### 2020-03-19 bris4_v5.cpp

@@ -162,5 +162,7 @@ int main(int argc, char *argv[]) {
   cout << "rank support size: " << sum << endl;
   cout << "bit vector size: " << bit_vec_size << endl;
   cout << "Data size: " << sizeof(uint8_t)*data_size << endl;
+  cout << "total data size:" << bit_vec_size + sizeof(uint8_t)*data_size/2 << endl;
+
   return 0;
 }

### 2020-03-19 bris8.cpp

@@ -145,5 +145,6 @@ int main(int argc, char *argv[]) {
   cout << "rank support size: " << sum << endl;
   cout << "bit vector size: " << bit_vec_size << endl;
   cout << "Data size: " << sizeof(uint8_t)*data_size << endl;
+  cout << "total data size:" << bit_vec_size + sizeof(uint8_t)*data_size << endl;
   return 0;
 }

### 2020-03-19 bris8_v5.cpp

@@ -142,5 +142,7 @@ int main(int argc, char *argv[]) {
   cout << "rank support size: " << sum << endl;
   cout << "bit vector size: " << bit_vec_size << endl;
   cout << "Data size: " << sizeof(uint8_t)*data_size << endl;
+  cout << "total data size:" << bit_vec_size + sizeof(uint8_t)*data_size << endl;
+
   return 0;
 }

### 2020-03-19 generate.cpp

@@ -21,11 +21,11 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-   int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
-  //int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- twolarge
+  // int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
+  // int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- twolarge
   // int maxvalues [8] = {1<<3,1<<4,1<<5,1<<6,1<<7,1<<8,1<<16,1<<31}; // <- vsmall
   //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- onelarge
-  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
+  int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
     int rand = dis(gen)%8;

### 2020-03-19 select8.cpp

@@ -48,6 +48,8 @@ int main(int argc, char *argv[]) {
   cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
   cout << "select support vector memory size: " << size_in_bytes(sls) + sizeof(select_support_mcl<>) << "bytes" << endl;
   cout << "Data vector memory size: " << sizeof(uint8_t) * data.size() << "bytes" << endl;
+  cout << "total data size:" << size_in_bytes(b) + sizeof(uint8_t)*data.size() << endl;
+
   srand((unsigned) time(0));
 
   cout << "number of numbers: " << original.size() << endl;
@@ -92,7 +94,7 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
     val = blokki >> offset;
     if(!val) {
       uint64_t blokki2 = *(b.data()+block+1);
-      val = val | (blokki2 <<(64-offset));
+      val = (blokki2 <<(64-offset));
     }
 
 

### 2020-03-19 select8_subarray.cpp

@@ -13,11 +13,9 @@
 using namespace std;
 using namespace sdsl;
 
-uint32_t bit_length = 4;
+uint32_t bit_length = 8;
 unsigned int random_accesses = 1000000;
 uint32_t cap = 1<<bit_length;
-int bsize = 64;
-
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
@@ -25,18 +23,8 @@ int main(int argc, char *argv[]) {
   }
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
-//  uint8_t *siv  = new uint8_t[8];
-//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
-//  uint64_t *x = (uint64_t *)&siv[0];
-//  *x = original[0]<<4;
-//  cout << *x << endl;
-//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
-//  exit(1);
 
   vector<uint32_t> data;
-
-//  REDO THIS:
-//  calculate bits on the fly and save number to block array intact
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
     vector<uint32_t> v = vb_encode_number(*i, cap);
@@ -44,35 +32,13 @@ int main(int argc, char *argv[]) {
     data.insert(data.end(), v.begin(), v.end());
   }
 
-
   bit_vector b(data.size(), 0);
-  uint8_t *iv  = new uint8_t[data.size()/2+4];
+  uint8_t *iv  = new uint8_t[data.size()];
   size_t index = 0;
-  size_t num_index = -1;
-  size_t num_p = 0;
-  uint8_t overflow = 0;
+
   for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
     b[index] = (*i>>bit_length) & 1;
-    if(b[index] == 1) {
-        uint64_t orig = original[num_p];
-        int diff = index-num_index;
-        int shiftsize = bsize-(diff*bit_length);
-        orig = (orig<<shiftsize)>>shiftsize;
-        num_index++;
-        if(num_index%2 == 1) {
-          if(diff == 16) {
-            overflow = orig>>60;
-          }
-          orig = orig<<4;
-        }
-        uint64_t *x = (uint64_t *)&iv[num_index/2];
-        *x = *x | orig;
[…]

### 2020-03-20 bris4_subarray.cpp
Subarray codes

@@ -1,0 +1,147 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 4;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[16];
+
+  std::array<std::vector<uint8_t>, 16> iv;
+  size_t index = 0;
+  for(int i = 0; i < 16; i++) {
+    vector<uint8_t> vec(data[i].size()/2 +1,0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+      uint8_t value = *j;
+      if(index%2 == 0) {
+        vec[index/2] = (value%cap)<<4;
+      }
+      else {
+        vec[index/2] += value%cap;
+
+      }
+//      cout << (int)*j <<" " << (int)value << " " << (int)vec[index/2] << endl;
+      bv[index] = (*j>>bit_length) & 1;
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
[…]

### 2020-03-20 bris4v5_subarray.cpp

@@ -1,0 +1,147 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 4;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[16];
+
+  std::array<std::vector<uint8_t>, 16> iv;
+  size_t index = 0;
+  for(int i = 0; i < 16; i++) {
+    vector<uint8_t> vec(data[i].size()/2 +1,0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+      uint8_t value = *j;
+      if(index%2 == 0) {
+        vec[index/2] = (value%cap)<<4;
+      }
+      else {
+        vec[index/2] += value%cap;
+
+      }
+//      cout << (int)*j <<" " << (int)value << " " << (int)vec[index/2] << endl;
+      bv[index] = (*j>>bit_length) & 1;
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
[…]

### 2020-03-20 bris8_subarray.cpp

@@ -1,0 +1,133 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[8];
+  std::array<std::vector<uint8_t>, 8> iv;
+  size_t index = 0;
+
+  for(int i = 0; i < 8; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  select_support_mcl<> sls[8];
+  rank_support_v<0> rb[8];
+  for(int i =0; i < 8; i++) {
+    select_support_mcl<> s(&b[i]);
+    rank_support_v<0> r(&b[i]);
+    sls[i] = s;
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
[…]

### 2020-03-20 bris8v5_subarray.cpp

@@ -1,0 +1,130 @@
+#include <sdsl/bit_vectors.hpp>
+#include <sdsl/int_vector.hpp>
+#include <string>
+#include <vector>
+#include <ctime>
+#include <chrono>
+#include <iterator>
+#include <iostream>
+#include "vbyte_helpers.hpp"
+#include <valgrind/callgrind.h>
+
+using namespace std;
+using namespace sdsl;
+
+uint32_t bit_length = 8;
+unsigned int random_accesses = 1000000;
+uint32_t cap = 1<<bit_length;
+int main(int argc, char *argv[]) {
+  if(argc<=1) {
+    cerr << "Give uint64_t file as parameter" << endl;
+    return 1;
+  }
+
+  vector<uint64_t> original = read_data_from_file(argv[1]);
+
+  std::array<std::vector<uint32_t>, 8> data;
+  for(int i = 0; i < 8; i++) {
+    vector<uint32_t> vec;
+    data[i] = vec;
+  }
+  //uint8_t *iv  = new uint8_t[8][original.size()];
+  int data_size = 0;
+  for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
+    vector<uint32_t> v = vb_encode_number(*i, cap);
+    v.front() += cap;
+    int level = 0;
+    for(auto j = v.rbegin(); j != v.rend(); j++) {
+      data[level].push_back(*j);
+      level++;
+      data_size++;
+    }
+  //  data.insert(data.end(), v.begin(), v.end());
+  }
+  bit_vector b[8];
+  std::array<std::vector<uint8_t>, 8> iv;
+  size_t index = 0;
+
+  for(int i = 0; i < 8; i++) {
+    vector<uint8_t> vec(data[i].size(),0);
+    bit_vector bv(data[i].size(), 0);
+    index = 0;
+    for (vector<uint32_t>::const_iterator j = data[i].begin(); j != data[i].end(); j++, index++) {
+        bv[index] = (*j>>bit_length) & 1;
+        vec[index] = *j%(cap);
+    }
+
+    iv[i] = vec;
+    b[i] = bv;
+  }
+
+
+  rank_support_v5<0> rb[8];
+  for(int i =0; i < 8; i++) {
+    rank_support_v5<0> r(&b[i]);
+    rb[i] = r;
+  }
+  //select_support_mcl<> sls(&b);
+
+//  cout << "continue-stop vector memory size: " << size_in_bytes(b) + sizeof(bit_vector) << "bytes" <<endl;
[…]

### 2020-03-20 generate.cpp

@@ -21,11 +21,11 @@ int main(int argc, char *argv[]) {
   filename.append(argv[1]);
   filename.append("_");
   filename.append(argv[2]);
-  // int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<31}; // <- all
+   int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<30}; // <- all
   // int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- twolarge
   // int maxvalues [8] = {1<<3,1<<4,1<<5,1<<6,1<<7,1<<8,1<<16,1<<31}; // <- vsmall
   //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- onelarge
-  int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
+  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
     int rand = dis(gen)%8;

### 2020-03-20 select4.cpp

@@ -25,18 +25,8 @@ int main(int argc, char *argv[]) {
   }
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
-//  uint8_t *siv  = new uint8_t[8];
-//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
-//  uint64_t *x = (uint64_t *)&siv[0];
-//  *x = original[0]<<4;
-//  cout << *x << endl;
-//  cout << (int)siv[0] << " " << (int)siv[1] << " " << (int)siv[2] << " " << (int)siv[3] << endl;
-//  exit(1);
 
   vector<uint32_t> data;
-
-//  REDO THIS:
-//  calculate bits on the fly and save number to block array intact
 
   for(vector<uint64_t>::const_iterator i = original.begin(); i != original.end(); i++) {
     vector<uint32_t> v = vb_encode_number(*i, cap);
@@ -115,15 +105,12 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
   for (vector<unsigned int>::const_iterator i = indices.begin(); i != indices.end(); i++) {
     index = *i;
 
+
     begin = index == 0 ? 0 : sls(index)+1;
-//    while(*(b.begin()+begin+diff)==0) diff++;
-//    bit_vector::iterator iter = b.begin()+begin;
-//    while(*iter == 0) {
-//      iter++;
-//    }
-//    uint8_t diff = std::distance(b.begin()+begin, iter);
-// TODO: try out with builtin_clz
-//    test = (uint64_t *)&iv[begin];
+
+
+
+
     int offset = begin%bsize;
     int block = begin/bsize;
     uint64_t blokki = *(b.data()+block);
@@ -146,11 +133,6 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
     if(0 && val != original[index]) {
       cout << "Did not match: " << val << " vs "  << original[index]   << endl;
       cout << "index " << index << ", begin " << begin << endl;
-      cout << bitset<64>(val) << endl;
-      cout << bitset<64>(original[index]) << endl;
-      cout << bitset<64>(original[index+1]) << endl;
-
-      exit(1);
     }
   }
 

### 2020-03-20 select4_subarray.cpp

@@ -13,9 +13,11 @@
 using namespace std;
 using namespace sdsl;
 
-uint32_t bit_length = 8;
+uint32_t bit_length = 4;
 unsigned int random_accesses = 1000000;
 uint32_t cap = 1<<bit_length;
+int bsize = 64;
+
 int main(int argc, char *argv[]) {
   if(argc<=1) {
     cerr << "Give uint64_t file as parameter" << endl;
@@ -32,13 +34,35 @@ int main(int argc, char *argv[]) {
     data.insert(data.end(), v.begin(), v.end());
   }
 
-  bit_vector b(data.size(), 0);
-  uint8_t *iv  = new uint8_t[data.size()];
-  size_t index = 0;
 
+  bit_vector b(data.size(), 0);
+  uint8_t *iv  = new uint8_t[data.size()/2+4];
+  size_t index = 0;
+  size_t num_index = -1;
+  size_t num_p = 0;
+  uint8_t overflow = 0;
   for (vector<uint32_t>::const_iterator i = data.begin(); i != data.end(); i++, index++) {
     b[index] = (*i>>bit_length) & 1;
-    iv[index] = *i;
+    if(b[index] == 1) {
+        uint64_t orig = original[num_p];
+        int diff = index-num_index;
+        int shiftsize = bsize-(diff*bit_length);
+        orig = (orig<<shiftsize)>>shiftsize;
+        num_index++;
+        if(num_index%2 == 1) {
+          if(diff == 16) {
+            overflow = orig>>60;
+          }
+          orig = orig<<4;
+        }
+        uint64_t *x = (uint64_t *)&iv[num_index/2];
+        *x = *x | orig;
+        if(num_index%2 == 1 && diff == 16) {
+            iv[8+(num_index/2)] = overflow;
+        }
+        num_index=index;
+        num_p++;
+    }
   }
 
   bit_vector::select_1_type sls(&b);
@@ -64,16 +88,14 @@ uint64_t *test;
 int end;
 uint8_t diff;
 uint64_t val;
-int bsize = 64;
-cout << "bsize" << bsize << endl;
 int shiftamt;
 uint64_t maxuint = 0-1;
 cout << maxuint << endl;
-uint64_t bitmasks[8];
+uint64_t bitmasks[16];
 val = 0;
-for(int i = 0;i < 8; i++) {
-  val = val << 8;
-  val = val | 0xFF;
[…]

### 2020-03-20 bris4_subarray.cpp
remove fact check

@@ -123,7 +123,7 @@ int main(int argc, char *argv[]) {
       val = (val<<bit_length) + e;
       z = z^val;
 
-      if(1 && val != original[*i+counter]) {
+      if(0 && val != original[*i+counter]) {
         cout << "Did not match: " << val << " vs " << original[*i+counter] << endl;
         cout << original[*i+counter]%cap << " " << original[*i+counter]/cap << endl;
       }

### 2020-03-20 bris8_subarray.cpp

@@ -109,7 +109,7 @@ int main(int argc, char *argv[]) {
       val = (val<<bit_length) + iv[level][index];
       z = z^val;
 
-      if(1 && val != original[*i+counter]) {
+      if(0 && val != original[*i+counter]) {
         cout << "Did not match: " << val << " vs " << original[*i+counter] << endl;
         cout << original[*i+counter]%cap << " " << original[*i+counter]/cap << endl;
       }

### 2020-03-20 bris8v5_subarray.cpp

@@ -106,7 +106,7 @@ int main(int argc, char *argv[]) {
       val = (val<<bit_length) + iv[level][index];
       z = z^val;
 
-      if(1 && val != original[*i+counter]) {
+      if(0 && val != original[*i+counter]) {
         cout << "Did not match: " << val << " vs " << original[*i+counter] << endl;
         cout << original[*i+counter]%cap << " " << original[*i+counter]/cap << endl;
       }

### 2020-03-20 select4_subarray.cpp

@@ -138,7 +138,7 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
 //      if((int)diff != sls(index+counter+1) - sls(index+counter) -1) {
 //        cout << "Diff:" << (int)diff << " sls diff:" << (sls(index+counter+1) - sls(index+counter) -1) << endl;
 //      }
-      if(1 && val != original[index+counter]) {
+      if(0 && val != original[index+counter]) {
         cout << "diff: " << (int)diff << endl;
         cout << "Did not match: " << val << " vs "  << original[index+counter]   << endl;
         cout << "index " << index+counter << ", counter: " << counter << endl;

### 2020-03-20 select8_subarray.cpp

@@ -114,7 +114,7 @@ chrono::steady_clock::time_point time_begin = chrono::steady_clock::now();
 //      if((int)diff != sls(index+counter+1) - sls(index+counter) -1) {
 //        cout << "Diff:" << (int)diff << " sls diff:" << (sls(index+counter+1) - sls(index+counter) -1) << endl;
 //      }
-      if(1 && val != original[index+counter]) {
+      if(0 && val != original[index+counter]) {
         cout << "Did not match: " << val << " vs "  << original[index+counter]   << endl;
         cout << "index " << index+counter << ", counter: " << counter << endl;
         cout << "bit_offset " << bit_offset << ", total_diff " << (int)total_diff << endl;

### 2020-03-22 gradu/msc_english_contents.tex
subarray results

@@
 
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    xlabel={average integer byte length},
+    ylabel={ms per million queries},
+    xmin=1, xmax=8,
+    ymin=0, ymax=4000,
+    xtick={1,2,3,4,5,6,7,8},
+    ytick={0,500,1000,1500,2000,2500,3000,3500,4000},
+    legend pos=north west,
+    ymajorgrids=true,
+    grid style=dashed,
+]
+
+\addplot[
+color=blue,
+mark=square,
+]
+coordinates {
+(1,394.90)(2,654.60)(3,976.20)(4,1282.50)(5,1711.70)(6,2256.70)(7,2954.30)(8,3326.50)
+};
+\addplot[
+color=red,
+mark=ball,
+]
+coordinates {
+(1,776.70)(2,1484.30)(3,2246.60)(4,2570.30)(5,3476.20)(6,4144.80)(7,4760.40)(8,6771.50)
+};
+\addplot[
+color=green,
+mark=triangle,
+]
+coordinates {
+(1,393.70)(2,732.00)(3,1087.70)(4,1445.50)(5,1893.60)(6,2423.50)(7,3133.10)(8,3612.80)
+};
+\addplot[
+color=black,
+mark=diamond,
+]
+coordinates {
+(1,804.10)(2,1711.80)(3,2640.60)(4,3080.70)(5,4126.30)(6,5022.10)(7,5788.50)(8,8258.20)
+};
+\addplot[
+color=orange,
+mark=square*,
+]
+coordinates {
+(1,422.00)(2,508.40)(3,547.90)(4,636.30)(5,645.60)(6,664.50)(7,683.70)(8,703.30)
+};
+\addplot[
+color=blue,
+mark=star,
+]
+coordinates {
+(1,569.40)(2,694.80)(3,676.80)(4,691.70)(5,708.30)(6,716.50)(7,780.10)(8,812.20)
+};
+\legend{bris8, bris4, bris8v5, bris4v5, select8, select4 }
+
+TODO: create a dataset with low numbers and some large ones (every 40th or so)
+
+ - compare and explain
+
+\end{axis}
+\end{tikzpicture}
+\caption{Rank and select performances with subarray fetching} \label{graph:subarray}
+\end{figure}
+
  - TODO figure out why 4bit crashes sometimes with 500M\_all

### 2020-03-22 bris4_subarray.cpp

@@ -23,8 +23,8 @@ int main(int argc, char *argv[]) {
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
 
-  std::array<std::vector<uint32_t>, 8> data;
-  for(int i = 0; i < 8; i++) {
+  std::array<std::vector<uint32_t>, 16> data;
+  for(int i = 0; i < 16; i++) {
     vector<uint32_t> vec;
     data[i] = vec;
   }
@@ -123,7 +123,7 @@ int main(int argc, char *argv[]) {
       val = (val<<bit_length) + e;
       z = z^val;
 
-      if(0 && val != original[*i+counter]) {
+      if(0   && val != original[*i+counter]) {
         cout << "Did not match: " << val << " vs " << original[*i+counter] << endl;
         cout << original[*i+counter]%cap << " " << original[*i+counter]/cap << endl;
       }

### 2020-03-22 bris4v5_subarray.cpp

@@ -23,8 +23,8 @@ int main(int argc, char *argv[]) {
 
   vector<uint64_t> original = read_data_from_file(argv[1]);
 
-  std::array<std::vector<uint32_t>, 8> data;
-  for(int i = 0; i < 8; i++) {
+  std::array<std::vector<uint32_t>, 16> data;
+  for(int i = 0; i < 16; i++) {
     vector<uint32_t> vec;
     data[i] = vec;
   }

### 2020-03-22 bris8_subarray.cpp

@@ -59,12 +59,9 @@ int main(int argc, char *argv[]) {
   }
 
 
-  select_support_mcl<> sls[8];
   rank_support_v<0> rb[8];
   for(int i =0; i < 8; i++) {
-    select_support_mcl<> s(&b[i]);
     rank_support_v<0> r(&b[i]);
-    sls[i] = s;
     rb[i] = r;
   }
   //select_support_mcl<> sls(&b);

### 2020-04-01 gradu/msc_english_contents.tex
updates to gradu

--- 2020-03-19 gradu/msc_english_contents.tex
@@
 
-The most intuitive way to calculate block length of i-th element is from $select_1$(i+1) and subtract the previously calculated start block index. This however causes a second 
-$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until the next 1. Alternatively, the block length can be calculated by reading an integer 
-from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
+The most intuitive way to calculate block length of i-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
+$select_1$ query, which is costly. A much faster option is to calculate the block length by reading a word from the bit array and bit-shifting the word until the bit representing the start of the i-th
+element is on the rightmost end and then counting the trailing zeros.
+This counts the number of blocks in the i-th element that do not have the continuation bit set.
+The number is incremented by one to
+count in the ending block with the continuation bit.
+
+Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
 Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
@@
 
-\chapter{Sub-array access}
-After locating and decompressing one value from a compressed data structure, compressing the subsequent items gets a lot easier.
-Because the next compressed value is usually located next to the fetched
-in memory, locating the data is not needed. This is especially beneficial to the VByte decoding with $select$ method, where calculating the start byte location takes majority of he runtime. In 
-addition to this, the length of the next compressed integer is faster to calculate because already manipulated data in memory.
+\chapter{Subarray access}
 
-The shown $rank$ method by Brisaboa et al.
-can also effectively decompress the next value but some modifications are needed.
-When fetching a longer subarray, the goal becomes to minimize $rank$ calls.
-If an additional pointer array is stored during subarray decompression, only one $rank$ call for each accessed block level is needed.
-This can be observed from Figure~\ref{bris_ds}; the blocks of a subarray
-are located next to each other in different block levels.
+After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements is significantly easier. Calculating the memory location of the next element is not 
+needed because the next compressed value is usually located next to the previously fetched element.
+This is especially beneficial to the VByte decoding with $select$ method, where calculating the start
+byte location takes majority of the runtime. In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory.
+
+Single direct access with $select$ needs to calculate the length of the wanted element. This is done by reading a 64-bit word from the continuation bit array, bit-shifting until the bit representing 
+the first block is on the rightmost end and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
+and then counting trailing zeros. Once the word is shifted through, another word is read and process is continued.
+
+The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
+When a new block level is reached first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored. Next time the same level is reached, the
+previous stored is incremented by one and that value is used instead of the $rank_1$ query.
+The amount of $rank_1$ calls needed is equal to the number of $rank_1$ calls needed to decompress the largest
+number in the subarray.
 
@@
 
- - TODO get results from this
 \chapter{Experimental results}
@@
 
-Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
+\section{VB decoding comparison}
+Six different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
 library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers.
@@
 \begin{itemize}
+  \item Bris4 - Bris implementation with block size 4, using rank support v
   \item Bris4v5 - Bris implementation with block size 4, using rank support v5
@@
 
-In rank, the v5 version naturally has worse performance because more calculations are needed to locate the correct index. As a trade-off, it requires 
+The v5 version of rank naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
 significantly less memory, as seen on Figure~\ref{table:supportsize}.
@@
     grid style=dashed,
+    width=13cm, height=8cm
 ]
@@
 
-Figures \ref{table:results8bit} and \ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures. The data sets consist of 50 million 64-bit numbers and 
-the compressed size is a sum of continuation bit array and the data blocks.
+\section{Memory usage}
+Table~\ref{table:results8bit} and Table~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
+ of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks.
 
@@
 
- - TODO figure out why 4bit crashes sometimes with 500M\_all
+\section{Subarray access results}
 
- - see how cache matters
+Figure~\ref{graph:subarray} shows results for subarray fetching.
+The data sets used contained mostly 4-bit integers and a few 32-bit integers randomly among them.
+The number in the x-axis corresponds to the
+number of large integers within 1000 integers and the y-axis has the average time of fetching 50 subsequent numbers from random location.
 
- - better dataset might be something with larger (like 60bit) numbers
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    xlabel={number of 32bit integers per 1000 integers},
+    ylabel={ms per million subarray queries},
+    xmin=1, xmax=16,
+    ymin=400, ymax=600,
+    xtick={1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
+    ytick={300,400,500,600,700,800,900,100},
+    legend pos=north west,
+    ymajorgrids=true,
+    grid style=dashed,
+    width=13cm, height=8cm
+]
 
- - subarray access
+\addplot[
+color=blue,
+mark=square,
+]
+coordinates {
+(1,405.80)(2,416.10)(3,426.10)(4,436.20)(5,444.40)(6,455.20)(8,472.50)(10,490.10)(12,503.50)(14,523.90)(16,539.60)(18,553.90)(20,564.80)(25,596.60)
+};
+\addplot[
+color=blue, dashed,
+mark=square,
+]
+coordinates {
+(1,407.30)(2,422.40)(3,436.70)(4,451.30)(5,466.20)(6,481.80)(8,513.50)(10,553.10)(12,580.40)(14,616.30)(16,653.80)(18,685.80)(20,713.40)(25,810.70)
+};
+\addplot[
+color=red,
+mark=triangle,
+]
+coordinates {
+(1,406.20)(2,418.50)(3,429.40)(4,440.60)(5,451.40)(6,461.20)(8,480.90)(10,499.50)(12,520.50)(14,534.60)(16,549.00)(18,570.10)(20,591.00)(25,625.20)
+};
+\addplot[
+color=red, dashed,
+mark=triangle, 
+]
+coordinates {
+(1,423.50)(2,442.20)(3,461.20)(4,479.70)(5,497.50)(6,519.30)(8,559.40)(10,601.10)(12,642.60)(14,688.50)(16,728.40)(18,769.20)(20,793.00)(25,893.50)
+};
+\addplot[
+color=black,
+mark=ball,
+]
+coordinates {
+(1,427.00)(2,428.10)(3,428.70)(4,429.40)(5,430.00)(6,430.50)(8,432.20)(10,433.50)(12,434.30)(14,433.80)(16,435.70)(18,436.90)(20,437.90)(25,440.80)
+};
+\addplot[
+color=black, dashed,
+mark=ball, 
+]
+coordinates {
+(1,461.20)(2,461.00)(3,462.70)(4,463.70)(5,465.20)(6,464.80)(8,466.40)(10,468.40)(12,467.90)(14,469.60)(16,472.60)(18,472.70)(20,475.80)(25,480.60)
+};
 
- - bris has advantage of different size blocks (in different levels)
+\legend{bris8, bris4, bris8v5, bris4v5, select8, select4 }
 
- - figure out why data size matters
-
- - do we need to support search()?
-
- - $rank$ vs rank? should formulas have a specific style?
+\end{axis}
+\end{tikzpicture}
+\caption{Rank and select performances with subarray fetching} \label{graph:subarray}
+\end{figure}
 
@@
 The data needs
-less manipulation because the data is already written as it is, but calculating the integer block lengths might benefit from SIMD.
+less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
 
@@
 It is good to conclude with a summary of findings. You can also use separate chapter for discussion and future work. These details you can negotiate with your supervisor.
+
+TODO:
+
+ - see how cache matters
+
+ - bris has advantage of different size blocks (in different levels)
+
+ - figure out why data size matters
+
+ - do we need to support search()?
+
+ - $rank$ vs rank? should formulas have a specific style?

### 2020-04-01 bris4.cpp

@@ -55,7 +55,6 @@ int main(int argc, char *argv[]) {
       }
       else {
         vec[index/2] += value%cap;
-
       }
 //      cout << (int)*j <<" " << (int)value << " " << (int)vec[index/2] << endl;
       bv[index] = (*j>>bit_length) & 1;

### 2020-04-12 gradu/msc_english_contents.tex
Updated pdf

@@
 \chapter{Introduction}
-Enormous datasets are a common case in today's applications. Compressing the datasets is beneficial, because they 
-naturally decrease memory requirements but also are faster when compressed data is read from disk \citep{Zob95}.
+Enormous datasets are a common case in today's applications. Compressing these datasets is often beneficial, because doing so 
+naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95}.
 
-One of the leading methods of data compression is variable-length coding \citep{Sal99}, where frequent sequences of data
-are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
+Fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
+frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
 not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
-data compression algorithms are inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill. However, 
-random access of compressed data is useful in compressed data structures. It saves storage space, bandwidth and increases the likelihood
-of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data from the beginning.
+data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
+need to fulfill. However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
+increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
+from the beginning.
 
-A variable-byte encoding based integer compression method with constant time random access was first introduced by \citep{Bri09}. They used clever block 
-reorganizing and $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
+A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
+reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
 has been widely adopted.
 
-An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ implementation by Brisaboa et al is 
-shown with different implementations of $rank$ and $select$ and several different data sets. Proposed method does not use data 
+In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. Comparison to the $rank$-based 
+implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$ and several different data sets. The proposed method does not use data 
 reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
@@
 
-$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better. Additionally, the
-proposed $select$ method offers quick access to next or previous element.
+The $rank$ method is found superior when dataset contains mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based 
+method performs better. Additionally, the proposed $select$ method offers fast access to the next or previous element.
 
-\chapter{Variable-bytes}
+\chapter{Variable-byte encoding of integers}
 
@@
 width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
-is generally faster to decode \citep{Bri09} and sometimes preferred due to its simplicity.
+is generally faster to decode \citep{Bri09} and so is sometimes preferred, also  due to its simplicity of implementation.
 
@@
 words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
-advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number 
-instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the integers are small. 
-In contrast, rare words have only a few entries but the integers stored are larger. These lists are excellent data sets for VB encoding.
+advanced search purposes. Usually these lists are preprocessed to an inverted index, storing each number as its difference to the previous number 
+instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but inverted index encoding makes the integers are small. 
+In contrast, rare words have only a few entries but the integers stored are larger. These lists are thus excellent candidates for VB encoding.
 
-Variable-byte encoding originates from and is used in MIDI music file format \citep{Mid96} and several applications have a similar implementation of VB. Apache 
-Lucene has vInt datatype. Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
- Microsoft .Net framework offers "7BitEncodedInt" in BinaryReader and BinaryWriter classes and IBM DB2 has a variable byte  \citep{Bha09}.
+Variable-byte encoding originates from and is used in the MIDI music file format \citep{Mid96} and several applications have a similar implementation of VB. Apache 
+Lucene has the vInt datatype \citep{ApaLV}. The Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
+ Microsoft .Net framework offers "7BitEncodedInt" in BinaryReader and BinaryWriter classes and IBM DB2 uses variable byte encoding to store record identifier lists \citep{Bha09}.
 
-Later, VB encoding was to 
-be found efficient in compressing integer lists. It was first experimented as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. 
-That yielded excellent records, and since then many different approaches have been introduced.
+Later, VB encoding was found to be efficient for compressing integer lists. It was first experimented as a tool for compressing inverted index lists of word locations in documents by \citep{Sch02}. 
+In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced.
 
-More recent studies have taken a look into the machine code level for VB and applied SIMD (Single instruction, multiple data) instructions to VB \citep{Lem18,Pla15}. The bit 
+More recent studies have taken a look into the machine code level for VB and applied SIMD (Single instruction, multiple data) instructions for VB decoding \citep{Lem18,Pla15}. The bit 
 operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
 
-Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. Their encoding process assigns short bit array codes to small integers, 
-which is why they outperform VB encoding on datasets with a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, the compressable data sets will have
-the focus on slightly larger, unordered numbers.
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the aforementioned data sets. Their encoding process assigns short bit sequences to small integers, 
+which is why they outperform in trems of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, 
+the data sets used for experimentation will have the focus on slightly larger, unordered numbers.
 
@@
 VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The continuation bit is set to 1 only
-on the block containing the least significant bits of the integer. This information is used in decoding to signal if the current integer continues in the next chunk.
+on the block containing the least significant bits of the integer. This information is used in decoding to signal whether or not the current integer continues in the next chunk.
 
 For example, the standard bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. When split to blocks of length 4, it becomes \texttt{0000 0000 0010 1010}. The empty blocks 
-are removed and then continuation bits are added.
-1 is added to the least significant block and 0 to the other block, resulting in \texttt{00010 11010}.
-A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}.
-Prepend adds an element to the beginning of the list
-and extend adds all the elements of the second list to the end of the first list.
+are removed and then continuation bits are added. 1 is prepended to the least significant block and 0 to the other blocks, resulting in \texttt{00010 11010}. A pseudo code for VB encoding with 
+block length 4 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list 
+and \textsc{extend} adds all the elements of the second list to the end of the first list.
 
@@
 
-Smaller block length can yield a better compression rate at the cost of more bit manipulation, while bigger block lenghts need less bit manipulation and 
+Smaller block lengths can yield a better compression rate at the cost of more bit manipulation and therefore possibly slower decompression, while bigger block lenghts need less bit manipulation and 
 offer less effective compression. On the other hand, bigger block length means a smaller percentage of added continuation bits.
-Generally block length of 7 has been used because it tends to perform well on average and handling chunks as bytes is 
+Generally, a block length of 7 has been popular because it tends to perform well on average and handling chunks as bytes is 
 convenient \citep{Man08}.
@@
 VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
-are removed from all the chunks and the blocks are concatenated to form the original integer. A pseudo code implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.
-Append adds an element to the end of the list. If block length is changed, additional bit operation steps when reading the data may be needed.
+are removed from all the chunks and the resulting blocks are concatenated to form the original integer.
+A pseudo code implementation of VB decoding with a block length of 7 is shown in Figure~\ref{vbyte_dec}.
+\textsc{Append} adds an element to the end of the list. If the block length is changed, additional bit operation steps when reading the data may be needed.
 
 As how the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
-extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
+extracted and added to $n$, making $n$ = \texttt{10} (bit representation of 2).
 The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
@@
 \section{Rank and Select}
-Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
-$select_1$(B, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep{gbmp2014sea}. To use rank or select in indexing, the
+Rank and select are two operations on a bit array B. $Rank_1$(B, i) gives the number of 1 bits between B[0] and B[i-1], inclusive, and 
+$select_1$(B, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. To use rank or select in indexing, the
 set 1 bits in B should reflect to the element locations in the encoded data.
@@
 For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
-has to be close to the length of the data, because encoded element lenghts vary. VB encoding has several advantages with search and rank: the data is compressed in 
+has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
 blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
@@
 \section{Rank and Select implementation}
+\label{chap:implementations}
 
 The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
- of both rank and select to support the bit array. Also a few useful functions were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
+ of both rank and select to support the bit array. Also a few useful functions from the library were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
 size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
-the number of 1's in the data.
+the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
+compressed sequence by much less than one bit per element.
+Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$.
+All these were chosen, because they are faste and can be created over a regular bit array.
 
-The data structure of rank has two layers. First layer is the superblock array. For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
-The second layer is the relative count block. It stores a relative count of 1's for every 64th bit. To calculate rank(i), the superblock index is s = i/512 and 
-relative count block index is r = i-s / 64, both being integer divisions. Then number of 1 bits from index r to i are calculated. All three values added together add up to rank(i). 
-Rankv5 is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, but reduces 
-memory requirement to a quarter.
+The data structure to support $rank_v$ has two layers.
+The first layer is the superblock array.
+For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
+The second layer is the relative count block. For every 64th bit inside a superblock, it stores the number of 1's since the start of this superblock. To calculate rank(i), the superblock index is 
+s = i/512 and relative count block index is r = i-s / 64, both being integer divisions. Then number of 1 bits from index r to i are calculated from the original bit array. All three values added 
+together add up to rank(i).
+$Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit.
+This causes the final bit calculation to be more costly,
+but reduces memory requirement to a quarter.
 
-$Select$ data structure works similarly to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
-relative to the superblock. With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until required amount of bits is reached.
+The $rank$ data structures are based on \citep{Vig08}.
+A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks.
+This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits.
+The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock.
+Therefore the remaining 7 relative blocks take 63 bits and fit into one
+64-bit word.
+The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each.
+The first relative block again equals to 0, so the rest
+of the blocks fit to a 64-bit word.
 
-In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only differiating factor is the manual bit count from relative
-block's index to wanted index. This is at most the size of one relative block and thus both of the functions work in constant time \citep{Gon05}.
+The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+relative to the superblock.
+With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached.
+The $select_{MCL}$
+ is a practical variant of a PAT tree \citep{Cla97}.
+
+In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
+block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}.
 
@@
 Rankv5 & \text{6.25\% of bit array}\\
-Select & \text{~8-12\% of bit array}\\
+Select & \text{~6-15\% of bit array (TODO: check this}\\
 \hline
@@
 \section{DAC via rank}
-Directly addressable codes in VB was first introduced by \citep{Bri09}. In their solution, the chunks are sorted in separate arrays by their significance. For each integer, the block 
-of the least significant chunk is stored in the first array $A_1$, and its bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
-stored to $A_2$ and the bit to $B_2$ and so on. After the data is set, rank data structure is built for each bit array. The data structure is implicit and does not require additional space apart
+Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
+of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$.
+Then if the number was stored in multiple chunks, the block of the next chunk is
+stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
 from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.
@@
 
-To decode an element from index i, first block is fetched from $A_1$[i].
-Because each encoded element is composed of at least one block, the first block is obtainable via straight indexing.
-With small integers, this
-is all that is required.
-If the data has a lot of small integers, the $rank$ method has a huge advantage.
-In addition, the block length of this method can change within different array levels.
-This opens
-a possibility for further optimization.
+To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
+With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
+different array levels. This opens a possibility for further compression optimization.
 
-However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, i). $Rank_0$ 
-returns the number of zeros in the bit array preceding index i. In other words, the number of elements before $i$ that continue to the next array. 
+However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
+returns the number of zeros in the bit array preceding index $i$. In other words, the number of elements before $i$ that continue to the next array. 
 This in turn means that $i$ now has the index of the element's next block in array $A_2$.
-This process is repeated until the i-th bit of the bit array of current level is set.
+This process is repeated until the i-th bit of the bit array of the current level is set.
 The resulting integer is constructed from the blocks of the fetched chunks.
@@
 the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
-consists of a few bit operations.
+consists of just a few bit operations.
 
 To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
-only on it's last block, $Select_1$(i) returns the location of the end byte of i-th element. Therefore the start of j-th element in the block array is at 
-block $b_s$ = $select_1$(j-1) + 1. Implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.
+only on its last block, $Select_1$($i$) returns the location of the end byte of $i$-th element. Therefore the start of  $j$-th element in the block array is at 
+block $b_s$ = $select_1$($j$-1) + 1. In this thesis, the implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.
 
@@
 For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$(i) gets the index location
-of needed block. Because the block length differs from byte length, the the two values need to be calculated. First, the starting byte location $s$ needs to be calculated 
-$(s = select_1(i) \times 4) \div 8$, the latter operand being the integer division.
-Second, the first block might not be at the start of the byte $s$.
-This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$
-is needed (in the example case, o equals to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.
+of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
+The first block might not be at the start of the byte $s$.
+This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4).
+Then a 32-bit word $w$ is read from memory from byte location $s$.
 
 At this point, $w$ contains the wanted bits, but also has extra bits from the previous and next compressed integers.
-If offset was not zero, there are trailing $o$ bits in $w$ from the previous integer.
-These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the requested value. Because most systems are
-able to read memory only from byte addresses, there is a corner case where the value is stored in too many bytes for a word.
-In this setup, this happens if compressed length was 8 blocks and offset was 4.
-Then another step of reading another byte and storing the remaining bits to $w$ is required.
-This is entirely avoidable if integer length is constrained.
-In this case maximum integer length would be 28 bits.
+If the offset was not zero, there are $o$ trailing bits in $w$ from the previous integer.
+These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the desire value. Most systems are
+able to read memory only from byte addresses, which leads to a corner case where the compressed integer is stored in more bytes than are contained in a word. In the aforementioned situation, this 
+happens if compressed length was 8 blocks and offset was 4. Then another step of reading another byte and storing the remaining bits to $w$ is required. This is entirely avoidable if the integer 
+length is constrained. In this case the maximum integer length would be 28 bits.
 
-The most intuitive way to calculate block length of i-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
+The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
 $select_1$ query, which is costly. A much faster option is to calculate the block length by reading a word from the bit array and bit-shifting the word until the bit representing the start of the i-th
@@
 This counts the number of blocks in the i-th element that do not have the continuation bit set.
-The number is incremented by one to
-count in the ending block with the continuation bit.
+This number is then incremented by one to
+also count the ending block which has the continuation bit set to 1.
 
 Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
-Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
+Different block sizes need extra calculation to get the block location from the byte array.
+In the example, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns
 a bit mask for k bytes.
@@
 
-Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. Length of the compressed 
-element is obtainable e.g. via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$. Because element length in this case is 3, blocks not belonging to the element
-are removed either with a bitmask or by double bit shifting, resulting in decoded k-th element.
+Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. The length of the compressed 
+element is 3, which is obtainable, e.g., via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$ and blocks not belonging to the element
+are removed either with a bitmask or by double bit shifting, resulting in k-th element being decoded.
 
@@
 \end{tikzpicture}
-\caption{Data} \label{VBarrays}
+\caption{Relation of continuation bits and data blocks in VB encoding} \label{VBarrays}
 \end{figure}
@@
 \section{Elias-Fano encoding}
-Direct access via select can also be implemented with Elias-Fano (TODO citation to Elias-Fano?), but the data has to be sorted. In Elias-Fano encoding, the data is at first split into low and 
-high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bits is trivial. The higher bits are gap-encoded and then encoded in unary. 
+Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted, which is usually a given in inverted index lists of search engines.
+In Elias-Fano encoding \citep{Eli74, Vig13},
+the data is at first split into low and 
+high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
 For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}.
@@
 
-After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements is significantly easier. Calculating the memory location of the next element is not 
-needed because the next compressed value is usually located next to the previously fetched element.
-This is especially beneficial to the VByte decoding with $select$ method, where calculating the start
+After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements becomes significantly easier.
+Calculating the memory location of the next element is not
+needed because the next compressed value is usually located next to the previously fetched element. This is especially beneficial to the select-based VByte decoding, where calculating the start 
 byte location takes majority of the runtime. In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory.
 
-Single direct access with $select$ needs to calculate the length of the wanted element. This is done by reading a 64-bit word from the continuation bit array, bit-shifting until the bit representing 
-the first block is on the rightmost end and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
-and then counting trailing zeros. Once the word is shifted through, another word is read and process is continued.
+A single direct access with $select$ needs to calculate the length of the desired element.
+This is done by reading a 64-bit word from the continuation bit array, bit-shifting until the bit representing
+the first block is on the rightmost end, and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
+and then counting trailing zeros. Once the word is shifted through, another word is read and process continues.
 
 The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
-When a new block level is reached first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored. Next time the same level is reached, the
-previous stored is incremented by one and that value is used instead of the $rank_1$ query.
-The amount of $rank_1$ calls needed is equal to the number of $rank_1$ calls needed to decompress the largest
-number in the subarray.
+When a new block level is reached for the first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored.
+The next time the same level is reached, the
+previously stored result is incremented by one and that value is used instead of issuing another $rank_1$ query. This way the number of $rank_1$ calls needed is equal to the number of $rank_1$ calls 
+needed to decompress the largest number in the subarray.
 
- - examples / visualizaton?
+ - examples / visualizaton
 
@@
 The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
-The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG} and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
+The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
 
@@
 Six different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
-library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers.
+library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers. The $rank$ and $select$ implementations are explained in \autoref{chap:implementations}.
 
 \begin{itemize}
-  \item Bris4 - Bris implementation with block size 4, using rank support v
-  \item Bris4v5 - Bris implementation with block size 4, using rank support v5
-  \item Bris8 - Bris implementation with block size 8, using rank support v
-  \item Bris8v5 - Bris implementation with block size 8, using rank support v5
-  \item Select4 - Proposed implementation with block size 4, using select support mcl
-  \item Select8 - Proposed implementation with block size 8, using select support mcl
+  \item Bris4 - Bris implementation with block size 4, using $rank_v$
+  \item Bris4v5 - Bris implementation with block size 4, using $rank_{v5}$
+  \item Bris8 - Bris implementation with block size 8, using $rank_v$
+  \item Bris8v5 - Bris implementation with block size 8, using $rank_{v5}$
+  \item Select4 - Proposed implementation with block size 4, using $select_{MCL}$
+  \item Select8 - Proposed implementation with block size 8, using $select_{MCL}$
 \end{itemize}
@@
 
-All implementations read the data set from a file and compress it to memory, randomize one million indexes and stored to an array. The times shown in Table~\ref{table:results1} 
+All implementations read the data set from a file and compress it to memory, randomize one million query indexes and store results to an array. The times shown in Table~\ref{table:results1} 
 are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated multiple times and an average of the runtimes is taken.
@@
 \centering
-\caption{Results in milliseconds, smaller is better.\label{table:results1}}
+\caption{Results in milliseconds, smaller is better. Smallest time is bolded
+\label{table:results1}}
 \begin{tabular}{l||c c c c c c c} 
@@
  \hline \hline 
-all (5M)   & 191.80 & 471.50 & 234.00 & 569.40 & 149.20 & 175.90 \\
-all (50M)   & 269.20 & 573.80 & 323.00 & 785.00 & 245.40 & 275.00 \\
-all (500M)   & 298.50 & 677.80 & 368.20 & 941.80 & 355.90 & 382.20 \\
-twolarge (5M)   & 96.70 & 275.50 & 110.70 & 329.00 & 125.60 & 151.90 \\
-twolarge (50M)   & 140.40 & 379.70 & 159.80 & 456.80 & 227.70 & 242.20 \\
-twolarge (500M)   & 168.20 & 441.00 & 191.50 & 584.70 & 330.20 & 341.00 \\
-onelarge (5M)   & 25.80 & 40.10 & 28.30 & 49.20 & 95.90 & 100.40 \\
-onelarge (50M)   & 49.80 & 91.30 & 50.90 & 105.70 & 212.70 & 213.20 \\
-onelarge (500M)   & 55.00 & 100.00 & 58.50 & 124.00 & 315.80 & 307.70 \\
-onlysmall (5M)   & 16.50 & 20.20 & 16.50 & 23.80 & 84.70 & 88.40 \\
-onlysmall (50M)   & 31.20 & 51.80 & 32.10 & 57.00 & 201.20 & 198.80 \\
-onlysmall (500M)   & 34.10 & 56.00 & 35.00 & 63.10 & 297.90 & 293.80 \\
+all (5M)   & 191.80 & 471.50 & 234.00 & 569.40 & \textbf{149.20} & 175.90 \\
+all (50M)   & 269.20 & 573.80 & 323.00 & 785.00 & \textbf{245.40} & 275.00 \\
+all (500M)   & \textbf{298.50} & 677.80 & 368.20 & 941.80 & 355.90 & 382.20 \\
+twolarge (5M)   & \textbf{96.70} & 275.50 & 110.70 & 329.00 & 125.60 & 151.90 \\
+twolarge (50M)   & \textbf{140.40} & 379.70 & 159.80 & 456.80 & 227.70 & 242.20 \\
+twolarge (500M)   & \textbf{168.20} & 441.00 & 191.50 & 584.70 & 330.20 & 341.00 \\
+onelarge (5M)   & \textbf{25.80} & 40.10 & 28.30 & 49.20 & 95.90 & 100.40 \\
+onelarge (50M)   & \textbf{49.80} & 91.30 & 50.90 & 105.70 & 212.70 & 213.20 \\
+onelarge (500M)   & \textbf{55.00} & 100.00 & 58.50 & 124.00 & 315.80 & 307.70 \\
+onlysmall (5M)   & \textbf{16.50} & 20.20 & 16.50 & 23.80 & 84.70 & 88.40 \\
+onlysmall (50M)   & \textbf{31.20} & 51.80 & 32.10 & 57.00 & 201.20 & 198.80 \\
+onlysmall (500M)   & \textbf{34.10}  & 56.00 & 35.00 & 63.10 & 297.90 & 293.80 \\
 
@@
 
-The 8-bit versions tend to perform better than the 4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4bit versions achieve 
+The 8-bit versions tend to perform better than the 4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4-bit versions achieve 
 better compression in data sets containing small integers. Select versions do not depend as much on the block size, because all the blocks in the compressed element are read at the same time. 
@@
 Table~\ref{table:results8bit} and Table~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
- of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks.
+ of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks. For $rank$ and $select$ methods used in this thesis, the total 
+space taken by the extra data structures is only a small fraction of the overall size.
 
@@
 as is the relation between the memory requirement of $rank$ and the total number of blocks. Data structures in versions with smaller block size naturally take more space, but smaller block 
-size allows a better data compression in some data sets and therefore a smaller total size. This can be seen when comparing the two forementioned figures: All 4-bit versions of the algorithms require
+size allows better compression in some data sets and therefore a smaller total size. This can be seen when comparing the two forementioned figures: All 4-bit versions of the algorithms require
 more space than the 8-bit versions, but the compressed size of the data is significantly smaller in the bottom two data sets.
@@
 The number in the x-axis corresponds to the
-number of large integers within 1000 integers and the y-axis has the average time of fetching 50 subsequent numbers from random location.
+number of large integers within 1000 integers and the y-axis has the average time of fetching 50 subsequent numbers from an initially random location.
 
@@
 \end{tikzpicture}
-\caption{Rank and select performances with subarray fetching} \label{graph:subarray}
+\caption{Rank and select performance with subarray fetching} \label{graph:subarray}
 \end{figure}
 
-\chapter{Future work}
+\chapter{Conclusions Future work}
 
-There has already been a few successful attempts to decompress VB encoded integers with SIMD \citep{Lem18,Pla15}.
-It is worthwhile to check if the presented approach benefits from SIMD.
-The data needs
-less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
+It is good to conclude with a summary of findings. You can also use separate chapter for discussion and future work. These details you can negotiate with your supervisor.
+
+There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the select-based approach presented 
+in this thesis benefits from SIMD. The data needs less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
 
@@
 
-\chapter{Conclusions\label{chapter:conclusions}}
-
-It is good to conclude with a summary of findings. You can also use separate chapter for discussion and future work. These details you can negotiate with your supervisor.
-
 TODO:
-
  - see how cache matters
-
- - bris has advantage of different size blocks (in different levels)
 
@@
 
- - do we need to support search()?
+ - add subarray examples / visualization
 
- - $rank$ vs rank? should formulas have a specific style?
+ - write more about subarray access in results
+
+ - more text / references
+
+ - write conclusions
+
+ - write introduction
+
+ - write abstract

### 2020-04-14 gradu/msc_english_contents.tex
Adds subarray examples

@@
 reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
-has been widely adopted.
+has been widely adopted \citep[e.g.][]{Kon17}.
 
@@
 The $rank$ method is found superior when dataset contains mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based 
-method performs better. Additionally, the proposed $select$ method offers fast access to the next or previous element.
+method performs better. Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
+very convenient and fast. A solution for subarray fetching and comparison is proposed in this thesis.
 
@@
 
-A single direct access with $select$ needs to calculate the length of the desired element.
-This is done by reading a 64-bit word from the continuation bit array, bit-shifting until the bit representing
+\section{Subarray access with $select$}
+A single direct access with $select$ needs to calculate the length of the desired element. This is done by reading a word from the continuation bit array, bit-shifting until the bit representing 
 the first block is on the rightmost end, and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
@@
 
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+[node distance=0pt,
+ start chain = A going right,
+ start chain = B going right,
+ arrow/.style = {draw=#1,-{Stealth[]}, 
+                shorten >=1mm, shorten <=1mm}, % styles of arrows
+ arrow/.default = black,
+ X/.style = {rectangle, draw,% styles of nodes in string (chain)
+                minimum width=2ex, minimum height=5ex,
+                outer sep=0pt, on chain},
+ Y/.style = {rectangle, draw=none,% styles of nodes in string (chain)
+                minimum width=0ex, minimum height=5ex,
+                outer sep=0pt, on chain}
+ ]
+
+\node[X, join] (start) {...};
+\node[X] {0};
+\node[X] {1};
+\node[Y] {};
+\node[X] (byte1) {0};
+\node[X] {1};
+\node[X] {0};
+\node[X] {1};
+\node[X] {0};
+\node[X] (select){1};
+\node[X] {0};
+\node[X] {0};
+\node[Y] {};
+\node[X] (byte2) {1};
+\node[X] {1};
+\node[X] {0};
+\node[X] {0};
+\node[X] {1};
+\node[X] {0};
+\node[X] {1};
+\node[X] {0};
+\node[Y] {};
+\node[X] (byte3){0};
+\node[X] {1};
+\node[X] {0};
+\node[X] {1};
+\node[X] {1};
+\node[X] {0};
+\node[X] {1};
+\node[X] {0};
+\node[Y] {};
+\node[X] (byte4){1 };
+\node[X] {...};
+
+\node (index)[above=2ex of select]{k-1th 1};
+\node [below=1ex of byte1]{$B_1$};
+\node [below=1ex of byte2]{$B_2$};
+\node [below=1ex of byte3]{$B_3$};
+\node [below=1ex of byte4]{$B_4$};
+\draw[arrow] (index.south) to (select.north);
+\end{tikzpicture}
+\caption{Continuation bit array, the bits are grouped by the bytes they are in} \label{subarray}
+\end{figure}
+
+Figure~\ref{subarray} shows an example of a continuation bit array. The index of the 1 bit returned by the $select$ query issued for the starting index is marked, and the desired starting index 
+$i$ = $select$($k-1$) +1 is the following bit on the right.
+The byte where $i$-th bit is located can be calculated by doing an integer division $i$/8.
+A 16-bit word $w$ is read from the byte location $B_1$.
+It is beneficial to read as large word as possible, but 16-bits is used in this example for the sake of simplicity.
+Next the starting bit is shifted to the start of the word, the offset being 6 in this case.
+After the operation, $w$ contains bits \texttt{00110010 10000000}, least significant bit being on the left. The length of the current element is $l$ = 1 + \textsc{TrailingZeros}($w$). The number of 
+trailing zeros can be efficiently calculated with GCC function \texttt{\_\_builtin\_ctz} or \texttt{bits::lo} from the 'SDSL-lite' library.
+
+The length of the desired element $l$ and the location $i$ can now be used to decompress desired element.
+For the next element, $i$ is incremented by $l$ and $w$ is bit shifted $l$ bits and the process is
+repeated.
+When $w$ has no 1 bits, another word is read from the following bytes.
+A few extra steps are required, because the old $w$ might have continuation 0 bits left for the current word.
+The number of leftover
+zeros is added to the first length calculation after word switch.
+This number can be calculated either by adding bit shift amounts together and subtracting the result from word-length.
+Another option is to count
+leading zeros of the old $w$ before the first switch. In the example, one zero is left over from the byte $B_2$.
+
+\section{Subarray access with $rank$}
 The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
@@
 
- - examples / visualizaton
+ - examples / visualizaton for rank?
 
@@
 
-\chapter{Conclusions Future work}
+\chapter{Conclusions and Future work}
 

### 2020-04-14 gradu/msc_english_contents.tex
Fixes

@@
 620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
-60201 & \texttt{\underline{1}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
+60201 & \texttt{\underline{0}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
 
@@
 
-\section{Subarray access with $select$}
+\section{Subarray access with select}
 A single direct access with $select$ needs to calculate the length of the desired element. This is done by reading a word from the continuation bit array, bit-shifting until the bit representing 
@@
 
-\section{Subarray access with $rank$}
+\section{Subarray access with rank}
 The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 

### 2020-04-15 gradu/msc_english_contents.tex
Updates

@@
 reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
-has been widely adopted \citep[e.g.][]{Kon17}.
+has been widely adopted \citep[e.g.][]{Kon17, Sha16}.
 
@@
 \section{Rank and Select implementation}
-\label{chap:implementations}
 
@@
 \hline \hline 
-Rank   & \text{25\% of bit array} \\
-Rankv5 & \text{6.25\% of bit array}\\
-Select & \text{~6-15\% of bit array (TODO: check this}\\
+$Rank_v$   & \text{25\% of bit array} \\
+$Rank_{v5}$ & \text{6.25\% of bit array}\\
+$Select$ & \text{~6-15\% of bit array (TODO: check this}\\
 \hline
@@
 \section{DAC via rank}
+\label{chap:implementations}
 Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
@@
 \chapter{DAC with select query}
-
+\label{chap:select_impl}
 Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
@@
 The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
-The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
+The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
+and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
 
 \section{VB decoding comparison}
-Six different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
-library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers. The $rank$ and $select$ implementations are explained in \autoref{chap:implementations}.
-
+Six different VB decoding implementations were compared:
 \begin{itemize}
@@
 
+The Bris-algorithms are implementations based on \citep{Bri09} using two different $rank$ data structures from the SDSL-library \citep{gbmp2014sea}. The $select$-algorithm is the one proposed in 
+this thesis and it is based on the $select_{MCL}$ data structure from the SDSL-library.
+Two different block sizes were used to illustrate how block size affects the runtime and memory requirement.
+Sizes
+8 and 4 were chosen because they are relatively easy to implement.
+
+Each implementation is compressing and dompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
+and \autoref{chap:select_impl}, respectively.
+
+\pagebreak
 Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).
@@
 
-The v5 version of rank naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
-significantly less memory, as seen on Figure~\ref{table:supportsize}.
+The $rank_{v5}$ naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
+significantly less memory than $rank_v$, as seen on Figure~\ref{table:supportsize}.
 
@@
     grid style=dashed,
-    width=13cm, height=8cm
+    width=13cm, height=8.5cm
 ]
@@
 \section{Memory usage}
-Table~\ref{table:results8bit} and Table~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
+Tables~\ref{table:results8bit} and~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
  of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks. For $rank$ and $select$ methods used in this thesis, the total 
@@
 
+The subarray access experiments were run similarly to the previous experiments.
+The data sets used contained 50 million numbers each.
+Most of the numbers were 4-bit long, with a varying amount of 32-bit
+integers randomly among them.
+The numbers in the x-axis correspond to the number of 32-bit integers per 1000 integers.
+One million indexes were randomized and both the index numbers and the data sets were
+preloaded into memory.
+A subarray of length 50 was decompressed from each randomized index location.
+Some precaution was used in the index randomizing to prevent reading from an out of bounds location.
+
 Figure~\ref{graph:subarray} shows results for subarray fetching.
-The data sets used contained mostly 4-bit integers and a few 32-bit integers randomly among them.
-The number in the x-axis corresponds to the
-number of large integers within 1000 integers and the y-axis has the average time of fetching 50 subsequent numbers from an initially random location.
+The probability of having a long integer within the subarray range is visible from the results of the $rank$ based methods.
+When the density of the long integers increase, so does the need to access multiple
+block layers and thus the average time increases.
+The results of the $select$ based methods visualize the static need of singular $select$ query.
+The 4-bit version of $select$ requires a few more instructions
+during the block location phase and therefore is a little bit slower overall.
 
@@
     xtick={1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
-    ytick={300,400,500,600,700,800,900,100},
+    ytick={300,400,500,600,700},
     legend pos=north west,
@@
 \end{tikzpicture}
-\caption{Rank and select performance with subarray fetching} \label{graph:subarray}
+\caption{$Rank$ and $select$ performance with subarray fetching} \label{graph:subarray}
 \end{figure}
@@
 
- - add subarray examples / visualization
-
- - write more about subarray access in results
-
  - more text / references

### 2020-04-16 gradu/msc_english_contents.tex
more text2

@@
 
+Variable-byte encoding is a method for compressing integers of different byte length. An additional $continuous bit$ is added to the data to denote whether the compressed integer
+is continued to the next byte or not. A light-weight data structure is built over the continuous bit array. The fact that the compressed data blocks are of the same size is used
+with the data structure to allow direct access to the compressed variable-byte data.
+
 A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
@@
 
-The $rank$ method is found superior when dataset contains mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based 
-method performs better. Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
+Because of the block reorganizing, the $rank$ does not infact need a $rank$ query when fetching small elements. This makes the $rank$ method superior when dataset contains 
+mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based method performs better.
+
+Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
 very convenient and fast. A solution for subarray fetching and comparison is proposed in this thesis.
@@
 
-It is good to conclude with a summary of findings. You can also use separate chapter for discussion and future work. These details you can negotiate with your supervisor.
+When the dataset contains mostly small integers, the $rank$ based method by Brisaboa et al.
+is superior to the proposed $select$ based method.
+The benefit of using a direct array index instead of a $rank$
+call for the first block gives a huge advantage.
+The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements.
+This can be seen
+from figure~\ref{graph:data_comparison}.
+
+The order of how data is stored with $select$ compression enables an efficient fetching of a subarray.
+Only one $select$ query is needed in the beginning, and the decoding of the elements past the first one
+consists of just a few bit operations.
 
@@
 
- - write conclusions
-
  - write introduction

### 2020-04-17 gradu/gradu.tex
mods
new stuff
Updated pdf
more text

@@
 \usepackage{tikz}             % For making vector graphics (hard to learn but powerful)
+\usetikzlibrary{arrows.meta,scopes,chains,matrix}
 %\usepackage{wrapfig}        % For nice text-wrapping figures (use at own discretion)
@@
 \usepackage[noend]{algpseudocode}
+\usepackage{pgfplots} 
 \singlespacing               %line spacing options; normally use single
@@
 %% Replace parameters with your information.
-\title{something about variable-byte encoding}
+\title{Select-based random access \\ to variable-byte encodings}
 
@@
 \classification{\protect{\ \\
-\  General and reference $\rightarrow$ Document types  $\rightarrow$ Surveys and overviews\  \\
-\  Applied computing  $\rightarrow$ Document management and text processing  $\rightarrow$ Document management $\rightarrow$ Text editing
+\  Information systems $\rightarrow$ Data management systems  $\rightarrow$ Data structures $\rightarrow$ Data layout $\rightarrow$ Data compression \  \\
+\  probably $\rightarrow$ something $\rightarrow$ else $\rightarrow$ too \\
 }}
@@
 \begin{abstract}
-Use this otherlanguage environment to write your abstract in another language if needed.
 
-Topics are classified according to the ACM Computing Classification System
-(CCS), see 
-\url{https://www.acm.org/about-acm/class}:
-check command \verb+\classification{}+. A small set of paths (1--3) should be used, starting from any top nodes
-referred to bu the root term CCS leading to the leaf nodes. The elements
-in the path are separated by right arrow, and emphasis of each element individually can be indicated
-by the use of bold face for high importance or italics for intermediate
-level. The combination of individual boldface terms may give the reader
-additional insight. 
+Enormous datasets are a common occurence today and compressing them is often beneficial.
+Variable-byte encoding is a method for compressing integers of different byte lenghts.
+It removes the zero bytes
+and adds an additional \textit{continuation bit} to each byte to denote whether the compressed integer continues to the next byte or not.
+Two bit array data structures, $rank$ and $select$  are built on top of the array of
+continuation bits.
+These data structures enable a fast direct access to the elements in the compressed data.
+The $rank$ implementation performs better on data containing mostly small integers.
+The $select$
+implementation benefits on larger integers.
+An experiment on fetching a subarray is explained and implemented.
+The $select$ implementation has significant advantages on subarray fetching due to how the data
+is compressed.
+
 \end{abstract}

### 2020-04-17 gradu/msc_english_contents.tex
more text

@@
 
-Variable-byte encoding is a method for compressing integers of different byte length. An additional $continuous bit$ is added to the data to denote whether the compressed integer
+Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuous bit} is added to the data to denote whether the compressed integer
 is continued to the next byte or not. A light-weight data structure is built over the continuous bit array. The fact that the compressed data blocks are of the same size is used
@@
 Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the aforementioned data sets. Their encoding process assigns short bit sequences to small integers, 
-which is why they outperform in trems of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, 
+which is why they outperform in terms of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, 
 the data sets used for experimentation will have the focus on slightly larger, unordered numbers.
@@
 \State $B \gets $ continuation bit array
-\State $begin \gets \Call{Select}{index}$
+\State $begin \gets \Call{Select}{B, index}$
 \State $len \gets \Call{CalculateLength}{begin}$
@@
  - write abstract
+
+ - remove lists from algorithms and data sets in results, explain whe data sets were roughly chosen like that

### 2020-04-17 gradu/msc_english_contents.tex
more references

@@
 Variable-byte encoding originates from and is used in the MIDI music file format \citep{Mid96} and several applications have a similar implementation of VB. Apache 
-Lucene has the vInt datatype \citep{ApaLV}. The Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
+Lucene has the vInt datatype \citep{ApaLV} which works well with short inverted index lists \citep{Wan09}.
+The Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
  Microsoft .Net framework offers "7BitEncodedInt" in BinaryReader and BinaryWriter classes and IBM DB2 uses variable byte encoding to store record identifier lists \citep{Bha09}.
@@
 data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
-structures. It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
+structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.
+
+It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
 list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
@@
 The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
-$select_1$ query, which is costly. A much faster option is to calculate the block length by reading a word from the bit array and bit-shifting the word until the bit representing the start of the i-th
-element is on the rightmost end and then counting the trailing zeros.
+$select_1$ query, which is costly.
+A much faster option is to calculate the block length from the continuation bit array.
+This can be done by reading a word from the bit array and bit-shifting the word
+until the bit representing the start of the i-th element is on the rightmost end and then counting the trailing zeros.
 This counts the number of blocks in the i-th element that do not have the continuation bit set.
-This number is then incremented by one to
-also count the ending block which has the continuation bit set to 1.
+This number is then incremented by one to also count the ending block which has the continuation bit set to 1.
 
@@
 
- - write introduction
+ - write more introduction
 
- - write abstract
+ - write more abstract
+
+ - find ACL classifications
+
+ - write keywords
 

### 2020-04-17 gradu/msc_english_contents.tex
more references

@@
 data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
-need to fulfill. However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
+need to fulfill.
+
+However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
 increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
@@
 \section{VB encoding}
-VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The continuation bit is set to 1 only
-on the block containing the least significant bits of the integer. This information is used in decoding to signal whether or not the current integer continues in the next chunk.
+The name \textit{variable-byte} is reflected in how the integers are stored.
+An array stores all the elements with the same bit length.
+This is very inefficient if a lot of the values are very small but the
+array also has to support very large integers.
+VB encoding attempts to eliminate the unneeded leading zeros.
+The bits in an integer are split into blocks of length $b$ and empty blocks from the beginning are
+discarded.
+Because the lengths of the integers may now be different, the data cannot be stored as it is.
+Instead a \textit{continuation bit} is added in front of each block to form a $chunk$.
+This bit is set
+to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer continues in the next chunk.
 
-For example, the standard bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. When split to blocks of length 4, it becomes \texttt{0000 0000 0010 1010}. The empty blocks 
-are removed and then continuation bits are added. 1 is prepended to the least significant block and 0 to the other blocks, resulting in \texttt{00010 11010}. A pseudo code for VB encoding with 
-block length 4 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list 
-and \textsc{extend} adds all the elements of the second list to the end of the first list.
+For example, the standard 16-bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. Assuming the block length is 4, it is split to \texttt{0000 0000 0010 1010}. The empty blocks 
+are removed from the beginning and then the continuation bits are added. 1 is added in front of the last block (containing the least significant bits) and 0 to the other blocks, resulting in 
+\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with byte length 4.
+
+\begin{table}
+\centering
+\begin{tabular}{l||c c c c c} 
+Original number & first block & second block & third block & fourth block &\\ 
+\hline \hline 
+4  & \texttt{\underline{1}0100}    &                             &                           &  &  \\
+17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
+620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
+60201 & \texttt{\underline{0}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
+
+\hline
+\end{tabular}
+\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
+\end{table}
+
+A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
+\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length.
 
@@
 convenient \citep{Man08}.
-
-\begin{table}
-\centering
-\begin{tabular}{l||c c c c c} 
-Original number & first block & second block & third block & fourth block &\\ 
-\hline \hline 
-4  & \texttt{\underline{1}0100}    &                             &                           &  &  \\
-17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
-620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
-60201 & \texttt{\underline{0}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\
-
-\hline
-\end{tabular}
-\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
-\end{table}
 
@@
 \section{Rank and Select}
-Rank and select are two operations on a bit array B. $Rank_1$(B, i) gives the number of 1 bits between B[0] and B[i-1], inclusive, and 
-$select_1$(B, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. To use rank or select in indexing, the
-set 1 bits in B should reflect to the element locations in the encoded data.
+$Rank$ and $select$ are two bit array operations. $Rank_1$(i) gives the number of bits set to 1 in the bit array before index i. $Select_1$(i) gives the index of i-th bit that is set to 1. 
+Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
+$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.
 
-For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
+To use rank or select in indexing, the set 1 bits in B should reflect to the element locations in the encoded data. For most compression algorithms, this requires B to be 
+created in addition to the existing data and the length of B usually 
 has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
 blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
-as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks up to i-th index and $select_1$(C, i) would give the location of the ending block 
+as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
 of i-th compressed element.
@@
 Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$.
-All these were chosen, because they are faste and can be created over a regular bit array.
+All these were chosen, because they are fast and can be created over a regular bit array.
 
@@
  - remove lists from algorithms and data sets in results, explain whe data sets were roughly chosen like that
+
+ - Chapter 3 start: Direct access to any element.... explain why
+
+ - 3.1 explain more what the bit array is
+
+ - 3.1 faste => fast
+
+ - a real life situation where to use subarray

### 2020-04-24 gradu/gradu.tex
Updated introduction and abstract

@@
 
-Enormous datasets are a common occurence today and compressing them is often beneficial.
-Variable-byte encoding is a method for compressing integers of different byte lenghts.
-It removes the zero bytes
-and adds an additional \textit{continuation bit} to each byte to denote whether the compressed integer continues to the next byte or not.
-Two bit array data structures, $rank$ and $select$  are built on top of the array of
-continuation bits.
-These data structures enable a fast direct access to the elements in the compressed data.
-The $rank$ implementation performs better on data containing mostly small integers.
-The $select$
-implementation benefits on larger integers.
-An experiment on fetching a subarray is explained and implemented.
-The $select$ implementation has significant advantages on subarray fetching due to how the data
-is compressed.
+Enormous datasets are a common occurence today and compressing them is often beneficial. Fast direct access to any element in the compressed data is a requirement in the field of compressed 
+data structures, which is not possible with traditional compression methods.
+
+Variable-byte encoding is a method for compressing integers of different byte lengths. It removes unused leading bytes and adds an additional \textit{continuation bit} to each byte to 
+denote whether the compressed integer continues to the next byte or not. An existing solution using \textsc{rank} data structure performs well in this given task. This thesis introduces an 
+alternative solution using \textsc{select} data structure and compares the two implementations. An experimentation is also done on retrieving a subarray from the compressed data structure.
+
+The $rank$ implementation performs better on data containing mostly small integers. The $select$ implementation benefits on larger integers. The $select$ implementation has significant 
+advantages on subarray fetching due to how the data is compressed.
 

### 2020-04-24 gradu/msc_english_contents.tex
more references
Updated introduction and abstract

@@
 
-Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuous bit} is added to the data to denote whether the compressed integer
-is continued to the next byte or not. A light-weight data structure is built over the continuous bit array. The fact that the compressed data blocks are of the same size is used
-with the data structure to allow direct access to the compressed variable-byte data.
+Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuation bit} is added to the data to denote whether the compressed integer
+is continued to the next byte or not. Two different light-weight data structures are built over the continuation bit array. $Rank_1$($i$) returns the number of 1 bits in the bit array between 
+indexes 0 and $i$-1. $Select_1$($i$) returns the index of the i-th 1 bit in the array. The fact that the compressed data blocks are of the same size is used
+with the data structures to allow direct access to the compressed variable-byte data.
 
 A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
-reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
+reorganizing and a $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
 has been widely adopted \citep[e.g.][]{Kon17, Sha16}.
@@
 
-Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
-very convenient and fast. A solution for subarray fetching and comparison is proposed in this thesis.
+Decompressing a subarray is experimented later in this thesis. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a 
+subarray decompression very convenient and fast. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared. $Select$ based method performs better due to its 
+single $select$ call requirement, while the $rank$ call amount depends on the size of the largest element in the subarray.
 
@@
 The name \textit{variable-byte} is reflected in how the integers are stored.
-An array stores all the elements with the same bit length.
-This is very inefficient if a lot of the values are very small but the
-array also has to support very large integers.
+When stored in an array, each element requires the same amount of space.
+This is very inefficient if a lot of the values are
+very small but the array also has to support very large integers.
 VB encoding attempts to eliminate the unneeded leading zeros.
-The bits in an integer are split into blocks of length $b$ and empty blocks from the beginning are
-discarded.
+The bits in an integer are split into blocks of length $b$ and empty blocks
+from the beginning are discarded.
 Because the lengths of the integers may now be different, the data cannot be stored as it is.
-Instead a \textit{continuation bit} is added in front of each block to form a $chunk$.
-This bit is set
-to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer continues in the next chunk.
+Instead a \textit{continuation bit} is added in front of each block to form
+a $chunk$. This bit is set to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer 
+continues in the next chunk.
 
@@
 \node[X, below=5ex of b1] (a2) {$A_{1,2}$};
-\node [left=1ex of a2]{\textbf{$A_1$ =}};
+\node [left=1ex of a2]{\textbf{$A_2$ =}};
 \node[X] {$A_{3,2}$};
@@
 However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
-returns the number of zeros in the bit array preceding index $i$. In other words, the number of elements before $i$ that continue to the next array. 
-This in turn means that $i$ now has the index of the element's next block in array $A_2$.
-This process is repeated until the i-th bit of the bit array of the current level is set.
-The resulting integer is constructed from the blocks of the fetched chunks.
-A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo}
-with block length of 8.
+returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
+This in turn means that $i$ is the index of desired element's next block in array $A_2$.
+This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current
+block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks.
+
+A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
+contiuation bit arrays $B_k$.
+
+As an example, the fourth element from the data structure shown in Figure~\ref{bris_ds} is decoded. The first block of data is fetched by directly accessing $A_1$[3]. Then the 
+continuation bit $B_1$[3] is checked.
+It is not set, which indicates that the element continues to the next block.
+The index of the next block is obtained via a \textsc{Rank}($B_1$, 3) call.
+There are two zero
+bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is set to 1 so the decompression is completed.
 
@@
 
- - write more introduction
-
- - write more abstract
-
  - find ACL classifications
@@
 
- - Chapter 3 start: Direct access to any element.... explain why
+ - a real life situation where to use subarray
 
- - 3.1 explain more what the bit array is
+ - explain rank/select in introduction
 
- - 3.1 faste => fast
-
- - a real life situation where to use subarray
+ - 1.1 research questions

### 2020-04-25 gradu/gradu.tex
Adds slightly more text

@@
 
-\keywords{ulkoasu, tiivistelmä, lähdeluettelo}
+\keywords{compression, data structure}
 \additionalinformation{\translate{\track}}
@@
 \classification{\protect{\ \\
-\  Information systems $\rightarrow$ Data management systems  $\rightarrow$ Data structures $\rightarrow$ Data layout $\rightarrow$ Data compression \  \\
-\  probably $\rightarrow$ something $\rightarrow$ else $\rightarrow$ too \\
+\  Information systems $\rightarrow$ Data management systems  $\rightarrow$ Data structures $\rightarrow$ Data layout $\rightarrow$ Data compression \  
 }}

### 2020-04-25 gradu/msc_english_contents.tex
Adds slightly more text

@@
 A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
-\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length.
+\textsc{extend} adds all the elements of the second list to the end of the first list.
+The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length.
+Since the block
+length is 7, the continuation bit is added in row 8 by adding $2^8$ = 128 to the last block.
 
@@
   \begin{minipage}{0.5\linewidth}
-\begin{algorithmic}[H]
+\begin{algorithmic}[1]
 \Function{VBEncodeNumber}{$n$}
@@
 
-To use rank or select in indexing, the set 1 bits in B should reflect to the element locations in the encoded data. For most compression algorithms, this requires B to be 
-created in addition to the existing data and the length of B usually 
-has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
+To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect to the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
+created in addition to the existing data.
+The length of the bit array in these cases usually has to be close to the length of the data, because encoded element lengths vary.
+This spatial increment makes
+the compression ineffective and therefore $rank$ and $select$ cannot be applied to any given compression algorithm.
+
+VB encoding has several advantages with $search$ and $rank$: the data is compressed in 
 blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
@@
 
-The data structure to support $rank_v$ has two layers.
-The first layer is the superblock array.
-For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
-The second layer is the relative count block. For every 64th bit inside a superblock, it stores the number of 1's since the start of this superblock. To calculate rank(i), the superblock index is 
-s = i/512 and relative count block index is r = i-s / 64, both being integer divisions. Then number of 1 bits from index r to i are calculated from the original bit array. All three values added 
-together add up to rank(i).
-$Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit.
-This causes the final bit calculation to be more costly,
-but reduces memory requirement to a quarter.
-
-The $rank$ data structures are based on \citep{Vig08}.
-A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks.
-This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits.
-The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock.
-Therefore the remaining 7 relative blocks take 63 bits and fit into one
-64-bit word.
-The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each.
-The first relative block again equals to 0, so the rest
-of the blocks fit to a 64-bit word.
-
-The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
-relative to the superblock.
-With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached.
-The $select_{MCL}$
- is a practical variant of a PAT tree \citep{Cla97}.
-
-In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
-block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}.
-
 \begin{table}
@@
 
+The data structure to support $rank_v$ has two layers.
+The first layer is the superblock array.
+For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
+The second layer is the relative count block. For every 64th bit inside a superblock, it stores the number of 1's since the start of this superblock.
+
+For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, both division operations being integer divisions.
+Then number of 1 bits
+from index $r$ to $i$ are calculated from the original bit array. All three values added 
+together add up to $rank$($i$).
+$Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit.
+This causes the final bit calculation to be more costly,
+but reduces memory requirement to a quarter.
+
+The $rank$ data structures are based on \citep{Vig08}.
+A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks.
+This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits.
+The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock.
+Therefore the remaining 7 relative blocks take 63 bits and fit into one
+64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.
+
+The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each. The first relative block again equals to 0, so the rest 
+of the blocks fit to a 64-bit word. The $rank_{v5}$ implementation requires an additional (64+64)/2048 = 6.25\% space.
+
+The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+relative to the superblock.
+With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached.
+The $select_{MCL}$
+ is a practical variant of a PAT tree \citep{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.
+
+In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
+block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}.
+
 \section{DAC via rank}
@@
 stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
-from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.
+from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
+contiuation bit arrays $B_k$.
 
@@
 With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
-different array levels. This opens a possibility for further compression optimization.
+different array levels which opens a possibility for further compression optimization.
 
@@
 This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current
-block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks.
-
-A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
-contiuation bit arrays $B_k$.
+block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.
 
@@
 In the example, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns
-a bit mask for k bytes.
+a bit mask for k bytes. In line 8, a long word is read from the memory starting at memory address $A$[$begin$].
 
@@
 \centering
-\begin{algorithmic}
+\begin{minipage}{0.5\linewidth}
+\begin{algorithmic}[1]
 \State $i \gets $ wanted index
@@
 \State $mask \gets \Call{ByteMask}{len}$
-\State $word \gets A[begin]$ \Comment{read a word from the byte array}
+\State $word \gets A[begin]$
 \State $word \gets word \mathbin{\&} mask$
@@
 \end{algorithmic}
+\end{minipage}
 \caption{Pseudo code of DAC with select with block length 8} \label{select_pseudo}
@@
 For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}.
-$Select_1$ has an interesting interaction with gap encoded unary arrays.
-$k = Select_1(B,i)$
+
+$Select_1$ has an interesting interaction with gap encoded unary arrays. $k = Select_1(B,i)$ 
 returns the location of i-th 1 in B. Up to index $k$, the bit array has had $i$ ones and $i-k$ zeros. The $i$ corresponds to the number of compressed integers in $B$ and every zero in the array 
@@
 The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
-The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
+The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The implementations were written in C++ and compiled with 
+\texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
 and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.
@@
 
-Each implementation is compressing and dompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
+Each implementation is compressing and decompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
 and \autoref{chap:select_impl}, respectively.
 
-\pagebreak
-Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).
+Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M). The content of the datasets were chosen as such to see how different implementations 
+work with different sized data.
 
@@
 call for the first block gives a huge advantage.
-The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements.
-This can be seen
-from figure~\ref{graph:data_comparison}.
+The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements until $select$
+performs better. This can be seen from figure~\ref{graph:data_comparison}.
 
@@
 TODO:
- - see how cache matters
-
- - figure out why data size matters
 
@@
 
- - find ACL classifications
+ - remove lists from algorithms and data sets in results?
 
- - write keywords
+ - a real life situation of where to use subarray
 
- - remove lists from algorithms and data sets in results, explain whe data sets were roughly chosen like that
-
- - a real life situation where to use subarray
-
- - explain rank/select in introduction
-
- - 1.1 research questions
+ - list research questions somewhere? (introduction 1.1?)

### 2020-04-25 gradu/msc_english_contents.tex
Adds slightly more text

@@
 \section{Rank and Select}
-$Rank$ and $select$ are two bit array operations. $Rank_1$(i) gives the number of bits set to 1 in the bit array before index i. $Select_1$(i) gives the index of i-th bit that is set to 1. 
-Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
+$Rank$ and $select$ are two array operations that are widely used in compressed data structures.
+$Rank_x$($i$) counts the number of occurences of $x$ before index $i$ in the data.
+$Select_x$($i$) returns the index of the i-th
+occurence of $x$ in the data.
+In this thesis, these two operations are assumed to work with a bit array.
+$Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index
+of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
 $B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.

### 2020-04-26 gradu/msc_english_contents.tex
Adds a few updates

@@
 \chapter{Introduction}
-Enormous datasets are a common case in today's applications. Compressing these datasets is often beneficial, because doing so 
-naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95}.
+Enormous datasets are a common case in today's applications. Some datasets are too large to handle in memory and thus must be stored in another, slower storage \citep{Hon10}. 
+Compressing these datasets is often beneficial, because doing so naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95, Wil99}.
 
@@
 
-However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
+However, efficient random access to compressed data is very useful in compressed data structures \citep{Ven13}. It saves storage space, bandwidth and 
 increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
@@
 
+ - TODO: introduce each chapter briefly
+
 \chapter{Variable-byte encoding of integers}
 
-Variable-byte (VB) encoding \citep{Wil99} is a method for compressing unsigned integers via omitting leading zero bits that would be present in a longer fixed 
-width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
-is generally faster to decode \citep{Bri09} and so is sometimes preferred, also  due to its simplicity of implementation.
+Variable-byte (VB) encoding \citep{Mid96, Wil99} is a method for compressing unsigned integers via omitting leading zero bits that would be present in a longer fixed 
+width word.
+In normal data sets such as text files or images, leading zero bits are usually not present in large quantitites and thus VB encoding loses in compression performance to generic algorithms
+like Huffman encoding \citep{Huf52} or Lempel-Ziv encoding \citep{Ziv77}.
 
-A good data set for VB encoding is a list of mostly small integers with a need to support larger ones. A search engine may use an inverted index of 
-words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
-advanced search purposes. Usually these lists are preprocessed to an inverted index, storing each number as its difference to the previous number 
-instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but inverted index encoding makes the integers are small. 
-In contrast, rare words have only a few entries but the integers stored are larger. These lists are thus excellent candidates for VB encoding.
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the integer data sets. Their encoding process assigns short bit sequences to small integers, 
+which is why they outperform in terms of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. VB is generally faster 
+to decode \citep{Wil99, Tro03} and so is sometimes preferred, also due to its simplicity of implementation.
+In this thesis, the data sets used for experimentation will have the focus on slightly larger,
+unordered integers.
 
@@
 
-Later, VB encoding was found to be efficient for compressing integer lists. It was first experimented as a tool for compressing inverted index lists of word locations in documents by \citep{Sch02}. 
-In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced.
+VB encoding was first experimented as a tool for compressing inverted index lists of word locations in documents by \citep{Sch02}. 
+In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced. A search engine may use an inverted index of 
+words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
+advanced search purposes. Usually these lists are preprocessed to an inverted index, storing each number as its difference to the previous number 
+instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but inverted index encoding makes the integers are small. 
+In contrast, rare words have only a few entries but the integers stored are larger.
+These lists are mostly small integers with a need to support larger ones and thus are excellent candidates for VB encoding.
 
@@
 operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.
-
-Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the aforementioned data sets. Their encoding process assigns short bit sequences to small integers, 
-which is why they outperform in terms of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. In this thesis, 
-the data sets used for experimentation will have the focus on slightly larger, unordered numbers.
 
@@
 Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$.
-All these were chosen, because they are fast and can be created over a regular bit array.
+These implementations were chosen because they are fast and can be created over a regular bit array.
 
@@
 In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
-block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}.
+block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}. The 
+implementations are explained in greater detail in \citep{gbmp2014sea}.
 
@@
 \node[X] {$C_{4,2}C_{4,1}$};
-\node[X] {$C_{5,1}$...};
+\node[X] {$C_{5,1}$};
+\node[X] {...};
 
@@
 happens if compressed length was 8 blocks and offset was 4. Then another step of reading another byte and storing the remaining bits to $w$ is required. This is entirely avoidable if the integer 
-length is constrained. In this case the maximum integer length would be 28 bits.
+length is constrained. In this case the maximum integer length of 28 bits would eliminate the handling of the second byte.
 
@@
 Different block sizes need extra calculation to get the block location from the byte array.
-In the example, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns
+In the pseudo code, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns
 a bit mask for k bytes. In line 8, a long word is read from the memory starting at memory address $A$[$begin$].
@@
 \section{Elias-Fano encoding}
-Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted, which is usually a given in inverted index lists of search engines.
+Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted.
+This is usually a given in inverted index lists of search engines.
 In Elias-Fano encoding \citep{Eli74, Vig13},
@@
 between bits is made to reduce the gap sizes.
-This approach can be further optimized by finding clusters in the integers and partitioning the indexes \citep{Ott14}.
+
+Elias-fano encoding is an excellent tool if the elements are sorted and the higher bits are reasonably small, because large integers cause a very inefficient unary encoding.
+This can be further optimized
+by finding clusters in the integers and partitioning the indexes \citep{Ott14}. Elias-Fano encoding is not compared in this thesis, because the restriction of sorted data is relaxed.
 
@@
 \section{Subarray access with rank}
-The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
-When a new block level is reached for the first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored.
-The next time the same level is reached, the
-previously stored result is incremented by one and that value is used instead of issuing another $rank_1$ query. This way the number of $rank_1$ calls needed is equal to the number of $rank_1$ calls 
-needed to decompress the largest number in the subarray.
+The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a subarray, the goal becomes to minimize $rank$ calls. 
+When a new block level is reached for the first time during a subarray query, the index is fetched with $rank$ as normal, and the result of the query is stored.
+When a new index is needed from a previously
+used level, the desired block is known to be right next to the previously fetched one.
+Therefore the stored index value is incremented by one and that value is used instead of issuing another $rank$ query.
+This way the number of $rank$ calls needed is equal to the number of $rank$ calls needed to decompress the largest number in the subarray.
 
- - examples / visualizaton for rank?
+The following example decompresses first four values from Figure~\ref{bris_ds} and assumes the maximum length of integers to be 4 blocks. Array $P$ is used to store the block indices. Initially $P$ = 
+[-1,-1,-1,-1], where -1 denotes that the value is unset and that level has not been reached yet.
+The start of the subarray is index 0.
+$P$[0] is -1, which means the value is unset.
+Because the $rank$ method
+does not need an actual $rank$ call in the first block array, $P$[1] = 0 and that value is used to fetch the first block.
+$B_1$[0] shows that the element is continued to the next block level.
+$P$[1] is also -1,
+so $P$[1] = $rank_0$($B_1$,0) is used to fetch the second block. $B_2$[0] indicates that the element is not continued anymore.
+
+Currently $P$ is [0,0,-1,-1].
+The location of the first block of the next element is fetched by increasing $P$[0] by 1 and then using the incremented value.
+The bit array shows that this element is only
+one block long.
+The first two blocks of the third element is fetched the same way.
+$P$[2] is unset, so $P$[2] = $rank_0$($B_2$,0) is used for the third block, which is the last block of that element.
+After
+these elements, $P$ is [2,1,0,-1].
+The index locations of the fourth element are fetched from $P$ the same way.
+In the end $P$ is [3,2,0,-1] and the process could potentially go on indefinitely with only a
+very small amount of $rank$ calls.
 
@@
 \section{VB decoding comparison}
-Six different VB decoding implementations were compared:
+Three different VB decoding implementations were compared with two different block lengths. Two $rank$ implementations used are based on \citep{Bri09} and use $rank_v$ and $rank_{v5}$ data structures 
+from the SDSL-library \citep{gbmp2014sea}. The different $rank$ data structures illustrate the trade-off between memory and speed. the $select$ algorithm is the one proposed in this thesis and it is 
+implemented using the $select_{MCL}$ data structure from the SDSL-library. Different block lengths were used to illustrate how block size affects the runtime and memory requirement. Lengths 
+8 and 4 were chosen because they are relatively easy to implement.
+
 \begin{itemize}
@@
 \end{itemize}
-
-The Bris-algorithms are implementations based on \citep{Bri09} using two different $rank$ data structures from the SDSL-library \citep{gbmp2014sea}. The $select$-algorithm is the one proposed in 
-this thesis and it is based on the $select_{MCL}$ data structure from the SDSL-library.
-Two different block sizes were used to illustrate how block size affects the runtime and memory requirement.
-Sizes
-8 and 4 were chosen because they are relatively easy to implement.
 
@@
 Tables~\ref{table:results8bit} and~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
- of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks. For $rank$ and $select$ methods used in this thesis, the total 
+ of 50 million 64-bit numbers.
+The compressed size column has a sum of continuation bit array and the data blocks.
+The tables show that the static requirement of $rank_{v5}$ is superior with these datasets.
+If the dataset had even larger integers, $select$ would be dominant. For $rank$ and $select$ methods used in this thesis, the total 
 space taken by the extra data structures is only a small fraction of the overall size.
@@
 The data structure memory usage for the bottom row is similar in both versions
-due to the fact that the continuation bit array is exactly the same.
+due to the fact that the continuation bit array is exactly the same in both cases.
 
@@
  - list research questions somewhere? (introduction 1.1?)
+
+ - should the c++ code be added as appendix?
+
+ - is metatext necessary? i.e. starting each chapter with "This chapter contains..."
+
+ - explain "X is done", "Y is used" in more detail (why is this done / why is this used)
+
+ - Some abbreviations are used. Should there be a list of them in the beginning?
+
+ - is Elias-Fano important or relevant to mention?
+
+ - should terms / practices be opened more (for people who don't know the field)
+
+results:
+
+ - what do the differences in speed or memory usage mean in practice. Is one better than the other?
+
+ - the differences in general need more explaining
+
+7. 
+ - write a recap about what was done

### 2020-04-27 gradu/msc_english_contents.tex
Adds research questions to intro and expalantion to conclusions

@@
 
-In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. Comparison to the $rank$-based 
-implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$ and several different data sets. The proposed method does not use data 
+In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. The proposed method does not use data 
 reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
@@
 
-Because of the block reorganizing, the $rank$ does not infact need a $rank$ query when fetching small elements. This makes the $rank$ method superior when dataset contains 
-mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based method performs better.
+Comparison of the proposed method and $rank$-based implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$. The $rank$ method 
+is assumed to perform better on data sets containing mostly small integers.
+It is assumed that larger numbers cause problems in performance.
+The performance of the proposed $select$ method is indifferent
+of the size of the elements in the data.
+Different kinds of data sets are used to see how each algorithm performs.
+The assumption is that $select$ works better with data sets with very large integers.
+It
+remains to be seen how large the integers need to be for the $select$ method to outperform the $rank$ method.
 
-Decompressing a subarray is experimented later in this thesis. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a 
-subarray decompression very convenient and fast. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared. $Select$ based method performs better due to its 
-single $select$ call requirement, while the $rank$ call amount depends on the size of the largest element in the subarray.
+Another interesting point is to compare the algorithms when decompressing a long subarray.
+Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element
+and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.
 
@@
 
-\chapter{Directly addressable codes} \label{chapter:DAC}
-
-The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
-data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
-structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.
-
-It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
-list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
-of each element to an array, but it adds a very large overhead which removes the benefit of compression.
-
-\section{Rank and Select}
+\chapter{Rank and Select}
 $Rank$ and $select$ are two array operations that are widely used in compressed data structures.
@@
 $Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index
-of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
+of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two operations are related to each other: when used on a bit array 
 $B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.
@@
 
+\chapter{Directly addressable codes} \label{chapter:DAC}
+
+The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
+data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
+structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.
+
+It is not natively possible to decode the $i$-th element in variable-byte compression algorithms, because the position of the element in the compressed
+list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
+of each element to an array, but it adds a very large overhead which removes the benefit of compression.
+
 \section{DAC via rank}
@@
 Then if the number was stored in multiple chunks, the block of the next chunk is
-stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
+stored to $A_2$ and its continuation bit to $B_2$ and so on.
+
+After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
 from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
@@
 To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
-With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
-different array levels which opens a possibility for further compression optimization.
+With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage.
+
+\pagebreak
 
@@
 
+Interestingly the block length of this method can change within different array levels which opens a possibility for further compression optimization. \citep{Bri09} experiments a few different block 
+length setups and leaves open the problem of finding optimal block lengths.
+While the problem of finding the optimal block length for each data set in regular VB encoding is interesting, the possibility to
+optimize it for each block level separately is surely more challenging.
 \begin{figure}[ht]
@@
 After
-these elements, $P$ is [2,1,0,-1].
-The index locations of the fourth element are fetched from $P$ the same way.
-In the end $P$ is [3,2,0,-1] and the process could potentially go on indefinitely with only a
-very small amount of $rank$ calls.
+these elements, $P$ is [2,1,0,-1]. The index locations of the fourth element are fetched from $P$ the same way. In the end $P$ is [3,2,0,-1] and the process could continue if needed.
 
@@
 
+This thesis explained variable-byte encoding, a compression method that stores the data in fixed length blocks and omits empty leading blocks.
+A continuation bit is added to each block to signal where an
+element ends and a new one begins.
+$Rank$ and $select$, the two bit array operations were also explained.
+Existing solution for direct access with VB using $rank$ was explained.
+A new solution for direct
+access with $select$ was proposed and results were compared between the two solutions. A method of extracting an array of data with VB encoding was proposed for both $rank$ and $select$.
+
 When the dataset contains mostly small integers, the $rank$ based method by Brisaboa et al.
@@
 consists of just a few bit operations.
+The amount of $rank$ queries depend on the length of the subarray and the average size of the elements.
+Therefore even a small amount of large elements in an array
+otherwise filled with small elements causes the runtime to go up.
+This leads to the conclusion that the $select$ method, if not strictly superior, is a strong contender and a good option for decompressing
+a subarray.
 
+\section{Future work}
 There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the select-based approach presented 
@@
  - a real life situation of where to use subarray
-
- - list research questions somewhere? (introduction 1.1?)
 
@@
 
-7. 
- - write a recap about what was done
+ - write more in conclusions

### 2020-04-28 gradu/msc_english_contents.tex
Adds more text

@@
 Comparison of the proposed method and $rank$-based implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$. The $rank$ method 
-is assumed to perform better on data sets containing mostly small integers.
+is expected to perform better on data sets containing mostly small integers.
 It is assumed that larger numbers cause problems in performance.
@@
 \draw[arrow] (index.south) to (select.north);
+\draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=10ex}]
+  (1.8,0) -- (9.5,0) node[midway,yshift=-6em]{The word $w$};
 \end{tikzpicture}
@@
 these elements, $P$ is [2,1,0,-1]. The index locations of the fourth element are fetched from $P$ the same way. In the end $P$ is [3,2,0,-1] and the process could continue if needed.
+
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+[node distance=0pt,
+ start chain = A going right,
+ arrow/.style = {draw=#1,-{Stealth[]}, 
+                shorten >=1mm, shorten <=1mm}, % styles of arrows
+ arrow/.default = black,
+ X/.style = {rectangle, draw,% styles of nodes in string (chain)
+                minimum width=7ex, minimum height=4ex,
+                outer sep=0pt, on chain},
+ Y/.style = {rectangle, draw,% styles of nodes in string (chain)
+                minimum width=4ex, minimum height=3ex,
+                outer sep=0pt, on chain}
+ ]
+
+\node[X] (a1) {$A_{1,1}$};
+\node [left=1ex of a1]{\textbf{$A_1$ =}};
+\node[X] {$A_{2,1}$};
+\node[X] {$A_{3,1}$};
+\node[X] {$A_{4,1}$};
+\node[X] {$A_{5,1}$};
+\node[X] {...};
+
+\node[X, below=0ex of a1] (b1) {0};
+\node [left=1ex of b1]{\textbf{$B_1$ =}};
+\node[X] {1};
+\node[X] {0};
+\node[X] {0};
+\node[X] {1};
+\node[X] {...};
+
+\node[X, below=5ex of b1] (a2) {$A_{1,2}$};
+\node [left=1ex of a2]{\textbf{$A_2$ =}};
+\node[X] {$A_{3,2}$};
+\node[X] {$A_{4,2}$};
+\node[X] {...};
+
+\node[X, below=0ex of a2] (b2) {1};
+\node [left=1ex of b2]{\textbf{$B_2$ =}};
+\node[X] {0};
+\node[X] {1};
+\node[X] {...};
+
+\node[X, below=5ex of b2] (a3) {$A_{3,3}$};
+\node [left=1ex of a3]{\textbf{$A_3$ =}};
+\node[X] {...};
+
+\node[X, below=0ex of a3] (b3) {1};
+\node [left=1ex of b3]{\textbf{$B_3$ =}};
+\node[X] {...};
+
+\node [Y, right=55ex of a1] (p1){-1};
+\node [left=2ex of p1]{\textbf{Initial:}};
+\node[Y] (p1_2) {-1};
+\node[Y] (p1_3) {-1};
+\node[Y] (p1_4) {-1};
+\node [above=1ex of p1]{\textbf{$P_1$}};
+\node [above=1ex of p1_2]{\textbf{$P_2$}};
+\node [above=1ex of p1_3]{\textbf{$P_3$}};
+\node [above=1ex of p1_4]{\textbf{$P_4$}};
+
+\node [Y, below=3ex of p1] (p2){0};
+\node [left=2ex of p2]{\textbf{After $C_1$:}};
+\node[Y] (p2_2){0};
+\node[Y] (p2_3){-1};
+\node[Y] (p2_4){-1};
+
+\node [Y, below=3ex of p2] (p3){1};
+\node [left=2ex of p3]{\textbf{After $C_2$:}};
+\node[Y] {0};
+\node[Y] {-1};
+\node[Y] {-1};
+
+\node [Y, below=3ex of p3] (p4){2};
+\node [left=2ex of p4]{\textbf{After $C_3$:}};
+\node[Y] {1};
+\node[Y] {0};
+\node[Y] {-1};
+
+\node [Y, below=3ex of p4] (p5){3};
+\node [left=2ex of p5]{\textbf{After $C_4$:}};
+\node[Y] {2};
+\node[Y] {0};
+\node[Y] {-1};
+
+\end{tikzpicture}
+\caption{TODO this}\label{bris_subarray}
+
+\end{figure}
+
+ - TODO: does this need a picture?
+
+ - TODO: is referring to Figure~\ref{bris_ds} enough or should this part have another image of the (maybe different) block setup
+
+ - TODO: is the explanation unclear?
 
@@
 Some precaution was used in the index randomizing to prevent reading from an out of bounds location.
+The
+time shown in the figure is milliseconds taken to decompress one million subarrays. Each result is an average of 10 runs.
 
@@
 during the block location phase and therefore is a little bit slower overall.
+It is worth noting that the lower bound of the figure starts at 400ms.
+This was done to clarify the results, which are are very
+close to each other.
 

### 2020-04-28 gradu/msc_english_contents.tex
Adds more text

@@
 and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.
-
- - TODO: introduce each chapter briefly
 
@@
 
-The following example decompresses first four values from Figure~\ref{bris_ds} and assumes the maximum length of integers to be 4 blocks. Array $P$ is used to store the block indices. Initially $P$ = 
-[-1,-1,-1,-1], where -1 denotes that the value is unset and that level has not been reached yet.
-The start of the subarray is index 0.
-$P$[0] is -1, which means the value is unset.
-Because the $rank$ method
-does not need an actual $rank$ call in the first block array, $P$[1] = 0 and that value is used to fetch the first block.
-$B_1$[0] shows that the element is continued to the next block level.
-$P$[1] is also -1,
-so $P$[1] = $rank_0$($B_1$,0) is used to fetch the second block. $B_2$[0] indicates that the element is not continued anymore.
-
-Currently $P$ is [0,0,-1,-1].
-The location of the first block of the next element is fetched by increasing $P$[0] by 1 and then using the incremented value.
+An example situation from a previous chapter is pictured in Figure~\ref{bris_subarray}.
+A subarray of four elements will be decoded from the start of the array.
+The starting location could very well be
+anything, but the start was chosen for simplicity.
+The array $P$ is used to store the block indices.
+Initially the values in $P$ are set to -1 to denote that the value is unset and that level has not been
+reached yet. When a block level is queried for the first time, the index value from the $rank$ call is stored in $P$ and used instead of further $rank$ calls.
+
+The subarray is in this case started from index 0.
+The value of $P$ for the current level is checked at the start of every block fetch.
+$P$[1] is unset, so a value needs to fetched and stored there.
+Because
+the $rank$ method does not need an actual $rank$ call in the first block array, $P$[1] is set to the starting index 0 and that value is used to fetch the first block.
+$B_1$[0] shows that the element is
+continued to the next block level.
+$P$[2] is -1 (because this is the first time the second level is reached), so a $P$[2] = $rank_0$($B_1$,0) call is needed to fetch the second block.
+$B_2$[0] indicates
+that the element is not continued anymore.
+
+For the location of the first block of the second element, $P[1]$ is checked.
+It is set, so the value it contains is increased by one and used as the index.
 The bit array shows that this element is only
 one block long.
-The first two blocks of the third element is fetched the same way.
-$P$[2] is unset, so $P$[2] = $rank_0$($B_2$,0) is used for the third block, which is the last block of that element.
-After
-these elements, $P$ is [2,1,0,-1]. The index locations of the fourth element are fetched from $P$ the same way. In the end $P$ is [3,2,0,-1] and the process could continue if needed.
+The first two blocks of the third element are fetched the same way.
+$P$[2] is unset and a  $P$[2] = $rank_0$($B_2$,0) is used to fetch the index.
+The fourth element does not reach to a new
+block level, so each block location is fetched from $P$.
 
@@
 \end{tikzpicture}
-\caption{TODO this}\label{bris_subarray}
+\caption{Index lookup array demonstration in $rank$ based subarray query}\label{bris_subarray}
 
 \end{figure}
-
- - TODO: does this need a picture?
-
- - TODO: is referring to Figure~\ref{bris_ds} enough or should this part have another image of the (maybe different) block setup
-
- - TODO: is the explanation unclear?
 

### 2020-04-29 gradu/msc_english_contents.tex
Adds more text

@@
 
-In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. The proposed method does not use data 
+In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail.
+The underlying question is if a $select$ based method is viable for random
+access. The proposed method does not use data 
 reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
@@
 $Rank_{v5}$ & \text{6.25\% of bit array}\\
-$Select$ & \text{~6-15\% of bit array (TODO: check this}\\
+$Select$ & \text{~8-23\% of bit array (depends on the data)}\\
 \hline
@@
 
+The implementations for these results can be found online at \href{https://github.com/mozzie/vbyte}{https://github.com/mozzie/vbyte}.
+
 \section{VB decoding comparison}
-Three different VB decoding implementations were compared with two different block lengths. Two $rank$ implementations used are based on \citep{Bri09} and use $rank_v$ and $rank_{v5}$ data structures 
+The list of implementations is depicted in Table~\ref{table:algorithms}.
+Three different VB decoding implementations were compared with two different block lengths.
+Two $rank$ implementations used are based on \citep{Bri09} and use $rank_v$ and $rank_{v5}$ data structures
 from the SDSL-library \citep{gbmp2014sea}. The different $rank$ data structures illustrate the trade-off between memory and speed. the $select$ algorithm is the one proposed in this thesis and it is 
@@
 
-\begin{itemize}
-  \item Bris4 - Bris implementation with block size 4, using $rank_v$
-  \item Bris4v5 - Bris implementation with block size 4, using $rank_{v5}$
-  \item Bris8 - Bris implementation with block size 8, using $rank_v$
-  \item Bris8v5 - Bris implementation with block size 8, using $rank_{v5}$
-  \item Select4 - Proposed implementation with block size 4, using $select_{MCL}$
-  \item Select8 - Proposed implementation with block size 8, using $select_{MCL}$
-\end{itemize}
+\begin{table}
+\centering
+\caption{Generated datasets
+\label{table:algorithms}}
+\begin{tabular}{l|c} 
+Algorithm & Explanation \\
+ \hline 
+  Bris4  & Bris implementation with block size 4, using $rank_v$ \\
+ Bris4v5 & Bris implementation with block size 4, using $rank_{v5}$ \\
+ Bris8   & Bris implementation with block size 8, using $rank_v$ \\
+ Bris8v5 & Bris implementation with block size 8, using $rank_{v5}$ \\
+ Select4 & Proposed implementation with block size 4, using $select_{MCL}$ \\
+ Select8 & Proposed implementation with block size 8, using $select_{MCL}$ \\
+
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M). The content of the datasets were chosen as such to see how different implementations 
-work with different sized data.
-
-\begin{itemize}
-  \item all - all integers randomly 1-4 bytes long
-  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
-  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
-  \item onlysmall - all integers 4 bits or less
-\end{itemize}
+Table~\ref{table:datasets} contains four differently constructed datasets that were used for the experiments. Each dataset was generated to three different sizes (5M, 50M, 500M). 
+The content of the datasets were chosen as such to see how different implementations work with different sized data.
+\begin{table}
+\centering
+\caption{Generated datasets
+\label{table:datasets}}
+\begin{tabular}{l|c} 
+Dataset & Explanation \\
+ \hline 
+all (5M)   & all integers randomly 1-4 bytes long \\
+twolarge   & one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte \\
+onelarge    & one eighth of integers 2 bytes long, rest 4 bits or less \\
+onlysmall    & all integers 4 bits or less \\
+
+\hline
+\end{tabular}
+\end{table}
 
@@
 A continuation bit is added to each block to signal where an
-element ends and a new one begins.
-$Rank$ and $select$, the two bit array operations were also explained.
-Existing solution for direct access with VB using $rank$ was explained.
-A new solution for direct
-access with $select$ was proposed and results were compared between the two solutions. A method of extracting an array of data with VB encoding was proposed for both $rank$ and $select$.
-
-When the dataset contains mostly small integers, the $rank$ based method by Brisaboa et al.
-is superior to the proposed $select$ based method.
-The benefit of using a direct array index instead of a $rank$
-call for the first block gives a huge advantage.
-The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements until $select$
-performs better. This can be seen from figure~\ref{graph:data_comparison}.
-
-The order of how data is stored with $select$ compression enables an efficient fetching of a subarray.
-Only one $select$ query is needed in the beginning, and the decoding of the elements past the first one
-consists of just a few bit operations.
-The amount of $rank$ queries depend on the length of the subarray and the average size of the elements.
-Therefore even a small amount of large elements in an array
-otherwise filled with small elements causes the runtime to go up.
-This leads to the conclusion that the $select$ method, if not strictly superior, is a strong contender and a good option for decompressing
-a subarray.
+element ends and a new one begins. $Rank$ and $select$, two array operations were explained and efficient implementations over bit arrays were introduced in detail.
+
+An efficient solution to direct access to any element in a compressed data structure was previously introduced by Brisaboa et al. Their solution is using VB encoding and a $rank$ data structure. 
+Another solution using $select$ was proposed in this thesis.
+Both solutions were experimented with a few different data structure implementations.
+As expected, the $rank$ version performed better with data sets that contain small integers.
+This is largely due to the
+clever block reorganizing technique that allows the first block to be fetched with just an array lookup.
+The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs
+to have a large portion of longer elements until $select$ performs better.
+
+All implementations were slightly modified to support subarray fetching. The order of how data is stored with $select$ compression enables an efficient fetching of a subarray. Only one $select$ query 
+is needed in the beginning, and the decoding of the elements past the first one consists of just a few bit operations.
+The amount of $rank$ queries depend on the length of the subarray and the average
+size of the elements.
+Therefore even a small amount of large elements in an array otherwise filled with small elements causes the runtime to go up.
+This leads to the conclusion that the $select$ method,
+if not strictly superior, is a strong contender and a good option for decompressing a subarray.
 
 \section{Future work}
-There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the select-based approach presented 
-in this thesis benefits from SIMD. The data needs less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
+There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the $select$ based approach presented 
+in this thesis benefits from SIMD. The data needs less manipulation because it is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
 
@@
 
- - more text / references
-
- - remove lists from algorithms and data sets in results?
-
- - a real life situation of where to use subarray
-
- - should the c++ code be added as appendix?
-
- - is metatext necessary? i.e. starting each chapter with "This chapter contains..."
-
- - explain "X is done", "Y is used" in more detail (why is this done / why is this used)
-
- - Some abbreviations are used. Should there be a list of them in the beginning?
-
- - is Elias-Fano important or relevant to mention?
-
- - should terms / practices be opened more (for people who don't know the field)
-
-results:
-
- - what do the differences in speed or memory usage mean in practice. Is one better than the other?
-
- - the differences in general need more explaining
-
- - write more in conclusions
+ - clean up code
+
+ - see if metatext is needed somewhere

### 2020-04-29 generate.cpp
modified generator

@@ -5,14 +5,14 @@ using namespace std;
 
 int main(int argc, char *argv[]) {
   if(argc<=2) {
-    cout  << "Usage: generate <uint64_t count> <number of bits in number>" << endl;
+    cout  << "Usage: generate <uint64_t count> <name>" << endl;
     exit(1);
   }
   int amount = atoi(argv[1]);
   if(amount == 0) {
     exit(1);
   }
-  int max = atoi(argv[2]);
+
   random_device rd;
   mt19937_64 gen(rd());
   uniform_int_distribution<unsigned long long> dis;
@@ -23,14 +23,12 @@ int main(int argc, char *argv[]) {
   filename.append(argv[2]);
    int maxvalues [8] = {1<<7,1<<8,1<<15,1<<16,1<<23,1<<24,1<<30,1<<30}; // <- all
   // int maxvalues [8] = {1<<7,1<<7,1<<7,1<<8,1<<8,1<<8,1<<16,1<<31}; // <- twolarge
-  // int maxvalues [8] = {1<<3,1<<4,1<<5,1<<6,1<<7,1<<8,1<<16,1<<31}; // <- vsmall
   //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<15}; // <- onelarge
-  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onebyte
+  //int maxvalues [8] = {1<<2,1<<2,1<<3,1<<3,1<<3,1<<4,1<<4,1<<4}; // <- onlysmall
   ofstream file(filename, ios::out | ios::binary);
   for (int n=0; n<amount; ++n) {
     int rand = dis(gen)%8;
     uint64_t i = dis(gen)%maxvalues[rand];
-//    uint64_t i = dis(gen)%max;
     file.write(reinterpret_cast<const char *>(&i), sizeof(uint64_t));
   }
   file.close();

### 2020-05-24 gradu/gradu.tex
Adds a few updates
Adds more text
Final fixes

@@
 \usepackage{pgfplots} 
+\usepackage{cleveref}
 \singlespacing               %line spacing options; normally use single
@@
 % Dr. or in Finnish toht. or tri or FT, TkT, Ph.D. or in Swedish... 
-\supervisors{Prof.~D.U.~Mind, Dr.~O.~Why}
-\examiners{Prof.~D.U.~Mind, Dr.~O.~Why}
+\supervisors{Dr.~Simon~Puglisi, Dr.~Juha~Kärkkäinen}
+\examiners{Dr.~Simon~Puglisi, Dr.~Juha~Kärkkäinen}
 
@@
 Enormous datasets are a common occurence today and compressing them is often beneficial. Fast direct access to any element in the compressed data is a requirement in the field of compressed 
-data structures, which is not possible with traditional compression methods.
+data structures, which is not easily supported with traditional compression methods.
 
 Variable-byte encoding is a method for compressing integers of different byte lengths. It removes unused leading bytes and adds an additional \textit{continuation bit} to each byte to 
-denote whether the compressed integer continues to the next byte or not. An existing solution using \textsc{rank} data structure performs well in this given task. This thesis introduces an 
-alternative solution using \textsc{select} data structure and compares the two implementations. An experimentation is also done on retrieving a subarray from the compressed data structure.
+denote whether the compressed integer continues to the next byte or not. An existing solution using a \textit{rank} data structure performs well in this given task. This thesis introduces an 
+alternative solution using a \textit{select} data structure and compares the two implementations. An experimentation is also done on retrieving a subarray from the compressed data structure.
 
@@
 \backmatter
-%\begin{appendices}
+\begin{appendices}
 
@@
 
-%\end{appendices}
+\end{appendices}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-05-24 gradu/msc_english_contents.tex
Final fixes

@@
 \chapter{Introduction}
-Enormous datasets are a common case in today's applications. Some datasets are too large to handle in memory and thus must be stored in another, slower storage \citep{Hon10}. 
+Enormous datasets are a common case in today's applications. Some datasets are too big to handle in memory and thus must be stored in larger, slower storage \citep{Hon10}. 
 Compressing these datasets is often beneficial, because doing so naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95, Wil99}.
 
-Fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
+A fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
 frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
-not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
-data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
-need to fulfill.
-
-However, efficient random access to compressed data is very useful in compressed data structures \citep{Ven13}. It saves storage space, bandwidth and 
-increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
-from the beginning.
+not trivial to determine the exact location of a certain element within the compressed data. If this is required, the usual 
+data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the the sought element is located. 
+Fortunately, such random access to elements is not a requirement compression algorithms usually need to fulfill.
+
+However, efficient random access to compressed data is very useful in compressed data structures \citep{Ven13}.  In addition to saving storage space and bandwidth, the compression 
+increases the likelihood of data already being in cache \citep{Sch02}.
 
 Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuation bit} is added to the data to denote whether the compressed integer
-is continued to the next byte or not. Two different light-weight data structures are built over the continuation bit array. $Rank_1$($i$) returns the number of 1 bits in the bit array between 
+is continued to the next byte or not.
+To facilitate random access, two different light-weight data structures are built over the continuation bit array.
+$Rank_1$($i$) returns the number of 1 bits in the bit array between
 indexes 0 and $i$-1. $Select_1$($i$) returns the index of the i-th 1 bit in the array. The fact that the compressed data blocks are of the same size is used
@@
 reorganizing and a $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
-has been widely adopted \citep[e.g.][]{Kon17, Sha16}.
-
-In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail.
+has been widely adopted \citep[see, e.g.,][]{Kon17, Sha16}.
+
+In this thesis, a novel alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail.
 The underlying question is if a $select$ based method is viable for random
-access. The proposed method does not use data 
-reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
-new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.
-
-Comparison of the proposed method and $rank$-based implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$. The $rank$ method 
-is expected to perform better on data sets containing mostly small integers.
-It is assumed that larger numbers cause problems in performance.
+access. Compared to \citeauthor{Bri09}, proposed method uses a very simple data 
+reorganizing, and capitalizes on the assumption that data is encoded the same way it is in the source. Data can then be read straight from the memory and reassembling the integer from 
+its encoded blocks is not needed.
+
+Comparison of the proposed method and the $rank$-based implementation by \citeauthor{Bri09} is provided with different implementations of $rank$ and $select$. The $rank$ method 
+is expected to perform better on data sets containing mostly small integers. It is assumed that larger numbers lead to slower random access for $rank$-based methods. 
 The performance of the proposed $select$ method is indifferent 
 of the size of the elements in the data.
-Different kinds of data sets are used to see how each algorithm performs.
+Different kinds of data sets are used to assess how each approach performs.
 The assumption is that $select$ works better with data sets with very large integers.
@@
 
-Another interesting point is to compare the algorithms when decompressing a long subarray.
-Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element
+The performance of the algorithms when decompressing a long subarray is also compared.
+Because of the way the data is stored, the proposed $select$ method offers fast access to the next or previous element
 and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.
@@
 
-Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the integer data sets. Their encoding process assigns short bit sequences to small integers, 
-which is why they outperform in terms of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. VB is generally faster 
+Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for integer data sets. Their encoding process assigns short bit sequences to small integers, 
+which is why they outperform, in terms of compression, VB encoding on datasets containing lots of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. VB is generally faster 
 to decode \citep{Wil99, Tro03} and so is sometimes preferred, also due to its simplicity of implementation.
@@
 
-VB encoding was first experimented as a tool for compressing inverted index lists of word locations in documents by \citep{Sch02}. 
+VB encoding was first experimented as a tool for compressing inverted index lists of word locations in documents by \citeauthor{Sch02}. 
 In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced. A search engine may use an inverted index of 
-words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
-advanced search purposes. Usually these lists are preprocessed to an inverted index, storing each number as its difference to the previous number 
-instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but inverted index encoding makes the integers are small. 
+words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in the document for 
+advanced search purposes. Usually these lists are preprocessed to form an inverted index, storing each number as its difference to the previous number 
+instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, which makes the inverted index encoded integers small.
 In contrast, rare words have only a few entries but the integers stored are larger.
@@
 
-\section{VB encoding}
-The name \textit{variable-byte} is reflected in how the integers are stored.
-When stored in an array, each element requires the same amount of space.
-This is very inefficient if a lot of the values are
+\section{Variable-byte encoding}
+The name \textit{variable-byte} reflects the way in which the integers are stored.
+When the integers are stored in an array, each element requires the same fixed amount of space.
+This is inefficient if a lot of the values are
 very small but the array also has to support very large integers.
@@
 from the beginning are discarded.
-Because the lengths of the integers may now be different, the data cannot be stored as it is.
+Because the lengths of the integers (i.e.
+the number of bytes used to represent them) may now be different, the data cannot be stored as it is.
 Instead a \textit{continuation bit} is added in front of each block to form
@@
 are removed from the beginning and then the continuation bits are added. 1 is added in front of the last block (containing the least significant bits) and 0 to the other blocks, resulting in 
-\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with byte length 4.
+\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with block length 4.
 
@@
 Since the block
-length is 7, the continuation bit is added in row 8 by adding $2^8$ = 128 to the last block.
+length is 7, the continuation bit is added in row 8 by adding $2^7$ = 128 to the last block.
 
@@
 
-Smaller block lengths can yield a better compression rate at the cost of more bit manipulation and therefore possibly slower decompression, while bigger block lenghts need less bit manipulation and 
-offer less effective compression. On the other hand, bigger block length means a smaller percentage of added continuation bits.
+Smaller block lengths can yield a better compression rate at the cost of more bit manipulation and therefore possibly slower decompression, while bigger block lengths need less bit manipulation and 
+offer less effective compression. On the other hand, a bigger block length means a smaller percentage of added continuation bits.
 Generally, a block length of 7 has been popular because it tends to perform well on average and handling chunks as bytes is 
@@
 
-\section{VB decoding}
+\section{Variable-byte decoding}
 VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
@@
 
-As how the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
+Continuing from where the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer.
+The block from the first chunk is
 extracted and added to $n$, making $n$ = \texttt{10} (bit representation of 2).
 The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
-changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
-$n = $\texttt{101010} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.
+changing $n$ = \texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
+$n$ = \texttt{101010} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.
 
@@
 
-To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect to the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
+To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
 created in addition to the existing data.
@@
 
-VB encoding has several advantages with $search$ and $rank$: the data is compressed in 
-blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
-as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
+VB encoding has several advantages with $select$ and $rank$: the data is compressed in 
+blocks of equal length which significantly shrinks the length of $B$. Moreover the bit array $C$ formed from the continuation bits already stores the locations of items and works
+as the needed indexing array. In this case, $rank_1$(C, i) would give the number of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
 of i-th compressed element.
@@
 The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
- of both rank and select to support the bit array. Also a few useful functions from the library were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
-size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
+ of both rank and select to support the bit array. A few useful functions from the library were also used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
+size requirements of the implementations used. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
 the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
@@
 
-For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, both division operations being integer divisions.
-Then number of 1 bits
-from index $r$ to $i$ are calculated from the original bit array. All three values added 
-together add up to $rank$($i$).
-$Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit.
+For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, with both division operations being integer divisions. Then 
+the number of 1 bits from index $r$ to $i$ are calculated from the original bit array.
+These three values add up to $rank$($i$).
+$Rank_{v5}$ is a lighter structure: its superblock size is 2048 and relative counts are taken for every 384th bit.
 This causes the final bit calculation to be more costly,
-but reduces memory requirement to a quarter.
+but reduces memory requirements to a quarter.
 
@@
 A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks.
-This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits.
-The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock.
+This means a relative block needs to be able to store a number up to 512 in itself, which takes $\log_2 512$ = 9 bits.
+The first relative block always equals 0, because a relative block counts the number of 1 bits from the start of the superblock.
 Therefore the remaining 7 relative blocks take 63 bits and fit into one
@@
 
-The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each. The first relative block again equals to 0, so the rest 
+The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $\log_2 2048$ = 11 bits each. The first relative block again equals 0, so the rest 
 of the blocks fit to a 64-bit word. The $rank_{v5}$ implementation requires an additional (64+64)/2048 = 6.25\% space.
 
-The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
+The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and the location of every 64th set bit is stored 
 relative to the superblock.
-With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached.
+With similar calculations, the location of the closest 64th set bit is calculated and then the bit array is iterated until the required amount of bits is reached.
 The $select_{MCL}$
- is a practical variant of a PAT tree \citep{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.
-
-In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
+ is a practical variant of a PAT tree, the data structure described by \cite{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.
+
+For both rank and select, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
 block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}. The 
-implementations are explained in greater detail in \citep{gbmp2014sea}.
+implementations are explained in greater detail in \cite{gbmp2014sea}.
 
@@
 
-The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
-data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
-structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.
+The ability to process large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to reduce the size of data so that it fits 
+into cache, memory or even hard drive while still allowing it to be accessed easily. Directly addressable codes (DAC), a direct access to any element in a compressed list or array, is 
+one of the basic building blocks in compressed data structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.
 
 It is not natively possible to decode the $i$-th element in variable-byte compression algorithms, because the position of the element in the compressed
-list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
-of each element to an array, but it adds a very large overhead which removes the benefit of compression.
+list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution would be to store the location 
+of each element to an array, but this adds a very large overhead which removes the benefit of compression.
 
@@
 
-After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
-from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
+After the data has been arranged this way, a $rank$ data structure is built for each bit array. The data structure does not require additional space on top of that used by the normal VB 
+encoding apart from that used by the $rank$ structure.
+Figure~\ref{bris_ds} contains a visualization of the data structure.
+The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and
 contiuation bit arrays $B_k$.
@@
 \end{tikzpicture}
-\caption{Data structure by Brisaboa et al., visualized}\label{bris_ds}
+\caption{Data structure by \citeauthor{Bri09}, visualized}\label{bris_ds}
 
@@
 This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current
-block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.
+block) is 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.
 
@@
 There are two zero
-bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is set to 1 so the decompression is completed.
-
-Interestingly the block length of this method can change within different array levels which opens a possibility for further compression optimization. \citep{Bri09} experiments a few different block 
+bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is 1 so the decompression is completed.
+
+Interestingly, the block length of this method can be allowed to change within different array levels which opens a possibility for further compression optimization.
+\citeauthor{Bri09} experiments with a few different block
 length setups and leaves open the problem of finding optimal block lengths.
@@
 \end{minipage}
-\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}
+\caption{Example pseudo code of DAC with rank by \citeauthor{Bri09}} \label{bris_pseudo}
 
@@
 Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
-in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
+in $B$ and a single $select_1$ query gets the desired starting point, while the aforementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
 the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
@@
 
-Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own array, which leaves the blocks in their own array. This allows the 
-compressed number to be read from the memory block and removes the need of block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
+Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own bit array, which leaves the blocks in their own array. This allows the 
+compressed number to be read from the memory block and removes the need to do block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
 when reading a word from the block byte array, the bits and bytes are already in correct order.
 
-For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$(i) gets the index location
+For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$($i$) gets the index location
 of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
@@
 If the offset was not zero, there are $o$ trailing bits in $w$ from the previous integer.
-These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the desire value. Most systems are
+These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the desired value. Most systems are
 able to read memory only from byte addresses, which leads to a corner case where the compressed integer is stored in more bytes than are contained in a word. In the aforementioned situation, this 
-happens if compressed length was 8 blocks and offset was 4. Then another step of reading another byte and storing the remaining bits to $w$ is required. This is entirely avoidable if the integer 
-length is constrained. In this case the maximum integer length of 28 bits would eliminate the handling of the second byte.
+happens if the compressed length was 8 blocks and offset was 4.
+In this case, another step of reading another byte and storing the remaining bits to $w$ is required.
+This is entirely avoidable if the integer
+length can be constrained. If the maximum integer length was set to 28 bits, the bits would be within a single word in memory and reading the extra byte would not be needed.
 
@@
 This can be done by reading a word from the bit array and bit-shifting the word
-until the bit representing the start of the i-th element is on the rightmost end and then counting the trailing zeros.
-This counts the number of blocks in the i-th element that do not have the continuation bit set.
+until the bit representing the start of the $i$-th element is on the rightmost end and then counting the trailing zeros (for example, by using the GCC function \texttt{\_\_builtin\_ctz}).
+The trailing zeros equal to the number of blocks in the $i$-th element that do not have the continuation bit set.
 This number is then incremented by one to also count the ending block which has the continuation bit set to 1.
@@
 In the pseudo code, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns
-a bit mask for k bytes. In line 8, a long word is read from the memory starting at memory address $A$[$begin$].
+a bit mask for k bytes. In line 8, a long word is read from memory starting at memory address $A$[$begin$].
 
@@
 element is 3, which is obtainable, e.g., via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$ and blocks not belonging to the element
-are removed either with a bitmask or by double bit shifting, resulting in k-th element being decoded.
+are removed for example with a bit mask, resulting in k-th element being decoded.
 
@@
 \end{tikzpicture}
-\caption{Relation of continuation bits and data blocks in VB encoding} \label{VBarrays}
+\caption{Relation of continuation bits and data blocks in VB encoding.} \label{VBarrays}
 \end{figure}
@@
 the data is at first split into low and 
-high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
-For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}.
+high bits. The low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
+For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as a unary bit array B = \texttt{0100100100010001}.
 
@@
 returns the location of i-th 1 in B. Up to index $k$, the bit array has had $i$ ones and $i-k$ zeros. The $i$ corresponds to the number of compressed integers in $B$ and every zero in the array 
-means an increment of 1 in the numbers. In other words, $i-k$ equals to the integer at position $i$ before gap-encoding. Therefore $select_1(i)$ - i returns the original higher bit number. The split 
+means an increment of 1 in the numbers.
+In other words, $i-k$ equals to the integer at position $i$ before gap-encoding.
+Therefore $select_1(i)$ - $i$ returns the original higher bit number.
+The split
 between bits is made to reduce the gap sizes.
 
-Elias-fano encoding is an excellent tool if the elements are sorted and the higher bits are reasonably small, because large integers cause a very inefficient unary encoding.
+Elias-Fano encoding is an excellent tool if the elements are sorted and the higher bits are reasonably small, because large integers cause a very inefficient unary encoding.
 This can be further optimized
-by finding clusters in the integers and partitioning the indexes \citep{Ott14}. Elias-Fano encoding is not compared in this thesis, because the restriction of sorted data is relaxed.
+by finding clusters in the integers and partitioning the indexes \citep{Ott14}. Elias-Fano encoding is not compared in this thesis, because the data considered is not necessarily sorted.
 
@@
 
-After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements becomes significantly easier.
-Calculating the memory location of the next element is not
-needed because the next compressed value is usually located next to the previously fetched element. This is especially beneficial to the select-based VByte decoding, where calculating the start 
-byte location takes majority of the runtime. In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory.
+After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements cam become significantly easier.
+Calculating the memory location of the next element may not
+be needed if the next compressed value is located next to the previously fetched element. This is especially beneficial to the select-based VByte decoding, where calculating the start 
+byte location takes majority of the runtime.
+In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory and immediately local to the last point of access.
 
@@
 the first block is on the rightmost end, and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
-and then counting trailing zeros. Once the word is shifted through, another word is read and process continues.
+and then counting trailing zeros. Once the word is shifted through, another word is read and the process continues.
 
@@
 \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=10ex}]
-  (1.8,0) -- (9.5,0) node[midway,yshift=-6em]{The word $w$};
+  (1.8,0) -- (9.5,0) node[midway,yshift=-6em]{word $w$};
 \end{tikzpicture}
@@
 $i$ = $select$($k-1$) +1 is the following bit on the right.
-The byte where $i$-th bit is located can be calculated by doing an integer division $i$/8.
-A 16-bit word $w$ is read from the byte location $B_1$.
+The byte where the $i$-th bit is located can be calculated by doing an integer division $i$/8.
+In the example, a 16-bit word $w$ is read from the byte location $B_1$.
 It is beneficial to read as large word as possible, but 16-bits is used in this example for the sake of simplicity.
 Next the starting bit is shifted to the start of the word, the offset being 6 in this case.
-After the operation, $w$ contains bits \texttt{00110010 10000000}, least significant bit being on the left. The length of the current element is $l$ = 1 + \textsc{TrailingZeros}($w$). The number of 
-trailing zeros can be efficiently calculated with GCC function \texttt{\_\_builtin\_ctz} or \texttt{bits::lo} from the 'SDSL-lite' library.
-
-The length of the desired element $l$ and the location $i$ can now be used to decompress desired element.
+After the operation, $w$ contains bits \texttt{00110010 10000000}, with the least significant bit being on the left.
+The length of the current element is $l$ = 1 + \textsc{TrailingZeros}($w$).
+The number of
+trailing zeros can be efficiently calculated with the GCC function \texttt{\_\_builtin\_ctz} (or \texttt{bits::lo} from the 'SDSL-lite' library).
+
+The length of the desired element $l$ and the location $i$ can now be used to decompress the desired element.
 For the next element, $i$ is incremented by $l$ and $w$ is bit shifted $l$ bits and the process is
@@
 The number of leftover
-zeros is added to the first length calculation after word switch.
+zeros is added to the first length calculation after reading a new word for $w$.
 This number can be calculated either by adding bit shift amounts together and subtracting the result from word-length.
@@
 \section{Subarray access with rank}
-The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a subarray, the goal becomes to minimize $rank$ calls. 
+The shown $rank$ method by \citeauthor{Bri09} described earlier can also effectively decompress the next value with slight modifications.
+When fetching a subarray, the goal becomes to minimize $rank$ calls.
 When a new block level is reached for the first time during a subarray query, the index is fetched with $rank$ as normal, and the result of the query is stored.
@@
 The starting location could very well be
-anything, but the start was chosen for simplicity.
+anywhere, but the start was chosen for simplicity.
 The array $P$ is used to store the block indices.
@@
 The value of $P$ for the current level is checked at the start of every block fetch.
-$P$[1] is unset, so a value needs to fetched and stored there.
+$P$[1] is unset, so a value needs to be fetched and stored there.
 Because
@@
 \end{tikzpicture}
-\caption{Index lookup array demonstration in $rank$ based subarray query}\label{bris_subarray}
+\caption{Index lookup array demonstration in $rank$ based subarray query.}\label{bris_subarray}
 
@@
 \section{VB decoding comparison}
-The list of implementations is depicted in Table~\ref{table:algorithms}.
-Three different VB decoding implementations were compared with two different block lengths.
-Two $rank$ implementations used are based on \citep{Bri09} and use $rank_v$ and $rank_{v5}$ data structures
-from the SDSL-library \citep{gbmp2014sea}. The different $rank$ data structures illustrate the trade-off between memory and speed. the $select$ algorithm is the one proposed in this thesis and it is 
+The list of implementations is depicted in Table~\ref{table:algorithms}. Three different VB decoding implementations were compared with two different block lengths. Two $rank$-based implementations 
+of VB used are based on \citep{Bri09} and use $rank_v$ and $rank_{v5}$ data structures 
+from the SDSL-library \citep{gbmp2014sea}.
+The different $rank$ data structures achieve different trade-offs between memory and speed.
+The $select$ algorithm is the one proposed in this thesis and it is
 implemented using the $select_{MCL}$ data structure from the SDSL-library. Different block lengths were used to illustrate how block size affects the runtime and memory requirement. Lengths 
@@
 
-Each implementation is compressing and decompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
-and \autoref{chap:select_impl}, respectively.
-
-Table~\ref{table:datasets} contains four differently constructed datasets that were used for the experiments. Each dataset was generated to three different sizes (5M, 50M, 500M). 
-The content of the datasets were chosen as such to see how different implementations work with different sized data.
+Each implementation compresses and decompresses 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \Cref{chap:implementations} 
+and \Cref{chap:select_impl}, respectively.
+
+Table~\ref{table:datasets} describes four different types of synthetic datasets that were used for the experiments. Each dataset was generated to three different sizes (5M, 50M, 500M). 
+The content of the datasets were chosen in order to see how different implementations work with different sized data and different magnitudes of integers.
 \begin{table}
@@
 \label{table:datasets}}
-\begin{tabular}{l|c} 
+\begin{tabular}{l|l} 
 Dataset & Explanation \\
@@
 
-All implementations read the data set from a file and compress it to memory, randomize one million query indexes and store results to an array. The times shown in Table~\ref{table:results1} 
-are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated multiple times and an average of the runtimes is taken.
+All implementations read the data set from a file and compress it in memory, randomize one million query indexes and store these indexes to an array. The times shown in Table~\ref{table:results1} 
+are times taken from looping through the index array and VB decoding the number at each index. Additionally, each run is iterated 10 times and an average of the runtimes is taken.
 
@@
 with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
-of one million $rank$ and $select$ calls with different average block lengths. The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior.
-
-The 8-bit versions tend to perform better than the 4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4-bit versions achieve 
-better compression in data sets containing small integers. Select versions do not depend as much on the block size, because all the blocks in the compressed element are read at the same time. 
-Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
-in every scenario. This is most likely due to cache, because the stored data set is smaller in the 4-bit version.
+of one million $rank$ and $select$ calls with different average block lengths.
+The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior, which leads to a much flatter curve.
+
+The 8-bit versions tend to perform better than the 4-bit versions.
+This is because the 8-bit version stores more data in one block and so less block readings are required.
+However, the 4-bit versions achieve
+better compression in data sets containing small integers. The select versions do not depend as much on the block size, because all the blocks in the compressed element are read at the same time. 
+Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though the 4-bit version should execute at least the same commands as the 8-bit version 
+in every scenario. This is most likely an effect of caching, because the stored data set is smaller in the 4-bit version, and thus more of it fits in higher levels of memory.
 
 The $rank_{v5}$ naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
-significantly less memory than $rank_v$, as seen on Figure~\ref{table:supportsize}.
+significantly less memory than $rank_v$, as seen on Table~\ref{table:supportsize}.
 
@@
  of 50 million 64-bit numbers.
-The compressed size column has a sum of continuation bit array and the data blocks.
+The compressed size column is the sum of the sizes of the continuation bit array and the data blocks.
 The tables show that the static requirement of $rank_{v5}$ is superior with these datasets.
@@
 
-The subarray access experiments were run similarly to the previous experiments.
+The subarray access experiments were run similarly to the experiments of previous sections.
 The data sets used contained 50 million numbers each.
-Most of the numbers were 4-bit long, with a varying amount of 32-bit
+Most of the numbers were 4 bits long, with a varying amount of 32-bit
 integers randomly among them.
@@
 The 4-bit version of $select$ requires a few more instructions
-during the block location phase and therefore is a little bit slower overall.
-It is worth noting that the lower bound of the figure starts at 400ms.
-This was done to clarify the results, which are are very
-close to each other.
+during the block location phase and is therefore slightly slower overall.
 
@@
 \end{tikzpicture}
-\caption{$Rank$ and $select$ performance with subarray fetching} \label{graph:subarray}
+\caption{$Rank$ and $select$ performance with subarray fetching. The scale of the y-axis was selected to clarify the difference of the results.} \label{graph:subarray}
 \end{figure}
@@
 
-An efficient solution to direct access to any element in a compressed data structure was previously introduced by Brisaboa et al. Their solution is using VB encoding and a $rank$ data structure. 
-Another solution using $select$ was proposed in this thesis.
+An efficient solution to direct access to any element in a compressed data structure was previously introduced by \citeauthor{Bri09}. Their solution uses VB encoding and a $rank$ data structure. 
+An alternative solution using $select$ was proposed in this thesis.
 Both solutions were experimented with a few different data structure implementations.
@@
 The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs
-to have a large portion of longer elements until $select$ performs better.
+to have a large portion of longer elements before $select$ performs better.
 
@@
 size of the elements.
-Therefore even a small amount of large elements in an array otherwise filled with small elements causes the runtime to go up.
+Therefore even a small amount of large elements in an array otherwise filled with small elements causes the runtime to increase significantly.
 This leads to the conclusion that the $select$ method,
-if not strictly superior, is a strong contender and a good option for decompressing a subarray.
+if not strictly superior, is a strong contender and a good option in particular when decompressing subarrays is required.
 
 \section{Future work}
-There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the $select$ based approach presented 
+There has already been a few successful attempts to decompress VB encoded integers sequentially with SIMD instructions \citep{Lem18,Pla15}.
+It is worthwhile to check if the $select$ based approach presented
 in this thesis benefits from SIMD. The data needs less manipulation because it is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.
@@
 
-TODO:
-
- - clean up code
-
- - see if metatext is needed somewhere
+TODO: a few paragraphs about haswell instructions

### 2020-05-24 gradu/msc_english_contents.tex
Final fixes

@@
 
-Another improvement to VB encoding is to see whether the compressed numbers can be changed.
+The $rank$ operation can be created with a small number of instructions utilizing a modern \textsc{POPCOUNT} machine instruction \citep{Zho13}. The $select$ approach is 
+known to be slower, ranging from 50 to 80 machine instructions. Intel's Haswell CPUs offer two new interesting machine instructions, \textsc{PDEP} and \textsc{TZCNT}. \textsc{PDEP} stores bits
+from its first operand based on the location of 1 bits in the second operand. \textsc{TZCNT} is a machine instruction to count trailing zeros. These two can be used to create a $select_1$ query on 
+a machine word with just 4 instructions \citep{Pan17a}. They have been tested with the $select$ of the SDSL-library and show an improvement of 2-15\% \citep{Pan17b}. Additionally, the 
+$select$ implementation proposed in this thesis uses a trailing zero count to calculate the length of the element. Some improvement in speed may happen if the implementation is changed 
+to use \textsc{TZCNT} instead.
+
+One more interesting improvement to VB encoding is to see whether the compressed numbers can be changed.
 This is trivial until the block length of the compressed number changes.
@@
 cache is needed and the supporting rank/select data structure needs to be recalculated.
-
-TODO: a few paragraphs about haswell instructions

### 2020-05-24 gradu/msc_english_contents.tex
Adds Haswell machine word text

@@
 from its first operand based on the location of 1 bits in the second operand. \textsc{TZCNT} is a machine instruction to count trailing zeros. These two can be used to create a $select_1$ query on 
-a machine word with just 4 instructions \citep{Pan17a}. They have been tested with the $select$ of the SDSL-library and show an improvement of 2-15\% \citep{Pan17b}. Additionally, the 
-$select$ implementation proposed in this thesis uses a trailing zero count to calculate the length of the element. Some improvement in speed may happen if the implementation is changed 
-to use \textsc{TZCNT} instead.
+a machine word with just 4 instructions \citep{Pan17a}. They have been tested with the $select$ algorithm of the SDSL-library (which is also used in the proposed method of this thesis) 
+and show an improvement of 2-15\% \citep{Pan17b}. Additionally, a trailing zero count function is used to calculate the length of the element. 
+Some improvement in speed may happen if the implementation is changed to use \textsc{TZCNT} machine instruction instead.
 

### 2020-06-06 gradu/msc_english_contents.tex
Clarifications

@@
 not trivial to determine the exact location of a certain element within the compressed data. If this is required, the usual 
-data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the the sought element is located. 
+data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the sought element is located. 
 Fortunately, such random access to elements is not a requirement compression algorithms usually need to fulfill.
@@
 of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two operations are related to each other: when used on a bit array 
-$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.
+$B$ = \texttt{0100 1101 \underline{1}011}, $select_1$(5) returns 8 (underlined) and therefore $rank_1$(8) returns 4.
 
@@
 blocks of equal length which significantly shrinks the length of $B$. Moreover the bit array $C$ formed from the continuation bits already stores the locations of items and works
-as the needed indexing array. In this case, $rank_1$(C, i) would give the number of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
+as the needed indexing array. In this case, $rank_1$(i) from $C$ would give the number of end blocks before the i-th index and $select_1$(i) would give the location of the ending block 
 of i-th compressed element.
@@
 Therefore the remaining 7 relative blocks take 63 bits and fit into one
-64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.
+64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 bits for the relative blocks, resulting in additional 128/512 = 25\% space.
 
@@
 \label{chap:select_impl}
-Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
+Using $select$ on the continuation bit array to achieve direct access is more intuitive than using $rank$. The element locations are already marked with 1's 
 in $B$ and a single $select_1$ query gets the desired starting point, while the aforementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
@@
 
-To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
+To use $select$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
 only on its last block, $Select_1$($i$) returns the location of the end byte of $i$-th element. Therefore the start of  $j$-th element in the block array is at 
@@
 
-For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$($i$) gets the index location
-of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
+For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits.
+First, the byte location of the first block is needed.
+The index location $x$ of the desired block
+ is calculated $x = select_1$($i$-1) + 1, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (x \times 4) \div 8$, the latter operand being for integer division.
 The first block might not be at the start of the byte $s$.
-This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4).
+This is why offset $o = (x \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4).
 Then a 32-bit word $w$ is read from memory from byte location $s$.
@@
 
-The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
+The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$($i$-1) from $select_1$($i$). This however causes an additional 
 $select_1$ query, which is costly.
@@
 
-After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements cam become significantly easier.
+After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements can become significantly easier.
 Calculating the memory location of the next element may not
@@
  \hline 
-all (5M)   & all integers randomly 1-4 bytes long \\
+all   & all integers randomly 1-4 bytes long \\
 twolarge   & one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte \\
@@
 with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
-of one million $rank$ and $select$ calls with different average block lengths.
+of one million $rank$ and $select$ calls with different average integer lengths.
 The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior, which leads to a much flatter curve.
@@
 and show an improvement of 2-15\% \citep{Pan17b}. Additionally, a trailing zero count function is used to calculate the length of the element. 
-Some improvement in speed may happen if the implementation is changed to use \textsc{TZCNT} machine instruction instead.
+Some improvement in speed may happen if the implementation is changed to use the \textsc{TZCNT} machine instruction instead.
 
